---
title: "05 Introduction to PISA"
# bibliography: references.bib
---

```{r}
#| eval: true
#| warning: true
#| echo: false

# only load this code if testing this file independently of overall project
if(!exists("test")){
  warning("RUNNING LOCAL FILE LOAD")
  
  library(glue)
  proj_folder <- "C:/Users/Peter/Google Drive/Kings/Code/MASTEMR/"
  proj_folder <- r"(/Users/k1765032/Google Drive/My Drive/MASTEMR/)"
  
  node <- Sys.info()
  if(node["nodename"] == "KCLJ7J6LFC74J"){ #Richard's PC
    proj_folder <- r"(/Users/k1765032/Google Drive/My Drive/MASTEMR/)"
  }else if(node["nodename"] == "KCL3Y7NKR3"){ #Pete's NEW Laptop
    proj_folder <- r"(G:\My Drive\Kings\Code\MASTEMR\)"
  }else if(node["nodename"] == "PETES_PC"){ #Pete's NEW Laptop
    proj_folder <- r"(C:\Users\Peter\Google Drive\Kings\Code\MASTEMR\)"
  }

  source(glue("{proj_folder}MASTEMR_setup.R"))
}
```

## Seminar activities

### Discussion activity

### Tasks

**Getting set up**

Ensure you have the PISA 2018 data set loaded. If you can see the PISA_2018 data frame in your enviroment window (at the top right of your screen), there is no need to reload.

```{r}
#| eval: false
#| code-fold: false
#| code-summary: "Loading the data"
# Load the PISA data set.

library(arrow)
library(tidyverse)

PISA_2018 <- read_parquet(r"[<folder>PISA_2018_student_subset.parquet]")
```

### Task1

::: question
-   Use geom_density to plot distributions
Plot the distribution of Japanese and UK mathematics scores - what patterns do you notice

```{r}
#| eval: true
#| code-fold: false
#| code-summary: "Loading the data"

JPUKMath<-PISA_2018%>%
  select(CNT, PV1MATH)%>%
  filter(CNT=="United Kingdom"|CNT=="Japan")

ggplot(data=JPUKMath,
       aes(x=PV1MATH, fill=CNT, alpha=0.5))+
  geom_density()

```
-   Examine gender differences

```{r}
#| eval: true
#| code-fold: false
#| code-summary: "Loading the data"

UKMathGender<-PISA_2018%>%
  select(CNT, PV1MATH, ST004D01T)%>%
  filter(CNT=="United Kingdom")

ggplot(data=UKMathGender,
       aes(x=PV1MATH, fill=ST004D01T, alpha=0.5))+
  geom_density()

```
-   Facet wrap gender differences across countries

```{r}
#| eval: true
#| code-fold: false
#| code-summary: "Loading the data"

MathGender<-PISA_2018%>%
  select(CNT, PV1MATH, ST004D01T)%>%
  filter(CNT=="United Kingdom"|CNT=="United States"|CNT=="Japan"| CNT=="Korea"|CNT=="Finland")

ggplot(data=UKMathGender,
       aes(x=PV1MATH, fill=ST004D01T, alpha=0.5))+
  geom_density()+
  facet_wrap(.~CNT)

```
::: 


::: question
-   Examine patterns of mathematics achievement by wealth

```{r}
#| eval: true
#| code-fold: false
#| code-summary: "Loading the data"

Wealthdata<-PISA_2018%>%
  select(CNT, WEALTH, PV1MATH)%>%
  group_by(CNT)%>%
  summarise(MeanWealth=mean(WEALTH, na.rm = TRUE),
            MeanMath=mean(PV1MATH, na.rm = TRUE))

ggplot(data=Wealthdata,
       aes(x=MeanWealth, y=MeanMath, alpha=0.5))+
  geom_point()+
  xlab("Wealth")+
  ylab("Mathematics score")

```
Add colour to outliers


```{r}
#| eval: true
#| code-fold: false
#| code-summary: "Loading the data"

Wealthdata<-PISA_2018%>%
  select(CNT, WEALTH, PV1MATH)%>%
  group_by(CNT)%>%
  summarise(MeanWealth=mean(WEALTH, na.rm = TRUE),
            MeanMath=mean(PV1MATH, na.rm = TRUE))%>%
  mutate(label=ifelse(MeanMath>550, "Red", "Blue"))

ggplot(data=Wealthdata,
       aes(x=MeanWealth, y=MeanMath, colour=label))+
  geom_point()+
  xlab("Wealth")+
  ylab("Mathematics score")

```

Add label to outliers


```{r}
#| eval: true
#| code-fold: false
#| code-summary: "Loading the data"

Wealthdata<-PISA_2018%>%
  select(CNT, WEALTH, PV1MATH)%>%
  group_by(CNT)%>%
  summarise(MeanWealth=mean(WEALTH, na.rm = TRUE),
            MeanMath=mean(PV1MATH, na.rm = TRUE))%>%
  mutate(label=ifelse(MeanMath>550, "Red", "Blue"))%>%
  mutate(labelname=ifelse(MeanMath>550, as.character(CNT), NA))
  

ggplot(data=Wealthdata,
       aes(x=MeanWealth, y=MeanMath, colour=label))+
  geom_point()+
  geom_text(aes(label=labelname), 
            colour="black", 
            check_overlap = TRUE)+
  xlab("Wealth")+
  ylab("Mathematics score")

```

Examine Likert responses by country using Facet plot

`ST125Q01NA` - How old were you when you started early childhood education? 

```{r}
#| eval: true
#| code-fold: false
#| code-summary: "Loading the data"

Culturedata<-PISA_2018%>%
  select(CNT, ST125Q01NA)%>%
  group_by(CNT)
  
ggplot(data=Culturedata,
       aes(x=ST125Q01NA))+
  geom_bar()+
  xlab("How old were you when you started early childhood education?")+
  theme(axis.text.x = element_text(angle = 90, vjust = 0.5, hjust=1))

```
Then use faceting to split the plots by country

```{r}
#| eval: true
#| code-fold: false
#| code-summary: "Loading the data"

Culturedata<-PISA_2018%>%
  select(CNT, ST125Q01NA)%>%
  filter(CNT=="United Kingdom"|CNT=="Germany" | CNT=="Belgium"|   CNT=="Austria"| CNT=="France" | CNT=="Poland" | CNT=="Estonia" | CNT=="Finland"| CNT=="Italy")
  
ggplot(data=Culturedata,
       aes(x=ST125Q01NA, fill=CNT))+
  geom_bar()+
  xlab("How old were you when you started early childhood education?")+
  theme(axis.text.x = element_text(angle = 90, vjust = 0.5, hjust=1))+
  facet_wrap(. ~ CNT)

```

Categorising Variables

```{r}
#| eval: true
#| code-fold: false
#| code-summary: "Loading the data"

WealthSci<-PISA_2018%>%
  select(CNT, WEALTH, PV1SCIE)%>%
  filter(CNT=="United Kingdom"|CNT=="Germany")%>%
  mutate(WTHCTG=ifelse(WEALTH>0, ifelse(WEALTH>1, "Very high", "High"), "Low"))%>%
  mutate(SciCat=ifelse(PV1SCIE>460, "Below Average","Above Average"))%>%
  group_by(CNT)%>%
  droplevels()

WealthSci<-as.data.frame(table(WealthSci))


  
ggplot(data=Culturedata,
       aes(x=ST125Q01NA, fill=CNT))+
  geom_bar()+
  xlab("How old were you when you started early childhood education?")+
  theme(axis.text.x = element_text(angle = 90, vjust = 0.5, hjust=1))+
  facet_wrap(. ~ CNT)

```

:::

::: callout-tip

:::

