---
title: "07 T-tests"
# bibliography: references.bib
editor: 
  markdown: 
    wrap: 72
---

```{r}
#| eval: true
#| warning: true
#| echo: false

# only load this code if testing this file independently of overall project
if(!exists("test")){
  warning("RUNNING LOCAL FILE LOAD")
  
  library(glue)
  proj_folder <- "C:/Users/Peter/Google Drive/Kings/Code/MASTEMR/"
  proj_folder <- r"(/Users/k1765032/Google Drive/My Drive/MASTEMR/)"
  
  node <- Sys.info()
  if(node["nodename"] == "KCLJ7J6LFC74J"){ #Richard's PC
    proj_folder <- r"(/Users/k1765032/Google Drive/My Drive/MASTEMR/)"
  }else if(node["nodename"] == "KCL3Y7NKR3"){ #Pete's NEW Laptop
    proj_folder <- r"(G:\My Drive\Kings\Code\MASTEMR\)"
  }else if(node["nodename"] == "Amys-MacBook-Air.local"){ #Amy's NEW Laptop
    proj_folder <- r"(/Users/amyobrien/Google Drive/My Drive/MASTEMR/)"
    
  }

  source(glue("{proj_folder}MASTEMR_setup.R"))
}
```

## T-tests

T-tests are statistical tests that determine if there are statistically
significant differences between the means of two groups. Consider the
data giving the results of girls and boys on a test:

**Boys' scores**

| Boy  | Score (/10) |
|------|-------------|
| A    | 7           |
| B    | 8           |
| C    | 6           |
| D    | 5           |
| E    | 8           |
| Mean | 6.8         |

**Girls' scores**

| Girls | Score (/10) |
|-------|-------------|
| A     | 9           |
| B     | 4           |
| C     | 7           |
| D     | 8           |
| E     | 8           |
| Mean  | 7.2         |

The means of the two groups are different (6.8 for boys and 7.2 for
girls) but the difference may simply be due to random variation. A
t-test can report the probability that a difference is means is due to
chance.

```{r}
#| eval: true
library(tidyverse)
Boys<-data.frame(Score=c(7,8,6,5,8))
Girls<-data.frame(Score=c(9,4,7,8,8))
print(mean(Boys$Score))
print(mean(Girls$Score))
t.test(Boys$Score, Girls$Score)

```

The `t.test` function returns a p-value greater than `0.05` so the mean
test-results can be assumed to the result of chance, and we conclude
there are no detectable differences by gender for this data set.

There are a number of types of t-tests:

• One-tailed test - checks if the mean of one group is different in one
direction (e.g. are scores only higher)

• Two-tailed tests - checks if the means are different in two
directions - i.e. are tests higher or lower

• Paired tests - compares the means of measurements from the same
individual or object (e.g. in a pre- and post-test - the comparison of
means and before and after scores compares the same student's scores
before and after)

• Unpaired t-tests - compares the means of two unrelated groups (for
example, are the mean science scores of boys and girls different)

For more information on t-tests, see [chapter
13](https://learningstatisticswithr.com/book/ttest.html), in Navaro's
Learning Statistics with R.

### Task 1 Choice of tests

Consider the following scenarios. Which (if any) of the t-tests would be
most of appropriate, including whether it would be a one or two tailed
test.

::: callout-note
One and two-tailed tests

Remember that a one-tailed test only looks for differences in one
direction from the mean (for example, that one sample has a higher mean
than the other). A two-tailed tests tests for the possibility of the
means of the two groups being higher or lower than each other.
:::

::: question
a)  An educational researcher wants to test a teaching approach with a
    group of 12 students to see if an intervention increases performance
    in their year 6 SATs results. Pupils are given a SATS paper before
    the intervention and again after. The results of these are
    normalised and then compared.

b)  A teacher wants to see if their pupils' GCSE computer science test
    scores are in line with the national average or not. The tests are
    out of 180 marks.

c)  The DfE want to compare uptake of STEM subjects pre and post
    pandemic to see if there has been any change in the percentage of
    students studying at least one STEM subject for A level.

d)  A school wants to compare salaries of staff between this academic
    year and last academic year to see whether there has been a
    significant increase.

e)  The same school as in d) wants to compare salaries of support staff
    and teaching staff to see if there is a significant difference.

f)  A researcher wants to find out how much time pupils spend looking at
    screens during lesson time, to see if there is a difference between
    year 7 and year 10 students. 10 pupils in year 7 and 10 pupils in
    year 10 are randomly selected across a school and observed
    throughout a school day in November. The times are recorded to the
    nearest minute.

g)  A business manager wants to check if the financial expenditure on
    STEM teachers per year is similar to the national average (based on
    average numbers of pupils on role)
:::

## T-tests

There are four assumptions that need to be met by dataset for a t-test
to be applicable: a) the data are continuous b) the sample was randomly
sampled c) The variance of the data in each group is similar
(homogeneity of variance) d) The data are approximately normally
distributed

::: callout-note
A rule of thumb researchers use is that if the ratio of variances of the
larger sample to the smaller is less than 4, then a t-test can be used.
:::

**Normality checking**

To check for normality we can use a number of different tests and
approaches, some of which will be covered in the seminar. For data sets
that are below 5000 observations then Sharpiro-Wilk is a good test to
use. For data sets above 5000 (as with the PISA data) then qqplots and
the Kolmogorov-Smirnov Test test are good approaches. Histograms can
also be used to help visualise the data and compare with the normal
distribution.

Are the PISA reading scores in the UK normally distributed?

Quantiles are sections of data, for example a type of quantile is a
quartile, a section of data containing a quarter of the data points, a
quintile contains 20% of the data and so on. A quintile-quantile plot,
plots the quantiles of two distributions against each other. If the
distributions are the same, a straight line is formed.

```{r}
#| eval: true
#| code-fold: false
# Create a UK reading subset
PISAUKRead<-PISA_2018%>%
  select(CNT, PV1READ)%>%
  filter(CNT=="United Kingdom")
# Perform a qqplot
qqnorm(PISAUKRead$PV1READ)
# If data are normally distributed, a qq plot should be a straight line.
# To assist judgement, you can add a straight line with:
qqline(PISAUKRead$PV1READ, col="red")
# The data in this case look normal.


```

A non-graphical approach to normality testing is the Kolmogorov-Smirnov
Test test. The package `dgof` contains `ks.test` which tests for
normality. The test is carried out by passing the data to
`ks.test(PISA_2018$PV1READ, "pnorm", mean=mean(PISA_2018$PV1READ), sd=sd(PISA_2018$PV1READ))`.
`"pnorm"` sets the test to testing for normality (`ks.test` can test for
different distributions), we also pass the function the mean (`mean`)
and standard deviation (`sd`) of our dataset. If the p-value is below
0.05, the data are normally distributed (or parametric).

```{r}
#| warning: false

# the Kolmogorov-Smirnov Test
library(dgof)
# Create a UK reading subset
PISAUKRead<-PISA_2018%>%
  select(CNT, PV1READ)%>%
  filter(CNT=="United Kingdom")
ks.test(PISAUKRead$PV1READ, "pnorm", mean=mean(PISAUKRead$PV1READ), sd=sd(PISAUKRead$PV1READ))
# The p-value is less than 0.05, assume the data are normally distributed
```

Finally, a third approach, which is harder to judge, is to plot a
histogram and to examine it for the normal shape.

```{r}
#| eval: true
#| code-fold: false
# Create a UK reading subset
PISAUKRead<-PISA_2018%>%
  select(CNT, PV1READ)%>%
  filter(CNT=="United Kingdom")
# Plot a histogram
ggplot(data=PISAUKRead,
       aes(x=PV1READ))+
  geom_histogram(binwidth = 5, fill="darkseagreen4")


```

### Task 2 - Normality checking

::: question
Determine if the following subsets of PISA data are normal (you will
need to use `select` and `filter`). Try to use all three methods of
normality checking (a) qqplots, b) the Kolmogorov-Smirnov test, c)
histrogram plotting, to reach your conclusion

Are the following data normally distributed: a) PV1MATH scores in the
UK; b) PV1MATH scores in Malta; c) Students in the UK who score more
than 750 on their science test.

```{r}
#| eval: false
#| code-fold: true
#| code-summary: "Answer"
#| warning: false
# a) PV1MATH scores in the UK
# Create UK Math subset
UKMath<-PISA_2018%>%
  select(CNT, PV1MATH)%>%
  filter(CNT=="United Kingdom")
  # Perform a qqplot
qqnorm(UKMath$PV1MATH)
# If data are normally distributed, a qq plot should be a straight line.
# To assist judgement, you can add a straight line with:
qqline(UKMath$PV1MATH, col="red")
# The data in this case look normal.

# Perform the KS test
ks.test(UKMath$PV1MATH, "pnorm", mean=mean(UKMath$PV1MATH), sd=sd(UKMath$PV1MATH))
# The p-value is 0.04783, less than 0.05, so we can assume the scores are normal

# Plot a histogram
ggplot(data=UKMath,
       aes(x=PV1MATH))+
  geom_histogram(binwidth = 5, fill="darkseagreen4")
# The data in this case look normal.

# b) PV1MATH scores in Malta

# Create Malta Math subset
MalMath<-PISA_2018%>%
  select(CNT, PV1MATH)%>%
  filter(CNT=="Malta")
  # Perform a qqplot
qqnorm(MalMath$PV1MATH)
# If data are normally distributed, a qq plot should be a straight line.
# To assist judgement, you can add a straight line with:
qqline(MalMath$PV1MATH, col="red")
# The data in this case look normal.

# Perform the KS test
ks.test(MalMath$PV1MATH, "pnorm", mean=mean(MalMath$PV1MATH), sd=sd(MalMath$PV1MATH))
# The p-value, 0.03173, less than 0.05, so we can assume the scores are normal

# Plot a histogram
ggplot(data=MalMath,
       aes(x=PV1MATH))+
  geom_histogram(binwidth = 5, fill="darkseagreen4")
# The data in this case look normal.

# c) UK science high achievers

# Create a UK over 750 Science score subset
UKHighSci<-PISA_2018%>%
  select(CNT, PV1SCIE)%>%
  filter(CNT=="United Kingdom" & PV1SCIE>750)
  # Perform a qqplot
qqnorm(UKHighSci$PV1SCIE)
# If data are normally distributed, a qq plot should be a straight line.
# To assist judgement, you can add a straight line with:
qqline(UKHighSci$PV1SCIE, col="red")
# The data in this case does not look normal.

# Perform the KS test
ks.test(UKHighSci$PV1SCIE, "pnorm", mean=mean(UKHighSci$PV1SCIE), sd=sd(UKHighSci$PV1SCIE))
# The p-value < 2.2e-16, less than 0.05, so we can assume the scores are normal

# Plot a histogram
ggplot(data=UKHighSci,
       aes(x=PV1SCIE))+
  geom_histogram(binwidth = 5, fill="darkseagreen4")
# The data in this case look normal.


```
:::

**Checking for Equal Variances**

The second assumption of a test is that the samples have equal variance.
Variance describes the spread of the data (standard deviation is the
square root of variance).

To check for equal variances, we can divide the variance of one sample
by the other sample. A rule of thumb researchers use is that if the
ratio of variances of the larger sample to the smaller is less than 4,
then a t-test can be used. R will check this for you, and if they are
unequal, will use the test for unequal variances (Welch's test).

To check for the ratio of variance, we can use the `var` function, and
calculate the ratio of the sample. We add `na.rm=TRUE` to the `var`
function to omit NA values. For example, determine the ratio of girls'
and boys' reading scores in the UK

```{r}
#| eval: true
#| code-fold: false
# Create a UK reading subset for boys
PISAUKReadM<-PISA_2018%>%
  select(CNT, PV1READ,ST004D01T)%>%
  filter(CNT=="United Kingdom"|ST004D01T=="Male")
# Create a UK reading subset for girls
PISAUKReadF<-PISA_2018%>%
  select(CNT, PV1READ,ST004D01T)%>%
  filter(CNT=="United Kingdom"|ST004D01T=="Female")
# Find the variance
MVar<-var(PISAUKReadM$PV1READ, na.rm =TRUE)
FVar<-var(PISAUKReadF$PV1READ, na.rm=TRUE)
MVar/FVar

```

The ratio of variances is 1.15 so the samples meet the variance
criterion for t-tests.

###Task 3 - Variance checking

Determine if the variance of these samples meets the criteria to carry
out a t-test: a) Boys and Girls maths scores in the UK b) Reading scores
in the US and the UK c) Wealth scores of boys and girls in the US

::: question
```{r}
#| eval: false
#| code-fold: true
#| code-summary: "Answer"

# a)
# Create a UK maths subset for boys
PISAUKMATHM<-PISA_2018%>%
  select(CNT, PV1MATH,ST004D01T)%>%
  filter(CNT=="United Kingdom"|ST004D01T=="Male")
# Create a UK reading subset for girls
PISAUKMATHF<-PISA_2018%>%
  select(CNT, PV1MATH,ST004D01T)%>%
  filter(CNT=="United Kingdom"|ST004D01T=="Female")
# Find the variance
MVar<-var(PISAUKMATHM$PV1MATH, na.rm =TRUE)
FVar<-var(PISAUKMATHF$PV1MATH, na.rm=TRUE)
MVar/FVar
# The ratio of variance is 1.132708 so a t-test can be performed

# b)
# Create a US reading score dataframe
USREAD<-PISA_2018%>%
  select(CNT, PV1READ)%>%
  filter(CNT=="United States")
# Create a UK reading score dataframe
UKREAD<-PISA_2018%>%
  select(CNT, PV1READ)%>%
  filter(CNT=="United Kingdom")
# Find the variance
UKVar<-var(UKREAD$PV1READ, na.rm =TRUE)
USVar<-var(USREAD$PV1READ, na.rm=TRUE)
UKVar/USVar
# The ratio of variance is 0.813 so a t-test can be performed

# c)
# Create a US Male Wealth  dataframe
USWEALTHM<-PISA_2018%>%
  select(CNT, WEALTH, ST004D01T)%>%
  filter(CNT=="United States"|ST004D01T=="Male")
# Create a US Female Wealth  dataframe
USWEALTHF<-PISA_2018%>%
  select(CNT, WEALTH, ST004D01T)%>%
  filter(CNT=="United States"|ST004D01T=="Female")
# Find the variance
MVar<-var(USWEALTHM$WEALTH, na.rm =TRUE)
FVar<-var(USWEALTHF$WEALTH, na.rm=TRUE)
MVar/FVar
# The ratio of variance is 1.05469 so a t-test can be performed




```
:::

::: callout-note
**Checking and Filtering for Outliers**

Often, we need to check and filter for outliers in the data. There are
many ways to do this, depending on the distribution of the data and how
extreme the outliers are that we want to remove. A good rule of thumb is
the median +/- 1.5 times the interquartile range (IQR), where the IQR is
the upper quartile - lower quartile. If the data is normally
distributed, then the mean +/- 1.5 time the standard deviation (sd) can
be used. If we only want to remove more extreme outliers, then 2 or even
3 times the IQR (or sd) from the median (or mean) can be used instead.

Once we have determined the upper and lower bound of our outliers, we
can remove any data points that lie outside of this and then perform our
statistical test on the data.

```{r}
#| eval: false
#| code-fold: true
#| code-summary: "Answer"
#| warning: false
# Create a subset of UK math scores
UKPISAmath<-PISA_2018%>%
  select(CNT, PV1MATH)%>%
  filter(CNT=="United Kingdom")
# Test for normality in current form
ks.test(UKPISAmath$PV1MATH, "pnorm", mean=mean(UKPISAmath$PV1MATH), sd=sd(UKPISAmath$PV1MATH))
# p-value = 0.04783 - only just normal
# Calculate value of scores at 1.5 x sd above and below the mean
outl_h_b<-mean(PISA_2018$PV1MATH, na.rm=TRUE) +1.5 * sd(PISA_2018$PV1MATH, na.rm=TRUE)
outl_l_b<-mean(PISA_2018$PV1MATH, na.rm=TRUE) -1.5 * sd(PISA_2018$PV1MATH, na.rm=TRUE)
# Filter values that are 1.5 x sd above and below the mean
UKPISAmath <- UKPISAmath %>%
 filter(PV1MATH> outl_l_b & PV1MATH < outl_h_b)
# Test for normality again
ks.test(UKPISAmath$PV1MATH, "pnorm", mean=mean(UKPISAmath$PV1MATH), sd=sd(UKPISAmath$PV1MATH))
# p-value now < 2.2e-16 - a much more normal distribution
```
:::

**Performing t-tests**

::: call-out note ***Types of t-test*** To set whether the t.test in R
is one-tailed or two-tailed you can set the function `alternative` in
`t.test` to "greater", "less" or "two.sided". To set the test as paired
or unpaired, you can set the `paired` variable to TRUE or FALSE.

```{r}
#| eval: true
#| code-fold: false
# Create a UK reading subset
UKSci<-PISA_2018%>%
  select(CNT, PV1SCIE,ST004D01T)%>%
  filter(CNT=="United Kingdom")
# Plot a two-sided, unpaired t-test
t.test(PV1SCIE ~ ST004D01T, data = UKSci, 
       paired = FALSE, alternative = "two.sided")


```

:::

To perform a t.test in R you use the `t.test` function. You need to pass
the function the data set you want to test (`data=USMath`), and specify
the groups to be compared (`PV1MATH ~ ST004D01T`) - here comparing the
means of the mathematics scores (`PV1AMTH`) by the gender variable
(`ST004D01T`). You can then specify if the test is two sided
(`alternative = "two.sided"`), and the type of test (`paired = FALSE` -
i.e. this is an unpaired test). For example, to determine if there are
statistically signfciant differences in boys and girls maths scores in
the United States:

```{r}
#| eval: true
#| code-fold: false
# Create a UK reading subset
USMATH<-PISA_2018%>%
  select(CNT, PV1MATH,ST004D01T)%>%
  filter(CNT=="United States")
# Plot a two-sided, unpaired t-test
t.test(PV1MATH ~ ST004D01T, data = USMATH, 
       paired = FALSE, alternative = "two.sided")


```

To interpret the outcome of the test, the key value is the p-value (in
this case, p-value = 0.008006). As the p-value is \<0.05 we can conclude
that there are significant differences between boys' and girls' mean
mathematics scores in the US.

### Task 3 T-tests

Check you have the `PISA_2018` dataset loaded.If not, click here to
download the parquet file:
[PISA_2018.parquet](https://drive.google.com/open?id=17Owl5F5fmuOWiAXQomVfQMr_9CYDHoBC&authuser=richardandrewbrock%40gmail.com&usp=drive_fs)

To load the data, use the code below:

```{r}
#| eval: false
#| warning: false
install.packages("arrow") # if you haven't already
library(arrow)
library(tidyverse)
PISA_2018 <- read_parquet("<location of the downloaded PISA_2018.parquet file>")
```

::: question
1.  Use the PISA dataset and the gender `ST004D01T` and `PV1MATH` (Math
    score) variables to perform a two-sided, unpaired t-test to
    determine if boys and girls perform differently on the test in the
    UK. Before performing the test, check for the normality of the data,
    and check for equal variance.

```{r}
#| eval: false
#| warning: false
#| code-fold: true
#| code-summary: "Answer"
# Are there differences between the mean scores of UK boys and girls in PISA mathematics?

# Select the gender (ST004D01T) and math score columns (PV1MATH)
# Filter the data to select UK responses

MaleUK <- PISA_2018 %>%
  select(CNT, ST004D01T, PV1MATH) %>%
  filter(CNT=='United Kingdom',
         ST004D01T=='Male')

FemaleUK <- PISA_2018 %>%
  select(CNT, ST004D01T, PV1MATH) %>%
  filter(CNT=='United Kingdom',
         ST004D01T=='Female')

# The conditions to do a t-test include that the data are normally distributed
# and there is homogeneity (similarity) of the variances (the squared standard deviations)
# Let us check the conditions are met by calculating first if the data sets are normally
# distributed using a qq plot

qqnorm(MaleUK$PV1MATH)
qqline(MaleUK$PV1MATH, col="red")

qqnorm(FemaleUK$PV1MATH)
qqline(FemaleUK$PV1MATH, col="red")

# The plots produces relatively straight lines so the distributions can be assumed to be normal
#
# We will then check the variances of the two data sets

VarM <- var(MaleUK$PV1MATH, na.rm=TRUE)
VarF <- var(FemaleUK$PV1MATH, na.rm=TRUE)
VarM / VarF

# The variance ratio is close to 1 (1.1)
# So our two conditions are met and can we can perform the t-test

# Create a UK Math subset
UKMATH<-PISA_2018%>%
  select(CNT, PV1MATH,ST004D01T)%>%
  filter(CNT=="United Kingdom")
# Plot a two-sided, unpaired t-test
t.test(PV1MATH ~ ST004D01T, data = UKMATH, 
       paired = FALSE, alternative = "two.sided")

# The p-value is <0.05 (4.061e-08) suggesting there are statistically
# differences between boys and girls


```

2.  Choose a different country of interest to compare performance in
    males and females in that country

```{r}
#| eval: false
#| warning: false
#| code-fold: true
#| code-summary: "Answer"
# Are there differences between the mean scores of boys and girls in China PISA mathematics?
#
# Select the gender (ST004D01T) and math score columns (PV1MATH)
# Filter the data to select UK responses

MaleChina <- PISA_2018 %>%
  select(CNT,ST004D01T, PV1MATH) %>%
  filter(CNT=='B-S-J-Z (China)', ST004D01T=='Male')

FemaleChina<-PISA_2018 %>%
  select(CNT,ST004D01T, PV1MATH) %>%
  filter(CNT=='B-S-J-Z (China)', ST004D01T=='Female')

# The conditions to do a t-test include that the data are normally distributed
qqnorm(MaleChina$PV1MATH)
qqline(MaleChina$PV1MATH, col="red")
qqnorm(FemaleChina$PV1MATH)
qqline(FemaleChina$PV1MATH, col="red")


# The plots produces relatively straight lines so the distributions can be # assumed to be normal
#
# We will then check the variances of the two data sets

VarM<-var(MaleChina$PV1MATH, na.rm=TRUE)
VarF<-var(FemaleChina$PV1MATH, na.rm=TRUE)
VarM/VarF

# The variance ratio is close to 1 (1.17)
# So our two conditions are met and can we can perform the t-test

# Create a UK Math subset
CHIMATH<-PISA_2018%>%
  select(CNT, PV1MATH,ST004D01T)%>%
  filter(CNT=="B-S-J-Z (China)")
# Plot a two-sided, unpaired t-test
t.test(PV1MATH ~ ST004D01T, data = CHIMATH, 
       paired = FALSE, alternative = "two.sided")

# The p-value is <0.05 (2.414e-07) suggesting there are statistically
# differences between boys and girls in mathematics in China


```

3.  Now try to compare the performance of males and females in all
    `OECD` countries in the PISA dataset. You may need to think about
    how to do this using R without repeating the test for each country

```{r}
#| eval: false
#| code-fold: true
#| code-summary: "Answer"

MaleOECD <- PISA_2018 %>%
  select(OECD, ST004D01T, PV1MATH) %>%
  filter(OECD=='Yes', ST004D01T=='Male')

FemaleOECD <- PISA_2018 %>%
  select(OECD, ST004D01T, PV1MATH) %>%
  filter(OECD=='Yes', ST004D01T=='Female')

qqnorm(MaleOECD$PV1MATH)
qqline(MaleOECD$PV1MATH, col="red")
qqnorm(FemaleOECD$PV1MATH)
qqline(FemaleOECD$PV1MATH, col="red")


# The plots produces relatively straight lines so the distributions can be assumed to be normal

# We will then check the variances of the two data sets

VarM <- var(MaleOECD$PV1MATH, na.rm=TRUE)
VarF <- var(FemaleOECD$PV1MATH, na.rm=TRUE)
VarM / VarF

# The variance ratio is close to 1 (1.1)
# So our two conditions are met and can we can perform the t-test

# The p-value is <0.05 (2.2e-16) suggesting there are statistically
# differences between boys and girls

OECD <- PISA_2018 %>%
  select(OECD, ST004D01T, PV1MATH) %>%
  filter(OECD=='Yes')

t.test(PV1MATH ~ ST004D01T, data = OECD, 
       paired = FALSE, alternative = "two.sided")

```

4.  Are there significant differences in mean wealth scores of boys and
    girls in the UK?

```{r}
#| eval: false
#| code-fold: true
#| code-summary: "Answer"

MaleUK <- PISA_2018 %>%
  select(ST004D01T, WEALTH) %>%
  filter(ST004D01T=='Male')

FemaleUK <- PISA_2018 %>%
  select(ST004D01T, WEALTH) %>%
  filter(ST004D01T=='Female')

qqnorm(MaleUK$WEALTH)
qqline(MaleUK$WEALTH, col="red")
qqnorm(FemaleUK$WEALTH)
qqline(FemaleUK$WEALTH, col="red")

# The plots here are somewhat skewed which might suggest not performing the test

# We will then check the variances of the two data sets

VarM <- var(MaleUK$WEALTH, na.rm=TRUE)
VarF <- var(FemaleUK$WEALTH, na.rm=TRUE)
VarM / VarF

# The variance ratio is close to 1 (1.1)
# So our two conditions are met and can we can perform the t-test

WEALTH <- PISA_2018 %>%
  select(WEALTH, ST004D01T, CNT) %>%
  filter(CNT=='United Kingdom')

t.test(PV1MATH ~ ST004D01T, data = OECD, 
       paired = FALSE, alternative = "two.sided")

# p-value < 2.2e-16 so there are statistically significant differences between boys' and girls' reported wealth in the UK

```
:::

## Anova (multiple t-tests)

We have performed t-tests of the difference in mathematics scores
between boys and girls in the UK. We may now want to perform the same
t-test for all the countries in the sample. This can create a problem.
Remember that a p-value report the probability that an outcome is due to
chance. The more times we repeat a t-test, the more likely we are to see
an effect, even when none-exists.

To avoid potential artefacts from doing multiple t-tests, when we want
to compate the means of more than two groups, we use an anova (anlaysis
of variance) test.

To determine if there is a statistically significant difference between
groups we run an anova calculation, using the `aov` function in R. The
dependent variable is the one we are interested in explaining, the
independent variables are the factors we think might explain the
variance.

`aov(data, dependent_var ~ independent_var + independent_var + ...)`

For example, we used t-tests to look at differences in scores between a
pair of countries (e.g the UK and the US). We can now compare across all
the countires.

We pass the function the data we wish to focus on (`PV1MATH`) and then
indicate we wish to look for variation in the `PV1MATH`, by `CNT`. We
then summarise the result to get a table.

```{r}
#| eval: TRUE
#| code-fold: false

ALLPISAMATH<-PISA_2018%>%
  select(CNT, PV1MATH)

resaov <- aov(data=ALLPISAMATH, PV1MATH ~ CNT)
sumresaov <- summary(resaov)
sumresaov

```

The resulting table has two rows: `CNT` residuals. The `aov` function
looks to determine if there is a difference in the variance across
countires. After determining any variation from those two variables, any
remaining variation is associated with residuals - you can think of this
as the unexplained variation that isn't associated with the vectors we
specified (i.e. the country).

First, look at the `Pr(>F)` column. This is a test of significance and
reports if the groups in that row show statistically significant
variation. In our case, country returns `Pr(>F) <2e-16 ***` - which is
over `0.05`. This suggests there are significant differences across
countries.

This is not terribly interesting so far(!) - but we can do two post-hoc
tests (i.e. tests after the first test, to find out the effect size and
between which countries differences exist)

## Eta-squared

Now knowing that there are significant differences between the two class
groups, we can next determine how much variation in the test score is
explained by class. To do this we calculate a variable called
eta-squared.

Eta-squared gives the proportion of variance explained by each variable.
The eta squared function is in the package `lsr` so we will install
`lsr`, and then use the result of our anova (`resaov`) to calculate the
eta squared variable using the function `etaSquared`. To report the
value of eta as a percentage we need to multiply the output of eta by
100.

Eta squared tells us the proportion of the total variance that is
explained by a variable (you can also think of it as an effect size). An
eta squared value of 1 indicates all the variance of a sample is explain
by some variable and 0 means the variable is not responsible for any of
the observed variance. We can multiply the eta-squared score by 100 to
get a value for the percentage of variance explained.

The percentage of variance explained is a useful figure. For example, it
has been reported that schools only account for 14% of the variance in
progress 8 scores, whilst family explains 43% (@wilkinson2018assessing).

```{r}

library(lsr)
eta <- as.data.frame(etaSquared(resaov))
eta <- eta*100
eta

```

The important column here is the `eta.sq` column - it tells us that the
country explains 26% of the variance in test scores. (The second column
contains information about partial eta-squared, which we won't go into,
but are used when the results of each measure are *not independent* i.e.
one result influences another).

::: callout-note
As a rule of thumb, a percentage variance explained of 1% is considered
small, 6% medium and 14% and more large
:::

## Tukey's HSD

When an anova test reports that are some statistically significant
differences between groups, it does not imply there are statistically
significant differences between all subgroups. For example, if the anova
reports statistically significant differences by age, statistically
significant differences might exist between 11, 12 and 13-year old
students, but not between 13 and 14-year olds.

We can use an additional anova test, **Tukey's Honest Significant
Difference** test (or Tukey's HSD for short), to find out which pairs of
subgroups have statistically significant differences in means.

Consider the question: Are there statistically significant differences
in the variance of science scores of the UK, US, France and Germany? To
determine if such differences exist, we create a new subset for those
countries' science scores, and then run an anova test by country,
reporting the eta squared value.

```{r}
PISAMULTI <- PISA_2018 %>%
  select(CNT, PV1SCIE)%>%
  filter(CNT=="United Kingdom" | CNT=="United States" | CNT=="Germany" | CNT=="France")
  # filter(CNT %in% c("United Kingdom", "United States", "Germany", "France"))

resaov<-aov(PV1SCIE ~ CNT, data=PISAMULTI)
sumresaov<-summary(resaov)
sumresaov

eta<-as.data.frame(etaSquared(resaov))
eta<-eta*100
eta
```

The anova results tell us there are significant differences between the
countries, which account for 0.4% of variance in scores. We can then run
a Tukey HSD test to determine which countries have significant
differences between mean scores.

```{r}
TukeyHSD(resaov)
```

From that we get a table with p values (p adj) for different pairs of
countries. Note that these are below 0.005 for all pairs of countries,
except the US and the UK. So we can conclude there are significant
differences in science scores between all countries except the UK and
the US.

## Task 4 -Multiple t-tests with anova

::: question
For the UK, France, Thailand, and the US determine if there are
statistically significant differences in variation in mathematics
scores.
:::

```{r}
#| code-fold: true
#| code-summary: answer
#| eval: true
PISAMULTI<-PISA_2018 %>%
  select(CNT, PV1MATH)%>%
  filter(CNT=="United Kingdom" | CNT=="Thailand" | CNT=="United States" | CNT=="France")


resaov<-aov(PV1MATH ~ CNT, data=PISAMULTI)
sumresaov<-summary(resaov)
sumresaov

eta<-as.data.frame(etaSquared(resaov))
eta<-eta*100
eta

TukeyHSD(resaov)
# There are statistically significant differences in mathematics scores for all pairs of countries

```

::: question
For the UK, are there statistically significant differences in reading
score by students who attend different types of school (`ISCEDO`), e.g.
Vocational, General etc.
:::

```{r}
#| code-fold: true
#| code-summary: answer
#| eval: true
UKSchRead<-PISA_2018 %>%
  select(CNT, PV1READ, ISCEDO)%>%
  filter(CNT=="United Kingdom")


resaov<-aov(PV1READ ~ ISCEDO, data=UKSchRead)
sumresaov<-summary(resaov)
sumresaov

eta<-as.data.frame(etaSquared(resaov))
eta<-eta*100
eta

TukeyHSD(resaov)
# There are statistically significant differences between Pre-vocational and general, and vocational and pre-vocational, but not between vocational and general.

```

::: question
For the UK, use `ifelse` on the WEALTH variable to create categories for four quartiles (divide the data into top; upper-mid; lower-mid; and bottom). Use anova to determine between which groups there are statistically signficant differences in science scores

Hint: You can use `quantiles<-quantile(UKSci$PV1SCIE, prob=c(.25,.5,.75), na.rm=TRUE)` to give the values of PV1SCIE that account for 25%, 50% and 75% of the responses.
:::

```{r}
#| code-fold: true
#| code-summary: answer
#| eval: true
UKSci<-PISA_2018 %>%
  select(CNT, PV1SCIE, WEALTH)%>%
  filter(CNT=="United Kingdom")%>%
  na.omit()%>%
  droplevels()%>%
  select(PV1SCIE, WEALTH)
# Find the cut of points for 25%, 50% and 75% (quantiles) of WEALTH
quantiles<-quantile(UKSci$WEALTH, prob=c(.25,.5,.75), na.rm=TRUE)

UKSci<-UKSci%>%
  mutate(WEALTH=ifelse(WEALTH<quantiles[1], "Bottom", 
                        ifelse(WEALTH<quantiles[2], "Lower-mid",
                               ifelse(WEALTH<quantiles[3], "Upper-mid","Top"))))

resaov<-aov(PV1SCIE ~ WEALTH, data=UKSci)
sumresaov<-summary(resaov)
sumresaov

TukeyHSD(resaov)
# There are statistically significant differences between all groups by wealth

```



## Doing t-tests in R

<iframe width="560" height="315" src="https://www.youtube.com/embed/KdhAcLPq5B4?start=3" title="YouTube video player" frameborder="0" allow="accelerometer; autoplay; clipboard-write; encrypted-media; gyroscope; picture-in-picture" allowfullscreen>

</iframe>

You can find the code from the video below:

```{r}
#| label: ttest video code
#| eval: false
#| code-fold: true
#| code-summary: "Show the Code"

# Introduction to t-tests in R
#
# Download data from /Users/k1765032/Library/CloudStorage/GoogleDrive-richardandrewbrock@gmail.com/.shortcut-targets-by-id/1c3CkaEBOICzepArDfjQUP34W2BYhFjM4/PISR/Data/PISA/subset/Students_2018_RBDP_none_levels.rds
# You want the file: Students_2018_RBDP_none_levels.rds
# and place in your own file system
# change loc to load the data directly. Loading into R might take a few minutes
install.packages("nortest")

library(tidyverse)
library(nortest)
loc <- "/Users/k1765032/Library/CloudStorage/GoogleDrive-richardandrewbrock@gmail.com/.shortcut-targets-by-id/1c3CkaEBOICzepArDfjQUP34W2BYhFjM4/PISR/Data/PISA/subset/Students_2018_RBDP_none_levels.rds"
PISA_2018 <- read_rds(loc)

# Are there differences between the mean scores of UK boys and girls in PISA mathematics?
#
# Select the gender (ST004D01T) and math score columns (PV1MATH)
# Filter the data to select UK responses

MaleUK<-PISA_2018 %>%
  select(CNT,ST004D01T, PV1MATH) %>%
  filter(CNT=='United Kingdom') %>%
  filter(ST004D01T=='Male')

FemaleUK<-PISA_2018 %>%
  select(CNT,ST004D01T, PV1MATH) %>%
  filter(CNT=='United Kingdom') %>%
  filter(ST004D01T=='Female')

# The conditions to do a t-test include that the data are normally distributed
# and there is homogeneity (similarity) of the variances (the squared standard deviations)
# Let us check the conditions are met by calculating first if the data sets are normally
# distributed using the Pearson test of normality from the nortest package

pearson.test(as.numeric(MaleUK$PV1MATH))
pearson.test(as.numeric(FemaleUK$PV1MATH))

# The p-values are over 0.05 so both distriburtions are normal
# Pearson chi-square normality test
#
# data:  as.numeric(MaleUK$PV1MATH)
# P = 75.714, p-value = 0.1936
# Pearson chi-square normality test
#
# data:  as.numeric(FemaleUK$PV1MATH)
# P = 74.06, p-value = 0.2589
#
# We will then check the variances of the two data sets

VarM<-var(MaleUK$PV1MATH)
VarF<-var(FemaleUK$PV1MATH)
VarM/VarF

# The variance ratio is close to 1 (1.1)
# So our two conditions are met and can we can perform the t-test

t.test(MaleUK$PV1MATH, FemaleUK$PV1MATH)

# The p-value is <0.05 (4.061e-08) suggesting there are statistically
# differences between boys and girls
```
