---
title: "Quantitative methods in the context of education research"
subtitle: "MA Education courses, School of Education, Communication & Society, King's College London"
execute:
  cache: false
editor:
  mode: source
date: "Last compiled on `r format(Sys.time(), '%d %B, %Y')`"

author:
  - name: Peter EJ Kemp
    orcid: 0000-0003-1131-0787
    email: peter.kemp@kcl.ac.uk
    # affiliations:
    #   - name: King's College London
  - name: Richard Brock
    orcid: 0000-0003-2497-8030
    email: richard.brock@kcl.ac.uk
    # affiliations:
    #   - name: King's College London
  - name: Amy O'Brien
    orcid: 0000-0003-3846-6538
    email: amy.l.o'brien@kcl.ac.uk
    # affiliations:
    #   - name: King's College London
  - name: Carla Finesilver
    orcid: 0000-0003-0136-7536
    email: carla.finesilver@kcl.ac.uk
    # affiliations:
    #   - name: King's College London
  - name: Lulu Healy
    orcid: 0000-0003-1698-5134
    email: Lulu.healy@kcl.ac.uk
    # affiliations:
    #   - name: King's College London
  - name: David Pepper
    orcid: 0000-0002-2228-7677
    email: david.pepper@kcl.ac.uk
    # affiliations:
    #   - name: King's College London

---

```{r setup}
    #| eval: true
    #| warning: false
    #| echo: false

    # options(max.print=10000)
    options(max.print=15)
    
  # only load this code if testing this file independently of overall project
  if(!exists("test")){
    warning("RUNNING LOCAL FILE LOAD")

    library(glue)
    node <- Sys.info()
	if(node["nodename"] == "KCL8KWNXY2"){
	  datafolder <- r"(C:\\Users\\w1926273\\OneDrive - King\'s College London\\Code\\SCARI\\)"
	  loc_amy <- r"(G:\\My Drive\\Kings\\Code\\MASTEMR\\data\\)"
	}else if(node["nodename"] == "PETES_PC"){
    pisa_data <- "C:/Users/peter/OneDrive - King\'s College London/Code/QERKCL_PISA/data/pisa/"
    base_location <- "C:/Users/peter/OneDrive - King\'s College London/Code/MASTEMR/"
	  proj_folder <- r"(C:/tmp/MASTEMR/)"
	}else if(node["nodename"] == "KCLMLVXD9XH6P"){
	  pisa_data <- "/Users/k1926273/Library/CloudStorage/OneDrive-King\'sCollegeLondon/Code/QERKCL_PISA/data/pisa/"
    base_location <- "/Users/k1926273/Library/CloudStorage/OneDrive-King\'sCollegeLondon/Code/MASTEMR/"
    proj_folder <- "/Users/k1926273/Desktop/MASTEMR/"
	}else if(node["nodename"] == "KCLJ7J6LFC74J"){ #Richard"s Mac
	  datafolder <- r"(/Users/k1765032/Google Drive/My Drive/PISR/data/PISA/)"
	  pisa_data <- r"[/Users/k1765032/Library/CloudStorage/OneDrive-King\'sCollegeLondon/QERKCL_PISA/data/pisa/]"
	  base_location <- "/Users/k1765032/Library/CloudStorage/OneDrive-King\'sCollegeLondon/Peter Kemp\'s files - MASTEMR/"
	  proj_location <- "/Users/k1765032/Library/CloudStorage/OneDrive-King\'sCollegeLondon/Peter Kemp\'s files - MASTEMR/"
	  proj_folder <- r"[/Users/k1765032/Documents/Teaching/STEM MA/Quantitative module/R/Temp/MASTEMR/]"
	}else if(node["nodename"] == "KCL3Y7NKR3"){ #Pete"s NEW Laptop
	  datafolder <- "G:/My Drive/Kings/Code/PISR/Data/PISA/)"
	  pisa_data <- "C:/Users/k1926273/OneDrive - King\'s College London/Code/QERKCL_PISA/data/pisa/)"
	}else if(node["nodename"] == "Amys-MacBook-Air.local"){ #Pete"s NEW Laptop
	  datafolder <- r"(/Users/amyobrien/Google Drive/My Drive/Data/PISA/)"
	  loc_amy <- r"(/Users/amyobrien/Google Drive/My Drive/MASTEMR/data/)"
	}else{
	  warning("Can\'t find PC name, using default folder location")
	  datafolder <- "D:/OneDrive/KCL OneDrive/OneDrive - King\'s College London/Code/SCARI/"
	}
	
	loc_amy <- glue("{base_location}data/")
    
    source(glue("{proj_folder}MASTEMR_functions.R"))
  }
      source(glue("{proj_folder}MASTEMR_functions.R"))
      source(glue("{proj_folder}MASTEMR_setup_libs.R"))
      source(glue("{proj_folder}MASTEMR_setup.R"))
      # OpenAnalysis_load_ALL(use_cache = TRUE, 
      #                       report_cache_folder=r\'[",MASTEMR_cache,"]\',
      #                       over_write=FALSE)

```

```{r}
#| error: false
#| eval: false
#| echo: false
```

```{r}
#| eval: true
#| echo: false
#| warning: false
PISA2018subset <- PISA_2022 %>%
  select(CNT, PV1MATH, PV1READ, PV1SCIE, HOMEPOS)%>%
  group_by(CNT)%>%
  summarise(Science=mean(PV1SCIE), Mathematics=mean(PV1MATH), Reading=mean(PV1READ), HOMEPOS=mean(HOMEPOS, na.rm=TRUE),
            n=n())

PISA2018subset<-pivot_longer(PISA2018subset, cols=c("Science","Reading","Mathematics"),names_to = "Subject",values_to = "Scores")

PISA2018subset<-PISA2018subset%>%
  mutate(label=ifelse(Scores>550, as.character(CNT),
                      ifelse(Scores<350, as.character(CNT),"")))

ggplot(PISA2018subset, aes(x=HOMEPOS, y=Scores, size=n))+
  geom_point(aes(colour=CNT), alpha=0.6)+
  geom_smooth(formula = 'y ~ x', method="lm", colour="darkgrey", size=0.5)+
  facet_grid(.~Subject)+
  xlab("Home possessions")+
  ylab("Score")+
  ggtitle("PISA 2022 test scores against home possessions")+ 
  theme_bw() +
  theme(legend.position = "none") +
  geom_label_repel(label=PISA2018subset$label, label.size = NA,
                   fill = alpha(c("white"),0.0))

```

This book has been written to accompany the MA STEM quantitative methods course at King's College London. It will take the reader through the process of learning the R statistical programming language through the analysis of the **2022 PISA survey data**. The book does not assume any prior knowledge and should suit the needs of those looking to learn R, learn statistics, learn how to analyse PISA data, or any combination of these. 

# Copyright and usage

This book is free to use under a [CC BY-NC-ND 4.0](https://creativecommons.org/licenses/by-nc-nd/4.0/deed.en) license.

All PISA products are published under the [Creative Commons Attribution-NonCommercial-ShareAlike 3.0 IGO](https://creativecommons.org/licenses/by-nc-sa/3.0/igo/) (CC BY-NC-SA 3.0 IGO) license. This includes the `.feather` and `.parquet` formatted datasets linked throughout the text. The PISA 2022 dataset is also available from the [OECD website](https://www.oecd.org/pisa/data/2022database/).

We have tested the software and datasets used in the making of this book, but cannot guarantee that they are free from errors. If you find any errors or have any questions, please contact [peter.kemp\@kcl.ac.uk](mailto:peter.kemp@kcl.ac.uk) and/or [richard.brock\@kcl.ac.uk](mailto:richard.brock@kcl.ac.uk). You are using this book at your own risk and we cannot be held responsible for any loss or damage arising from the use of this book (please don't overwrite your hard disks!).

To reference this book, please use:<!--@MASTEMR2023-->

> Kemp, Peter EJ, Richard Brock, Amy O’Brien, Carla Finesilver, Lulu Healy, and David Pepper. 2023. _Quantitative Methods in the Context of Education Research: The PISA 2022 data set_. London, United Kingdom: King’s College London. https://peterejkemp.github.io/v3/.

And to reference the PISA 2022 dataset, please use:<!-- @OECD2022results. -->

> OECD. 2022. _PISA 2022 Results (Volume i)_. OECD. https://www.oecd-ilibrary.org/docserver/53f23881-en.pdf.