---
title: "5 Gender differences"
# bibliography: references.bib
---

```{r}
#| eval: true
#| warning: true
#| echo: false

# only load this code if testing this file independently of overall project
if(!exists("test")){
  warning("RUNNING LOCAL FILE LOAD")
  
  library(glue)
  proj_folder <- "C:/Users/Peter/Google Drive/Kings/Code/MASTEMR/"
  proj_folder <- r"(/Users/k1765032/Google Drive/My Drive/MASTEMR/)"
  
  node <- Sys.info()
  if(node["nodename"] == "KCLJ7J6LFC74J"){ #Richard's PC
    proj_folder <- r"(/Users/k1765032/Google Drive/My Drive/MASTEMR/)"
  }else if(node["nodename"] == "KCL3Y7NKR3"){ #Pete's NEW Laptop
    proj_folder <- r"(G:\My Drive\Kings\Code\MASTEMR\)"
  }

  source(glue("{proj_folder}MASTEMR_setup.R"))
}
```

## Statistical analysis

There are probably statistical libraries in R to do every sort of test you will ever need, from the typical ANOVA to cutting edge machine learning. The full list of R packages sits on the [cran server](https://cran.r-project.org/web/packages/available_packages_by_name.html) and you can load packages as and when you need them at no cost. R comes pre-packaged with some common statistical tools, for example, the `t.test()` and linear model regression `lm()`. For other stats tools you'll need to install the package and load it (see @sec-load-run-pckges) before you can use the functions.

# T-tests

## Task 1

```{r}
#| eval: false
#| code-fold: true
#| code-summary: "Show the code"



```

## Task 2

```{r}
#| eval: false
#| code-fold: true
#| code-summary: "Show the code"



```

## Task 3

```{r}
#| eval: false
#| code-fold: true
#| code-summary: "Show the code"



```

### Doing t-tests in R

<iframe width="560" height="315" src="https://www.youtube.com/embed/KdhAcLPq5B4?start=3" title="YouTube video player" frameborder="0" allow="accelerometer; autoplay; clipboard-write; encrypted-media; gyroscope; picture-in-picture" allowfullscreen>

</iframe>

You can find the code from the video below:

```{r}
#| eval: false
#| code-fold: true
#| code-summary: "Show the code"

# Introduction to t-tests in R
#
# Download data from /Users/k1765032/Library/CloudStorage/GoogleDrive-richardandrewbrock@gmail.com/.shortcut-targets-by-id/1c3CkaEBOICzepArDfjQUP34W2BYhFjM4/PISR/Data/PISA/subset/Students_2018_RBDP_none_levels.rds
# You want the file: Students_2018_RBDP_none_levels.rds
# and place in your own file system
# change loc to load the data directly. Loading into R might take a few minutes
install.packages("nortest")

library(tidyverse)
library(nortest)
loc <- "/Users/k1765032/Library/CloudStorage/GoogleDrive-richardandrewbrock@gmail.com/.shortcut-targets-by-id/1c3CkaEBOICzepArDfjQUP34W2BYhFjM4/PISR/Data/PISA/subset/Students_2018_RBDP_none_levels.rds"
PISA_2018 <- read_rds(loc)

# Are there differences between the mean scores of UK boys and girls in PISA mathematics?
#
# Select the gender (ST004D01T) and math score columns (PV1MATH)
# Filter the data to select UK responses

MaleUK<-PISA_2018 %>%
  select(CNT,ST004D01T, PV1MATH) %>%
  filter(CNT=='United Kingdom') %>%
  filter(ST004D01T=='Male')

FemaleUK<-PISA_2018 %>%
  select(CNT,ST004D01T, PV1MATH) %>%
  filter(CNT=='United Kingdom') %>%
  filter(ST004D01T=='Female')

# The conditions to do a t-test include that the data are normally distributed
# and there is homogeneity (similarity) of the variances (the squared standard deviations)
# Let us check the conditions are met by calculating first if the data sets are normally
# distributed using the Pearson test of normality from the nortest package

pearson.test(as.numeric(MaleUK$PV1MATH))
pearson.test(as.numeric(FemaleUK$PV1MATH))

# The p-values are over 0.05 so both distriburtions are normal
# Pearson chi-square normality test
#
# data:  as.numeric(MaleUK$PV1MATH)
# P = 75.714, p-value = 0.1936
# Pearson chi-square normality test
#
# data:  as.numeric(FemaleUK$PV1MATH)
# P = 74.06, p-value = 0.2589
#
# We will then check the variances of the two data sets

VarM<-var(MaleUK$PV1MATH)
VarF<-var(FemaleUK$PV1MATH)
VarM/VarF

# The variance ratio is close to 1 (1.1)
# So our two conditions are met and can we can perform the t-test

t.test(MaleUK$PV1MATH, FemaleUK$PV1MATH)

# The p-value is <0.05 (4.061e-08) suggesting there are statistically
# differences between boys and girls
```
