---
title: "7 Inquiry learning in science"
---

```{r}
#| eval: true
#| warning: true
#| echo: false

# only load this code if testing this file independently of overall project
if(!exists("test")){
  warning("RUNNING LOCAL FILE LOAD")
  
  library(glue)
  proj_folder <- "C:/Users/Peter/Google Drive/Kings/Code/MASTEMR/"
  proj_folder <- r"(/Users/k1765032/Google Drive/My Drive/MASTEMR/)"
  
  node <- Sys.info()
  if(node["nodename"] == "KCLJ7J6LFC74J"){ #Richard's PC
    proj_folder <- r"(/Users/k1765032/Google Drive/My Drive/MASTEMR/)"
  }else if(node["nodename"] == "KCL3Y7NKR3"){ #Pete's NEW Laptop
    proj_folder <- r"(G:\My Drive\Kings\Code\MASTEMR\)"
  }

  source(glue("{proj_folder}MASTEMR_setup.R"))
}
```

<!--https://www.r-bloggers.com/2022/01/how-to-perform-eta-squared-in-r/-->

```{r}
#| code-fold: true
#| code-summary: inquiry preparation code
#| eval: true

# mapping used
# c(1 = "In all lessons",
#  2 = "In most lessons",
#  3 = "In some lessons",
#  4 = "Never or hardly ever",
#  NA = "Valid Skip",
#  NA = "Not Reached",
#  NA = "Not Applicable",
#  NA = "Invalid",
#  NA = "No Response")

inquiry_data <- PISA_2015 %>%
  select(ST098Q10NA, ST098Q07TA, ST098Q05TA, 
         ST098Q02TA, PV1SCIE, IBTEACH, TDTEACH) %>%
  mutate(ST098Q10NA = as.numeric(ST098Q10NA),
         ST098Q07TA = as.numeric(ST098Q07TA),
         ST098Q05TA = as.numeric(ST098Q05TA),
         ST098Q02TA = as.numeric(ST098Q02TA)) %>%
  drop_na() %>%
    mutate(inquiry_components =  ifelse(
                ST098Q10NA > 2 & ST098Q07TA > 2 & ST098Q05TA > 2 & ST098Q02TA > 2, "N",
              ifelse(
                ST098Q10NA > 2 & ST098Q07TA > 2 & ST098Q05TA > 2 & ST098Q02TA < 3, "A",
              ifelse(
                ST098Q10NA > 2 & ST098Q07TA > 2 & ST098Q05TA < 3 & ST098Q02TA < 3, "C",
              ifelse(
                ST098Q10NA > 2 & ST098Q07TA < 3 & ST098Q05TA < 3 & ST098Q02TA < 3, "I",
              ifelse(
                ST098Q10NA < 3 & ST098Q07TA < 3 & ST098Q05TA < 3 & ST098Q02TA < 3, "Q",
              99)))))) %>%
  mutate(inquiry_components = as.factor(inquiry_components)) %>%
  filter(inquiry_components != "99")
```

## Running the test

```{r}

res_aov <- aov(PV1SCIE ~ inquiry_components,
               data = inquiry_data)

summary(res_aov)

TukeyHSD(res_aov)

```

### Reporting the test

To put the ANOVA and Tukey tests in a format for reporting we can use the `easystats` package. Easystats is a collection of tools to help you statistically analyse data and present your findings. One you have installed and loaded `easystats` you can then use the `report(<stats_model>)` command to format your results for a paper you are writing:

```{r}
install.packages("easystats")
library("easystats") # Load the package every time you want to use these tools

report(res_aov)
```

### Plotting the results

-   Explain how to understand this
-   Could we have a ggplot version of this?

```{r}
plot(TukeyHSD(res_aov))
```

### Post tests function

Explanation needed here

```{r}
library(multcomp)
# Tukey HSD test:
post_test <- glht(res_aov,
                  linfct = mcp(inquiry_components = "Tukey"))
summary(post_test)
```

### Eta squared

Explanation needed here

```{r}
library(lsr)
etaSquared(res_aov)
```

## Seminar tasks

### Activity 1: run ANOVA tests exploring...

Need to set something simple for them to do here

```{r}
# code solution to go here


```

### Activity 2: prepare for and run Tukey HSD post hoc test with summary and plot of which contrasts between levels are statistically significant

```{r}
# code solution to go here


```

### Activity 3

```{r}
# code solution to go here


```
