---
# title: "Questions and Answers"
---

```{r}
#| eval: true
#| warning: true
#| echo: false

# only load this code if testing this file independently of overall project
if(!exists("test")){
  warning("RUNNING LOCAL FILE LOAD")
  
  library(glue)
  proj_folder <- "C:/Users/Peter/Google Drive/Kings/Code/MASTEMR/"
  proj_folder <- r"(/Users/k1765032/Google Drive/My Drive/MASTEMR/)"
  
  node <- Sys.info()
  if(node["nodename"] == "KCLJ7J6LFC74J"){ #Richard's PC
    proj_folder <- r"(/Users/k1765032/Google Drive/My Drive/MASTEMR/)"
  }else if(node["nodename"] == "KCL3Y7NKR3"){ #Pete's NEW Laptop
    proj_folder <- r"(G:\My Drive\Kings\Code\MASTEMR\)"
  }else if(node["nodename"] == "PETES_PC"){ #Pete's NEW Laptop
    proj_folder <- r"(C:\Users\Peter\Google Drive\Kings\Code\MASTEMR\)"
  }
  
  source(glue("{proj_folder}MASTEMR_functions.R"))
  OpenAnalysis_load_ALL(use_cache = TRUE, 
                        report_cache_folder=r"[C:\temp\cache\MASTEMR\]",
                        over_write=FALSE)
}
```

## Questions about R {#sec-QANDA}

**Why doesn't my `select`/`filter` statement work?**

When you are loading packages, sometimes different packages have the same function names in them, and the functions themselves will do very different things. For example, there is a `select` function in the `tidyverse`, but also *another* `select` function in the package `MASS` that does something very different. If we load the tidyverse before loading MASS, then the MASS version of select is the one that will be used?!

```{r}
#| echo: false
#| eval: false
#| warning: false
# if("package:broom" %in% search()) detach("package:broom", unload=TRUE)
# if("package:tidyr" %in% search()) detach("package:tidyr", unload=TRUE)
# if("package:dbplyr" %in% search()) detach("package:dbplyr", unload=TRUE)
# if("package:gt" %in% search()) detach("package:gt", unload=TRUE)
# if("package:dplyr" %in% search()) detach("package:dplyr", unload=TRUE)
# 
# if("package:multcomp" %in% search()) detach("package:multcomp", unload=TRUE)
# if("package:TH.data" %in% search()) detach("package:TH.data", unload=TRUE)
# if("package:MASS" %in% search()) detach("package:MASS", unload=TRUE)
# if("package:see" %in% search()) detach("package:see", unload=TRUE)
# if("package:stats" %in% search()) detach("package:stats", unload=TRUE)
# remove all packages
while(!is.null(sessionInfo()$loadedOnly)){
  lapply(names(sessionInfo()$loadedOnly), require, character.only = TRUE)
  invisible(lapply(paste0('package:', names(sessionInfo()$otherPkgs)), detach, character.only=TRUE, unload=TRUE, force=TRUE))    
}

```

```{r}
#| error: true
#| eval: false
library(tidyverse)
library(MASS)

diamonds %>% select(carat, cut, color)
```

To get around this make sure that you load the `tidyverse` *after* `MASS`, in fact you should always load the `tidyverse` last.

```{r}
#| error: true
library(MASS) 
library(tidyverse)

diamonds %>% select(carat, cut, color)
```

you can also specify the package that select comes from (in this case from a package within the `tidyverse` called `dplyr`):

```{r}
#| error: true
diamonds %>% dplyr::select(carat, cut, color)
```

**How can I unload packages**

If you are finding yourself with a conflict as mentioned above and want to `unload` packages, then you need to run the following code:

```{r}
#| eval: false
# adapted from: @mmfrgmpds https://stackoverflow.com/questions/7505547/detach-all-packages-while-working-in-r
while(!is.null(sessionInfo()$loadedOnly)){
  lapply(names(sessionInfo()$loadedOnly), require, character.only = TRUE)
  invisible(lapply(paste0('package:', names(sessionInfo()$otherPkgs)), detach, character.only=TRUE, unload=TRUE, force=TRUE))    
}
```

**Why am I getting the error 'could not find function "%>%"'**

The pipe operator `%>%` is loaded when you load the tidyverse package - make sure you have installed tidyverse and loaded it

```{r}
#| eval: false
install.packages("tidyverse")  # install
library(tidyverse)             # load
```

**I am getting the error: '"Error: Mapping should be created with `aes()` or `aes_()`."' when using ggplot**

This may be caused by having a bracket after the geom rather than before it

```{r}
#| eval: false

data<-data.frame(x=5:1,
                 y=10:6)
ggplot(data, aes(x, y) +           # Reproduce error message
         geom_point())

```

```{r}
#| eval: false

data<-data.frame(x=5:1,
                 y=10:6)
ggplot(data, aes(x, y)) +           # Fixed error by moving bracket
         geom_point()

```

## About the coursework

?filter
