---
title: "Twitter"
---

```{r setup}
library(tidyverse)
library(arrow)
library(glue)

fld <- "G:\\My Drive\\junk\\"

# load the twitter dataset
# twts <- read_parquet(glue("{fld}edutweets.parquet"))
# 
# lng <- nrow(twts)
# 
# tmp <- map(names(tweets), \(x){
# 
#   # x <- "lon"
#   x <- sym(x)
#   missing <- twts %>% filter(is.na({{x}})) %>% pull({{x}}) %>% length
#   present <- lng - missing
#   message(x, " : ", present, "/", lng, " (", round(100*present / lng, 1), "%)")
# })
# 
# 
# tweets <- twts %>% 
#   select(-any_of(c("id_str", "truncated", "in_reply_to_screen_name", "in_reply_to_status_id_str", "in_reply_to_user_id_str", "place_type", "place_name", "place_id", "lat", "lon", "expanded_url", "time_zone", "user_lang", "utc_offset", "country_code", "user_url", "geo_enabled", "protected", "favorited", "retweeted", "name"))) %>% 
#   filter(!is.na(text)) %>%
#   mutate(place_lon = signif(place_lon, 3),
#          place_lat = signif(place_lat, 3)) %>%
#   mutate(source_present = !is.na(source),
#         source = str_extract(source, 
#                 "(iPhone|Android|Twitter Web App|iPad|TweetDeck$|Hootsuite|Buffer|twittbot|SocialFlow|CareerArc|SocialNewsDesk|IFTTT|dlvr|Sprout Social|TweetDeck Web App)"),
#         source = ifelse(is.na(source), ifelse(source_present, "Other", "Missing"), source)) %>%
#   select(-source_present)
#     
# write_parquet(tweets, glue("{fld}edutweets_smaller.parquet"))

# tmp <- tweets %>% group_by(user_id_str) %>% count() %>% arrange(desc(n))
# tmp <- tweets %>% group_by(name) %>% count() %>% arrange(desc(n))

fld <- "G:\\My Drive\\junk\\"
fld <- "C:\\Users\\Peter\\Google Drive\\junk\\"
tweets <- read_parquet(glue("{fld}edutweets_smaller.parquet"))

users <- unique(tweets$user_id_str) %>% length

# location data
missing <- tweets %>% filter(is.na(place_lon)) %>% pull(place_lon) %>% length
present <- lng - missing
missing_per <- round(100*present / lng, 1)
# message(x, " : ", present, "/", lng, " (", round(100*present / lng, 1), "%)")
```

# Loading the dataset
The twitter dataset includes `r nrow(tweets)` tweets by `r users` users that include the word "teacher", "educator", "teach" or "teaching". This data covers anyone who tweeted in a short timeframe in April 2023.

# Where are the tweets coming from

First note that not everyone provides a location for their tweets, the actual number is quite low `r paste0(present, "/", lng, " (",missing_per,"%)")` this is hardly representative of all tweets happening. The data includes columns that tell you about where the tweets come from, including: `place_lon`, `place_lat`, `country`, `location`, `full_name`. Using the longitude and latitude data we can plot the `r present` tweeters:

```{r}
ggplot(data=tweets) + #%>% head(800000)) +
  geom_point(aes(x=place_lon, y=place_lat))
```

To get a table of countries we can get:

```{r}
tweets %>% filter(!is.na(country)) %>%
  group_by(country) %>% 
  count() %>% 
  arrange(desc(n)) %>% gt()
```

``` question
Using the twitter fields for location (`place_lon`, `place_lat`, `country`, `location`, `full_name`), find out which city had the most tweets
```

# Use of language

R has a range of corpus analysis tools, we are going to be looking at `quanteda` (you'll need to install it)

```{r}
#| error: false
#| eval: false
#| echo: false

library(quanteda)

# Most common words used
corp_tweets <- corpus(tweets$text)  # build a new corpus from all the tweets
summary(corp_tweets)

# add other data to the corpus
docvars(corp_tweets, "fav") <- tweets$favorite_count
docvars(corp_tweets, "retweets") <- tweets$retweet_count
docvars(corp_tweets, "followers") <- tweets$follower_count

summary(corp_tweets)

# we need to save this summary dataframe so we can plot the data
tweet_tokens <- summary(corp_tweets, n=length(tweets$favorite_count))

ggplot(data = tweet_tokens %>% head(50000), 
       aes(x = Tokens, y = retweets)) +
    geom_point(alpha=0.1)
```

# most common two word phrases

To turn our dataset into tokens that we can group, count and analyse relationships between, we use the `tokens` command

```{r}
#| error: false
#| eval: false
#| echo: false

data_tokens<- tokens(corp_tweets)
```

We can now start to analyse the tokens. As this dataset was built to look at tweets about teaching, don't be surprised to see some of words around teaching. `tokens`, by default, gives you everything back, including emojis and words such as `a` and `and`. To show all the tokens, use:


```{r}
#| error: false
#| eval: false
#| echo: false

We can now 
kwic(data_tokens, pattern = "terror")

# remove unnecessary tokens
simple_data_tokens <- tokens(data_tokens, remove_numbers = TRUE, remove_punct = TRUE)

# make a matrix of all the word frequencies
simple_data_tokens_dfm <- simple_data_tokens %>% dfm()

topfeatures(simple_data_tokens_dfm, 20)

```



```{r}
#| error: false
#| eval: false
#| echo: false

simple_data_tokens <- tokens_remove(simple_data_tokens, stopwords("en"))

simple_data_tokens_dfm <- simple_data_tokens %>% dfm()

topfeatures(simple_data_tokens_dfm, 20)



```

# does popularity mean you swear more?


# Use of emojis