---
title: "03 Descriptive Statistics"
---

```{r setup}
    options(max.print=15)

    library(here)
    source(here::here("MASTEMR_page_setup.R"))
```

# Pre-reading and pre-seminar task

Before the session, please read: @davis2013introduction [Link to chapter](https://ieas-szeged.hu/downtherabbithole/wp-content/uploads/2018/02/Lennard-J.-Davis-ed.-The-Disability-Studies-Reader-Routledge-2014.pdf#page=15)

**Getting set up**

Ensure you have the PISA 2022 data frame loaded. If you can see the `PISA_2022` data frame in your environment window (at the top right of your screen), there is no need to reload.

```{r}
#| eval: false
#| echo: true
#| code-fold: false
#| code-summary: "Loading the data"
# Load the PISA data.

library(arrow)
library(tidyverse)

PISA_2022 <- read_parquet(r"[<folder>PISA_2022_student_subset.parquet]")
```

# Descriptive statistics

## Using the command line for descriptive statistics

For further reading on descriptive statistics see [chapter 5](https://learningstatisticswithr.com/book/descriptives.html) of Navaro's Learning Statistics with R.

::: callout-tip
We are going to focus on the following variables in the PISA_2022 data frame:

-   ***CNT*** the country of the student.

-   ***HOMEPOS*** is a self-reported measure of a student's wealth, linked to the number of possessions students report having in their home (e.g. books, computers, cars, phones etc.). It is a numeric variable, with a mean of -0.447, minimum of -10.07 and a maximum of 15.24.

-   ***ESCS*** is the index of economic, social and cultural status. It might be thought of as a measure of economic and social status (with some cultural capital measures included). It is a numeric variable, with a mean of -0.310, minimum of -6.84 and a maximum of 7.38. It is constructed from three items: highest parental occupation (`HISEI`), highest level of parental education (`PARED`), and home possessions (`HOMEPOS`), including books in the home

-   ***PV1MATH, PV1SCIE, PV1READ*** are the plausible value scores for achievement tests in mathematics, science and reading, respectively. The full achievement tests are long, so each student only completes a subset of items (which still takes 2 hours). Statistical models are then used to calculate an overall score, based on the students' answers to the subset of questions, as if students had answered all the questions. Ten different approaches (ten different statistical models, that take different approaches to estimating the overalls score are used) to calculating a representative scores, plausible values are used, leading to ten different plausible values. In this course, will just use the first plausible value (PV1). This differs from the PISA recommendation for using the scores, but simplifies things for teaching. For more on plausible values see: [What are plausible values?](https://peterejkemp.github.io/#sec-PV)

-   ***ST004D01T*** is the gender variable and and can take the values: `Male`, `Female` or `NA`.
:::

The simplest way to find information about a data frame is to use the console. You can type commands to find out about a data frame directly into the console. To preform an action on a particular column (also called a vector), we use the `$` symbol. For example, to refer to country data (which is in the vector `CNT`) we would use `PISA_2022$CNT`

In the command line, if you want to find the mean of all the repsonses to the `HOMEPOS` (Home Possessions, a proxy for wealth) item you can type the following:

> mean(PISA_2022\$HOMEPOS)

Notice that you get this response: \[1\] NA. An NA in the data frame can occur for a number of reasons, for example it may indicate a response is missing or incomplete, hence the mean can't be calculated. To tell R to ignore `NA`s, we add `na.rm = TRUE` to a function:

> mean(PISA_2022\$HOMEPOS, na.rm = TRUE)

You can use the command line with a number of functions to find useful information about a data frame. R has a number of standard functions that might be useful for descriptive statistics. To find out details about data frames, you can use:

-   `nrow()` finds the number of rows (e.g. `nrow(PISA_2022)`)
-   `ncol()` finds the number of columns (e.g. `ncol(PISA_2022)`)
-   `names()` finds the names of all the columns (e.g. `names(PISA_2022)`)

If you are working on individual columns, e.g. `max(PISA_2022$PV1SCIE)`, you can use:

-   `mean()` - finds the arithmetic mean
-   `median()` - finds the median value
-   `min()` - finds the minimum value
-   `max()` - finds the maximum value
-   `sd()` - finds the standard deviation
-   `range()` - finds the range of values
-   `length()`- finds the number of items
-   `unique()` - finds the unique items

::: callout-tip
Maybe surprisingly, there is no function to calculate the *mode* in the `tidyverse` package. However, you can get one by loading the `modeest` package and using the most frequent value (`mfv`) function.

```{r}
#| eval: true
#| warning: false
#| echo: true
# Install the modeest package to calculate a mode

library(modeest)
library(tidyverse)

# The mode can be found with the most frequent value (mfv) function
# We can look at about number of books in the home (ST255Q01JA)
# Then use mfv to find the mode value (note the na.rm=TRUE to avoid NAs)

mfv(PISA_2022$ST255Q01JA, na.rm = TRUE)

# In the whole dataframe, the mode number of books is 26-100 books

```
:::

::: callout-tip
You can read a complete list of the items in PISA here: [PISA 2022 student survey item descriptors](https://docs.google.com/spreadsheets/d/1POM1kimLFIk88AZtWnTKMiTBFDwaWHTZ?rtpof=true&usp=drive_fs)
:::

A useful way to get a quick summary of what is in a `data.frame` is, the `summary` command. This command outputs the minimum, median, mean, maximum (and 1st and 3rd quartile values, i.e. the values at 25% and 75% of the range). For example, to get a sense of the science score variable (`PV1SCIE`) we can use:

```{r}
#| echo: true
#| eval: true
summary(PISA_2022$PV1SCIE)
```

# Filtering Data frames

We will now learn how to calculate means of subgroups of data frames, using the `filter` and `summarise` functions. We can use `filter` to focus on only a subset of our `data.frame`. For example, below, we can use `filter` to focus only on responses from UK students.

::: callout-tip
Note, in R the `=` and `==` operators have slightly different meanings. `=` can be used to assign a value, for example to set x to 10.

```{r}
#| echo: true
#| eval: true
x = 10
print(x)

```

By contrast, when checking is two items are equal, use `==`

```{r}
#| echo: true
#| eval: true
x <- data.frame(Colour=c("Red", "Green", "Blue", "Yellow"))
                # Create a data frame of colours

x <- x %>%
  filter(Colour=="Red") # Check if x is equalt to "Red" not the use of ==

print(x)

```
:::

In line 4 below, we `filter` for UK responses. Note that, in `filter` we use `==` rather than `=`.

Then we use `summarise` to calculate the means of the variables we are interested in, for example, students science (`PV1SCIE`), mathematics (`PV1MATH`) and reading scores. We can also find the total number of students entered in the UK, using `n()`, which counts the number of rows.

```{r}
#| eval: true
#| code-fold: false
#| code-summary: "Loading the data"
#| echo: true
# Selecting only UK responses and finding the mean of various variables

PISA_2022 %>%                         # <1>
  filter(CNT == "United Kingdom") %>% # <2> 
  summarise(MeanSci = mean(PV1SCIE),  # <3>
            MeanMath = mean(PV1MATH), # <4>
            Total = n())              # <5>

```

1.  line 1 passes the whole `PISA_2022` dataset and pipes it into the next line using `%>%`
2.  line 2 `filters` out any results that are not from the UK by finding all the rows where `CNT` equals `==` `"United Kingdom"`. Note the `==` for checking equality. The result is then piped (`%>%`) to the next line
3.  line 3 uses `summarise` to calculate, for the UK, the mean of `PV1SCIE` (the science score) and puts the result in a column called `MeanSci`
4.  line 4 uses `summarise` to calculate, for the UK, the mean of `PV1MATH` (the maths score) and puts the result in a column called `MeanMath`
5.  line 5 uses the `n()` function to count the number of students in the UK sample

::: callout-tip
If you want to filter by a different vector (that is, a different column in the table), don't forget to change the name of the vector in the filter command, for example, to find the mean mathematics and science scores, and total number of pupils who are girls (using the gender variable `ST004D01T`). We change the vector to filer on to `ST004D01T` and the condition to `Female`.

Don't forget to use the `%>%` and the `==` in `filter`!

```{r}
#| eval: true
#| echo: true
PISA_2022 %>% 
 filter(ST004D01T == "Female")%>%
  summarise(MeanSci= mean(PV1SCIE, na.rm=TRUE), 
            MeanMath = mean(PV1MATH, na.rm=TRUE), 
            Total=n())
```

You can add multiple filters by using the `&` operator which means AND. Later on we will also meet `|` (the vertical line symbol), which means OR. So if you want to find the scores of male students in the UK you would use:

```{r}
#| eval: true
#| echo: true
PISA_2022 %>% 
 filter(CNT == "United Kingdom" & ST004D01T == "Male")%>%
  summarise(MeanSci = mean(PV1SCIE), 
            MeanMath = mean(PV1MATH), 
            Total = n())
```
:::

Often we are interested in summary data across multiple subgroups. We can then tell R to `group_by`, for example, `group_by(CNT)`, to get a summary of data for subgroups.

```{r}
#| eval: true
#| echo: true
# Grouping by country and summarising
PISA_2022 %>%                            # <1>
 group_by(CNT) %>%                       # <2>
  summarise(MeanSci = mean(PV1SCIE),     # <3>
            MeanMath = mean(PV1MATH),    # <4>
            Total = n())                 # <5>
```

1.  line 1 passes the whole `PISA_2022` dataset and pipes it into the next line using `%>%`
2.  line 2 uses `group_by` to use the values in `CNT` to group the data - i.e. the calculate the means for each country
3.  line 3 uses `summarise` to calculate the mean of `PV1SCIE` (the science score) and puts the result in a column called `MeanSci`
4.  line 4 uses `summarise` to calculate the mean of `PV1MATH` (the maths score) and puts the result in a column called `MeanMath`
5.  line 5 uses the `n()` function to count the number of students the sample

::: callout-tip
In the console, R will truncate tables, so you might only see the first 10 countries of 80 in the data frame using the code above. One solution to this is to put the results of the summarising into a new data frame (e.g. `summarydata`) which you can then view from the environment window (and use for future processing). To do this you use the assign operator `<-`.

```{r}
#| eval: true
#| echo: true
# Grouping by country and summarising, and asigning to a new data frame
summarydata<-PISA_2022 %>% 
 group_by(CNT)%>% 
  summarise(MeanSci = mean(PV1SCIE), 
            MeanMath = mean(PV1MATH), 
            Total = n())
```
:::

# Creating summary tables and manipulating them

The PISA data frame is large (!) so it can often be helpful to create interim summary tables.

::: callout-tip
When using `summarise`, be careful to add a comma after each function, and check you have closed as many brackets as you open!
:::

A useful function is `table` which creates a summary table of the counts of unique entries in a data frame. For example, we might want to know how many boys and girls there are by country in the whole data set.

In the example below, we will use `select` which creates a subset of a table by columns. For example, if I want to create a table of gender type by country, we need only include two columns from the PISA_2022 data frame: country (`CNT`) and gender (`ST004D01T`). We use the command `select` to focus on those two: `select(CNT, ST004D01T)`.

To create a summary table, intuitively, we use the table function. There are two additional actions we need to do. First, because we piped the whole PISA_2022 data frame, if we apply table, then even though we have filtered by the UK, the data frame retains levels for all the other countries. If we don't remove these levels, we will get a large data frame with many zero entries for the countries we have filtered out. The function droplevels() removes the levels for other countries (i.e. everything other than the United Kingdom). Finally, the output of table is a datatype called (appropriately) a table. Data frames are more easily manipulated so we convert the table into a data frame using as.data.frame(table(SchoolType)).

```{r}
#| eval: true
#| echo: true

# Creating a summary data frame

GenderUK <- PISA_2022 %>%                          # <1>
  select(CNT, ST004D01T) %>%                       # <2>
  filter(CNT == "United Kingdom") %>%              # <3>
  droplevels()                                     # <4>

# Create a summary table
GenderUKSummary <- as.data.frame(table(GenderUK))  # <5>
GenderUKSummary                                    # <6>

```

1.  line 1 creates a new data frame, `GenderUK` into which the results manipulated `PISA_2022` will be placed - `PISA_2022` is piped to the next steps
2.  line 2 uses `select` to select the columns of interest `CNT` (country) and `ST004D01T` (gender)
3.  line 3 uses `filter` to filter for only the entries for the UK
4.  line 4 - as we have filtered for the UK, we would get 0 entires for all the other levels (all the other countries). To stop those 0s confusing our table we use `droplevels()` to ignore the levels we don't need
5.  line 5 - to create summary of the counts, we use `table` to create a summary table. We turn the table into a data frame with `as.data.frame` to make it easier to manipulate
6.  line 6 - print the table

You can open the `GenderUKSummary` data frame and see the summary data. Table has created a new column Freq which stores the results of the counts.

It might now be interesting to know what percentage the counts of genders represent. To achieve that, first we create a variable that is the total number of genders (to calculate the percentage). This variable is total, and we perform a simple sum on the count column - GenderUKSummary\$Freq.

We then use the mutate function. mutate allows you to add a new column to a table. You pipe the data frame to mutate, and begin by giving the name of the new column you want, in this case the percentage of schools of each type, we will call this PerSch mutate(PerSch=. Then we set the value of that column to the percentage calculation: 100\*(Freq/total). The Frequency count for each column will be multiplied by 100 and divided by the total.

```{r}
#| eval: true
#| echo: true

# Adding a percentage column

total <- sum(GenderUKSummary$Freq)       # <1>                                        
GenderUKSummary %>%                      # <2>   
  mutate(PerSch = 100*(Freq / total))    # <3>
```

1.  line 1 - to calculate a percentage, you first need to find the total of the frequency column in `GenderUKSummary`. We use `GenderUKSummary$Freq` to indicate the column and `sum` to find the total\
2.  line 2 - pipe `GenderUKSummary` so we can manipulate it
3.  line 3 - to add a new column (with the percentage in) we use `mutate`. We give the name of the new column `PerSch =` and set out the calculation we want to perform, dividing the frequency in each row, by the total and multiplying by 100: `100*(Freq / total)`

We can also get the same result by using the `sum` function inside `mutate.` Here `PerSch` is calculated for each row, taking the `Freq` value for that row and dividing it by the `sum` of `Freq` for all rows, i.e. calculating the percentage:

```{r}
#| eval: false
# We can also do this by:
GenderUKSummary %>%
  mutate(PerSch = 100*(Freq / sum(Freq)))

```

::: callout-tip
You can use the `round` function to display a given number of decimal places. Here, I have used `round( ,2)` to limit the percentage calculation to two significant figures.

```{r}
#| eval: true
#| echo: true

# Calculating Percentages of students on EHC plans
total=sum(GenderUKSummary$Freq)
GenderUKSummary%>%
  mutate(PerSch = round(100*(Freq / total),2))

```
:::

# Graphing

For more details on graphing, see @sec-graphing

## Bar graphs

Imagine we want to plot a graph of how many digital devices with screens students report in their home `ST253Q01JA` in the UK.

When plotting graphs, it makes things easier to have a `data.frame` of the data you will pass to ggplot - a bit like the final table of data you will actually plot when drawing a graph in real life.

To produce the graph, we will first create a new `data.frame` that we will use in the plot. I have called that `data.frame` `DDplot` (it is data on digital devices (DD) we will plot). To do this, take the `PISA_2022` `data.frame`, pipe it, and select `ST253Q01JA` and `CNT` and filter for the UK.

You have seen piping before. The new element here is using `ggplot`, one of R's graphing functions (you can find more details on how to use `geom_bar` to plot bar graphs, in the section: [Geom_bar)](https://peterejkemp.github.io/#geoms).

To plot a graph, you call ggplot and specify the data you want to use for the graph (in our case, the new `data.frame` we have created, `DDplot`).

The next layer of ggplot is the aesthetics (`aes`), i.e., what our graph will look like. First, we tell ggplot what data we want our `y` axis to represent, in this case the number of digital devices students have in their home `ST253Q01JA` . Finally, we specify we want a bar graph using `geom_bar()`.

```{r}
#| eval: true
#| code-fold: false
#| code-summary: "Show the code"
#| echo: true
# Graphing the numbers of digital devices students have
DDplot <- PISA_2022 %>%                # <1>
  select(CNT, ST253Q01JA) %>%          # <2>
  filter(CNT == "United Kingdom")      # <3>

# use the DDplot data to create a graph
ggplot(data = DDplot,                   # <4>
       aes(x = ST253Q01JA)) +           # <5>
  geom_bar()                          # <6>

```

1.  line 1 - create a new dataframe for plotting `DDplot` - we pipe PISA_2022 into the new data frame
2.  line 2 - `select` the columns we need, `CNT` (country) and `ST253Q01JA` the item on digital devices
3.  line 3 - 'filter' for only UK entries
4.  line 4 - to plot the chart, we pass the data to the plotting function `ggplot`
5.  line 5 - the first step in plotting is specifying the aesthetics (`aes`) - for a bar graph it is simple, we only need to state what is on the x-axis (`x = ST253Q01JA`) - the number of digital devices
6.  line 6 - we specify the geom we want, in this case a bar chart with `geom_bar()`

That is a good first attempt, but there are some aspects we will want to tidy up.

You can modify the graph by adding axis labels and titles using `xlab("label")`, `ylab("label")`, and `ggtitle("title")`.

```{r}
#| eval: true
#| code-fold: false
#| code-summary: "Show the code"
#| echo: true

# use digital device data to create a graph
# Adding axis labels and a title
ggplot(data = DDplot,                                              # <1>
       aes(x = ST253Q01JA)) +                                      # <2>
  geom_bar() +                                                     # <3>
  xlab("How many digital devices do you have?") +                  # <4>
  ylab("Count") +                                                  # <5>
  ggtitle("Counts of availability of digital devices in the UK")   # <6>

```

1.  line 1 - to plot the chart, we pass the data to the plotting function `ggplot`
2.  line 2 - the first step in plotting is specifying the aesthetics (`aes`) - for a bar graph it is simple, we only need to state what is on the x-axis (`x = ST253Q01JA`) - the number of digital devices
3.  line 3 - we specify the geom we want, in this case a bar chart with `geom_bar()`
4.  line 4 - we set the x-axis label with `xlab("")`
5.  line 5 - we set the y-axis label with `ylab("")`
6.  line 6 - we give a titleusing `ggtitle("")`

You can add colour, by specifying that the fill of the bars should be by the value of `IC009Q11NA` in the aesthetics: `aes(x=IC009Q11NA, fill=IC009Q11NA)`.

```{r}
#| eval: true
#| code-fold: false
#| code-summary: "Show the code"
#| echo: true

# Adding axis labels and a title
# And adding colour
ggplot(data = DDplot,
       aes(x = ST253Q01JA, fill = ST253Q01JA)) +                   # <2>
  geom_bar() +
  xlab("How many digital devices do you have?") +
  ylab("Count") +
  ggtitle("Counts of availability of digital devices in the UK")

```

2.  line 2 - to set the colour of the columns, within `aes` we specify the fill should be set by the variable `ST253Q01JA` (the number of digital devices)

To rotate the text on the x-axis you can add the text: `theme(axis.text.x = element_text(angle = 90, vjust = 0.5, hjust=1))`. Inserting `\n` into the title adds a line break to divide the long title. We can turn the legend off with `theme(legend.position = "none")`

```{r}
#| eval: true
#| code-fold: false
#| code-summary: "Show the code"
#| echo: true

# Rotating axis text and removing legend

ggplot(data = DDplot,
       aes(x = ST253Q01JA, fill = ST253Q01JA)) +
  geom_bar() +
  xlab("How many digital devices do you have?") +
  ylab("Count") +
  ggtitle("Counts of availability \n of digital devices in the UK") +     # <6>
  theme(axis.text.x = element_text(angle = 90, vjust = 0.5, hjust = 1)) + # <7>
  theme(legend.position = "none")                                         # <8>
  

```

6.  line 6 - note that the `\n` in the title inserts a line break
7.  line 7 - to rotate the text on the x-axis we use `theme(axis.text.x = element_text(angle = 90, vjust = 0.5, hjust = 1))`
8.  line 8 - we can turn off the legend by adding: `theme(legend.position = "none")`

You make wish to plot multiple series, for example, the interactive whiteboard data reported by boys and girls. In that case, we need to include the gender vector `ST004D01T` in the data frame to plot `DDplot`. To highlight the difference between boys and girls, in the aesthetics we set the fill colour by the gender variable: `fill=ST004D01T`. Finally, we need to tell ggplot to plot the bars side-by-side, rather than stacking them - you do this by specifying `position = position_dodge2()`.

```{r}
#| eval: true
#| code-fold: false
#| code-summary: "Show the code"
#| echo: true
# Graphing the number of digital devices students have by gender

DDplot <- PISA_2022 %>%
  select(CNT, ST253Q01JA, ST004D01T) %>%                              # <2>
  filter(CNT == "United Kingdom")

# Plotting two series - in this case, by gender

ggplot(data = DDplot,
       aes(x = ST253Q01JA, fill = ST004D01T)) +
  geom_bar(position = position_dodge2()) +                           # <6>
  xlab("How many digital devices do you have?") +
  ylab("Count") +
  ggtitle("Counts of availability \n of digital devices in the UK") +
  theme(axis.text.x = element_text(angle = 90, vjust = 0.5, hjust=1))

```

2.  line 2 - to plot two series, we include gender (`ST004D01T`) in the filter
3.  line 6 - the default for `ggplot` is to stack the bars on top of each other, `position = position_dodge2()` tells `ggplot` to plot the bars side by side.

If you take \``position = position_dodge2()` out, ggplot will default to stacking the bars. The long response for no devices (`"There are no <digital devices> with screens."`) is not terribly aesthetically pleasing. I will use `mutate` to change it - `recode` changes all the `"There are no <digital devices> with screens."` to `"None"` in the vector `ST253Q01JA`.

```{r}
#| eval: true
#| code-fold: false
#| code-summary: "Show the code"
#| echo: true
# Graphing the number of digital devices students have
DDplot <- PISA_2022 %>%
  select(CNT, ST253Q01JA, ST004D01T) %>%
  filter(CNT == "United Kingdom")

# Changing the longer label for a shorter for ease of plotting
# Change the long "There are no <digital devices> with screens" to "None"
DDplot<-DDplot %>%
  mutate(ST253Q01JA = recode(ST253Q01JA, "There are no <digital devices> with screens."="None"))

# Plotting two series - in this case, by gender

ggplot(data=DDplot,
       aes(x = ST253Q01JA, fill = ST004D01T)) +
  geom_bar() +
  xlab("How many digital devices do you have") +
  ylab("Count") +
  ggtitle("Counts of availability of digital devices in the UK")

```

You can add text, for example the counts, to the graphs, using `geom_text`. We set the `label` to `after_stat(count)`and `stat=count`. Because the bars have been positioned using `position_dodge`, you need to do the same for the labels. `vjust=-0.5` sets the height of the label over the bar.

```{r}
#| eval: true
#| code-fold: false
#| code-summary: "Show the code"
#| echo: true
#| warning: false
# Graphing the number of students who have digital devices by gender
DDplot <- PISA_2022 %>%
  select(CNT, ST253Q01JA, ST004D01T) %>%
  filter(CNT == "United Kingdom")

# Changing the longer label for a shorter for ease of plotting
DDplot<-DDplot%>%
  mutate(ST253Q01JA = recode(ST253Q01JA, 
                      "There are no <digital devices> with screens." = "None")) # <6>

# Plotting two series - in this case, by gender with text
ggplot(data = DDplot,
       aes(x = ST253Q01JA, fill = ST004D01T)) +
  geom_bar(position = position_dodge2()) +
  xlab("How many digital devices do you have?") +
  ylab("Count") +
  ggtitle("Counts of availability of digital devices in the UK") +
  geom_text(aes(label=after_stat(count)), stat = "count",                              # <13>
            position = position_dodge2(width = 0.9), vjust = -0.5, size = 3)   # <14>
  

```

6.  line 6 - the label `There are no <digital devices> with screens` is too long to fit on the plot. We can use the `recode` function, which switches all entries of `There are no <digital devices> with screens` to `None` which produces a neater plot
7.  line 13 - we use `geom_text` to add labels. Inside the `aes` we specific that we want the label to be counts (`label=..count..`) and tell ggplot to count up the numbers of devices (`stat = "count"`)
8.  line 14 - we use `position_dodge2(width = 0.9)` for the text, because we have used `position_dodge2` for the columns and set the vertical justification (`vjust = -0.5`) and font size (`size = 3`)

## Scatter graphs

To plot a scatter graph we use `geom_point` (see also: [Geom_Point section](https://peterejkemp.github.io/#sec-geom_point)), which works in a similar way to `geom_bar`.

**Example**

Imagine we want to plot a graph of mathematics scores `PV1MATH` (on the x-axis) against science scores `PV1SCIE` (y-axis) for students in the UK.

A above, we first want to create a `data.frame` to plot - in this case we have called it `plotdata`. We `select` only the columns we need (`PV1MATH`, `PV1SCIE`, and `CNT` to `filter` for the UK), and then `filter` for the UK.

We use the ggplot function, specifying `data = plotdata` (i.e. we want to plot the data in the specified data frame). As with `geom_bar`, we can use `ggplot` and first specify the data we want to plot (`ggplot(data = plotdata,`). Next, we set the aesthetic variables (`aes`) - to keep things simple, we will set only the `x` and `y` variables. Then we call `geom_point` to plot the points as a scatter graph.

```{r}
#| eval: true
#| code-fold: false
#| echo: true

# Filter to create a data frame of UK scores in science and mathematics

plotdata<-PISA_2022 %>%
  select(CNT, PV1SCIE, PV1MATH)%>%       # <2>
  filter(CNT == "United Kingdom")        # <3>

# Plot the data as a scatter graph with geom_point()

ggplot(data = plotdata,                  # <4>
       aes(x = PV1MATH, y = PV1SCIE)) +  # <5>
  geom_point()                           # <6>
```

2.  line 2 - select the columns we need for our new `plotdata` data frame, `CNT` (country), `PV1SCIE` (science score), and `PV1MATH` (maths score).
3.  line 3 - filer for UK data
4.  line 4 - pass the `plotdata` we created to `ggplot`
5.  line 5 - set the aesthetics, unlike the bar graph, for a scatter plot we set the variable to plot on the x and y axes (`aes(x = PV1MATH, y = PV1SCIE)`)
6.  line 6 - to plot a scatter plot, we use `geom_point`

We can make things more pleasing by adding more features. For example, we can add colour (`geom_point(colour = "blue")`) and rename the axes `labs(x = "Mathematics score", y = "Science score")`. Note when adding to `ggplot`, the `+` should come at the end of the line before the new addition to avoid an error.

```{r}
#| eval: true
#| code-fold: false
#| echo: true

# Filter to create a data frame of UK scores in science and mathematics
plotdata<-PISA_2022 %>%
  select(CNT, PV1SCIE, PV1MATH)%>%
  filter(CNT == "United Kingdom")

# Plot the data as a scatter graph with geom_point()

ggplot(data = plotdata,
       aes(x = PV1MATH, y = PV1SCIE)) +
  geom_point(colour = "blue") +                          # <6>
  labs(x = "Mathematics Score", y = "Science Score")     # <7>
```

6.  line 6 - inside `geom_point` we set `colour = "blue"` to plot blue points
7.  line 7 - we can use `labs()` to set the x and y axes labels

We can also add a line using (`geom_smooth(method='lm')`) - here `lm` specifies a linear plot (i.e. a straight line).

```{r}
#| eval: true
#| code-fold: false
#| warning: false
#| echo: true

# Filter to create a data frame of UK scores in science and mathematics

plotdata<-PISA_2022 %>%
  select(CNT, PV1SCIE, PV1MATH)%>%
  filter(CNT == "United Kingdom")

# Plot the data as a scatter graph with geom_point()

ggplot(data = plotdata,
       aes(x = PV1MATH, y = PV1SCIE)) +
  geom_point(colour = "blue") +
  labs(x = "Mathematics Score", y = "Science Score") +
  (geom_smooth(method = 'lm'))                           # <8>
```

8.  line 8 - `geom_smooth` is a general function for drawing lines. If we set the `(method - 'lm')` that uses the linear method (lm) giving a straight line.

We can also change the colour of the points by a variable in the data frame - for example by gender (`aes(colour=ST004D01T`) - NB In order to change colour by gender, the gender variable (`ST0040D01T`) must be included in the data set to plot. We can vary the size of points and make them slightly transparent (their alpha level): `geom_point(alpha=0.4, size=0.6)`, and move the legend to the bottom (`theme(legend.position = "bottom")`)

```{r}
#| eval: true
#| code-fold: false
#| echo: true
#| warning: false

# Create a data set of UK maths and science scores including gender

plotdata<-PISA_2022 %>%
  select(CNT, PV1SCIE, PV1MATH, ST004D01T) %>%
  filter(CNT == "United Kingdom")

# Plot the data as a scatter graph with geom_point()
ggplot(data = plotdata,
       aes(x = PV1MATH, y=PV1SCIE, colour = ST004D01T)) +  # <6>
  geom_point(alpha = 0.4, size = 0.6) +
  labs(x="Mathematics Score", y="Science Score") +
  (geom_smooth(method = 'lm')) +
  theme(legend.position = "bottom") +                     # <9>
  theme(legend.title = element_blank())                   # <10>

```

6.  line 6 - set the colour of the points by gender (`colour = ST004D01T`)
7.  line 9 - moves the legend to the bottom of the graph
8.  line 10 - removes the legend title

::: callout-tip
For a summary of all the elements of a graph you can change in `ggplot` - see this [help sheet](https://posit.co/wp-content/uploads/2022/10/data-visualization-1.pdf).
:::

# Seminar activities

## Task 1 - What makes a good graph?

::: question
Consider an example of a good and a bad graph / data representation.

- is the type of graph appropriate for the data? Can you think of a better choice?
- are the data fairly represented? Are there features that skew how the data might be read? (broken axes, different scales on different sub-graphs, improper scaling, missing labels, omitted data, unhelpful 3D features, poor use of colour)
- are there any unnecessary features, labels or data? What could be added? What is missing?
- does the graph tell a clear story? What features allow or prevent a clear message about the data coming across to the reader?

:::

## Task 2 - Using the command line

::: question
-   Using the command line, find out:

a)  The number of students (i.e. the number of rows) in the PISA 2022 data frame
b)  The number of items in our data frame (i.e. the number of columns)
c)  The mean, maximum and minimum science score (don't forget to use `$`)
d)  The unique values of `ST003D02T` - what information do you think this column holds?
:::

```{r}
#| eval: false
#| code-fold: true
#| code-summary: "Answer"
#| echo: true
# Using the command line
# a) Find the number of students (i.e. the number of rows) in the PISA 2022 data frame

nrow(PISA_2022)

# b) The number of items in our data frame (i.e. the number of columns)

ncol(PISA_2022)

# c) The mean, maximum and minimum science score (don't forget to use $)

mean(PISA_2022$PV1SCIE)
max(PISA_2022$PV1SCIE)
min(PISA_2022$PV1SCIE)

# d) The unique values of ST003D02T - what information do you think this column holds?

unique(PISA_2022$ST003D02T)

# This column contains students' birth months
# You can find out the subtitle of columns using

attributes(PISA_2022$ST003D02T)


```

## Task 3 - Using the `summary` function

::: question
-   Using `summary` find:

a)  The maximum and minimum of the HOMEPOS (Wealth) variable
b)  The mean reading score
c)  The minimum science score in the data set
d)  Consider the distributions of the reading and science scores, and comment on any differences.
:::

```{r}
#| eval: false
#| code-fold: true
#| code-summary: "Answer"
#| echo: true
# Using the command line

summary(PISA_2022$HOMEPOS)

summary(PISA_2022$PV1READ)

summary(PISA_2022$PV1SCIE)

```

## Task 4 - Creating summary tables

::: callout-tip
Make sure you have spelled the name of the variables PV1MATH, etc. correctly. They are case sensitive. You can use the function `colnames(PISA_2022)` to get a list of names and copy and paste them.
:::

::: question
-   Find the total number of students who responded in the United States, and their mean science, mathematics and reading scores. Compare that to the responses from the UK. Don't forget to pipe (`%>%`) each step!

-   Filter the data frame for the UK and `group_by` gender (which is `ST004D01T`). Use summarise to find the maximum, minimum and mean scores for boys and girls in mathematics in the UK.

-   Filter the data frame for the UK, the US, and `group_by` gender (which is `ST004D01T`) and country. Use summarise to compare mathematics and science achievement.
:::

```{r}
#| eval: false
#| code-fold: true
#| code-summary: "Answer"
#| echo: true
# Summarising responses in the US and UK and finding means

PISA_2022 %>% 
 filter(CNT == "United Kingdom" | CNT == "United States")%>%
  group_by(CNT)%>%
  summarise(MeanSci = mean(PV1SCIE), 
            MeanMath = mean(PV1MATH),
            MeanRead = mean(PV1READ),
            Total = n())

# Comparing male and female mathematics performance in the UK

PISA_2022 %>% 
 filter(CNT == "United Kingdom")%>%
  group_by(ST004D01T)%>%
  summarise(MeanUKMath = mean(PV1MATH),
            MaxUKMath = max(PV1MATH),
            MinUKMath = min(PV1MATH))

# Comparing male and female mathematics performance in the UK and US

PISA_2022 %>% 
  filter(CNT == "United Kingdom" | CNT== "United States" )%>%
  group_by(ST004D01T, CNT)%>%
  summarise(MeanMath = mean(PV1MATH),
            MaxMath = max(PV1MATH),
            MinMath = min(PV1MATH))

```

::: callout-tip
Don't forget to use the pipe operator `%>%` between each function!
:::

::: question
`WB171Q01HA` asks participants to think of the last time you had a break between classes at school: How did you feel: Happy. For students in France, find out the percentage of students who responded with the different options: Not at all A little Quite a bit Extremely Valid Skip Not Applicable Invalid No Response Missing. (Hint: don't forget to `droplevels()`).
:::

```{r}
#| eval: false
#| code-fold: true
#| code-summary: "Answer"
#| echo: true
# Finding the percentage of students who feel happy between lessons in France

WellData<-PISA_2022%>%
  select(CNT, WB171Q01HA)%>%
  filter(CNT == "France")%>%
  droplevels()

WellData<-as.data.frame(table(WellData))
Total = sum(WellData$Freq)
WellData<-WellData%>%
  mutate(WellData=round((Freq*100 / Total),1))

WellData

```

::: question
`ST251Q06JA` asks students if they have a musical instrument in their home. What percentage of students in the UK have no instruments in their home? What is the percentage in China?
:::

```{r}
#| eval: false
#| code-fold: true
#| code-summary: "Answer"
#| echo: true
# Finding the percentage of students with no musical instruments in the UK and China
# Select the relevant variables, filter for the countries and group - dropping levels to cut unnecessary countries
MusicData<-PISA_2022%>%
  select(CNT, ST251Q06JA)%>%
  filter(CNT == "United Kingdom"|CNT == "B-S-J-Z (China)")%>%
  group_by(ST251Q06JA, CNT)%>%
  droplevels()

# Convert to a data frame

MusicData<-as.data.frame(table(MusicData))

# Find the total number of students to calculate percentages
Total = sum(MusicData$Freq)

# Mutate to add a column with the percentage calculation
MusicData<-MusicData%>%
  mutate(PercComp = round((Freq*100 / Total), 1))
MusicData

```

::: callout-tip
Don't forget to use the pipe operator `%>%` between each function!
:::

## Task 5 - Bar graphs

::: question
-   `ST038Q08NA` asks students in the past 12 months, how often do other students spread nasty rumours about them. Plot a graph of how often students spread rumours in the UK and France.
:::

```{r}
#| eval: false
#| code-fold: true
#| code-summary: "Answer"
#| echo: true

# Select country and question data and Filter for the UK and France
Rumplot <- PISA_2022 %>%
  select(CNT, ST038Q08NA) %>%
  filter(CNT == "United Kingdom"|CNT == "France")

# use rumour data to create a graph
# Set x as the item on culture and fill by the country
# Remember to use position dodge to plot the bars side by side
ggplot(data = Rumplot,
       aes(x = ST038Q08NA, fill = CNT)) +
  geom_bar(position = "dodge2")+
  xlab("Do students spread rumours about you?")+
  ylab("Count")+
  ggtitle("Spreading rumours: France vs. the UK")+
  theme(axis.text.x = element_text(angle = 90, vjust = 0.5, hjust=1))
  # Rotate axis text

```

::: question
-   `PA195Q01JA` asks students how many books are in their home. Plot a bar graph for students in Germany. Then plot a graph splitting the data by gender (`ST004D01T`).
:::

```{r}
#| eval: false
#| code-fold: true
#| code-summary: "Answer"
#| echo: true
# Create a data set related to books in Germany and include gender
Bookplot <- PISA_2022 %>%
  select(CNT, PA195Q01JA, ST004D01T) %>%
  filter(CNT == "Germany")

# use book data to create a graph
ggplot(data = Bookplot,
       aes(x = PA195Q01JA, fill=PA195Q01JA)) +
  geom_bar(position = "dodge2") +
  theme(axis.text.x = element_text(angle = 90, vjust = 0.5, hjust = 1))+ 
  # To rotate the x-axis text
  xlab("Number of books in the home")+
  ylab("Count")+
  ggtitle("Number of books in the home for German students by gender")

# use fill by gender to split the data
ggplot(data = Bookplot,
       aes(x = PA195Q01JA, fill = ST004D01T)) +
  geom_bar(position = "dodge2") +
  theme(axis.text.x = element_text(angle = 90, vjust = 0.5, hjust = 1))+ 
  # To rotate the x-axis text
  xlab("Number of books in the home")+
  ylab("Count")+
  ggtitle("Number of books in the home for German students by gender")

```

::: question
-   Plot a bar graph of the mean mathematics scores of all the countries in the PISA data frame
-   Hint one: you will need to create a summary dataframe using `group_by` and `summarise` and then `geom_bar(stat="identity")`. You use `stat=identity` when you want `geom_bar` to plot the values in the `data.frame` (because you have already summarised) rather than counting the values.
-   Hint two: you can reorder the x-axis with the `reorder` function. Rather than a simple `x=CNT` you can put `x=reorder(CNT, -SciMean)` which will reorder the x axis in descending order (because of the `-` sign) of `SciMean`.
:::

```{r}
#| eval: false
#| code-fold: true
#| code-summary: "Answer"
#| echo: true
# Create a data set of science scores, and use group_by and summarise to create mean scores by country

Sciplot <- PISA_2022 %>%
  select(CNT, PV1SCIE) %>%
  group_by(CNT)%>%
  summarise(MeanSci = mean(PV1SCIE))

# Use geom_bar to plot the data

ggplot(data = Sciplot,
       aes(x = reorder(CNT, -MeanSci), y = MeanSci, fill = MeanSci)) +
  geom_bar(stat = "identity") +
  theme(axis.text.x = element_text(angle = 90, vjust = 0.5, hjust = 1))
```

## Task 6 - Scatter graphs

::: question
-   For students in the UK and Brazil, plot science scores (`PV1SCIE`) by the index of economic, social and cultural status (`ESCS`). (You may have come across this variable in your studies as socioeconomic status, similar to the idea of 'social class'.) Then try varying the colour of points by country, and add a line for each.
:::

```{r}
#| eval: false
#| code-fold: true
#| code-summary: "Answer"
#| echo: true
#| warning: false
plotdata<-PISA_2022 %>%
  select(CNT, PV1SCIE, ESCS)%>%
  filter(CNT == "United Kingdom"|CNT == "Brazil")
# Plot the data as a scatter graph with geom_point()
ggplot(data=plotdata,
       aes(x = ESCS,y = PV1SCIE, colour = CNT)) +
  geom_point(alpha = 0.2, size = 0.6)+
  labs(x = "Index of economic, social and cultural status",
       y = "Science Score")+
  (geom_smooth(method = 'lm'))+
  theme(legend.position = "bottom")+
  labs(CNT = "Country") # Changes ST004D01T to gender for the plot
```

::: question
-   For students in the UK, plot a graph of science scores (`PV1SCIE`) against reading scores (`PV1READ`). Add a straight line and vary the colour of points by students' gender.
:::

```{r}
#| eval: false
#| code-fold: true
#| code-summary: "Answer"
#| echo: true
plotdata<-PISA_2022 %>%
  select(CNT, PV1SCIE, PV1READ, ST004D01T) %>%
  filter(CNT == "United Kingdom")
# Plot the data as a scatter graph with geom_point()
ggplot(data = plotdata,
       aes(x = PV1READ, y = PV1SCIE, colour = ST004D01T)) +
  geom_point(alpha = 0.6, size = 0.6) +
  labs(x = "Reading Score", y = "Science Score") +
  geom_smooth(method = 'lm') +
  theme(legend.position = "bottom")+
  labs(colour = "Gender") # Changes ST004D01T to gender for the plot
```

::: question
-   Challenging task (!): Plot a graph of mean mathematics score (`PV1MATH`) by economic, social and cultural status (`ESCS`) and highlight countries with mathematics scores above 800. An outline of how to achieve this:

a)  Create a data frame of mean `PV1MATH` and `ESCS` by country using `group_by` and `summarise`. (Don't forget to use na.RM=TRUE)
b)  Use `mutate` and `ifelse` to add a new variable, called `text` which contains the names of countires with mathematics scores over 550.
c)  Use `geom_label` to add these data points to the x and y coordinates of the countries e.g. `geom_text_repel(PV1MATHmean, ESCSmean, label)`.
:::

```{r}
#| eval: false
#| code-fold: true
#| code-summary: "Answer"
#| echo: true
#| warning: false

# Create a data frame of ESCS scores and mathematics scores
plotdata<-PISA_2022%>%
  select(CNT, ESCS, PV1MATH)%>%
  group_by(CNT)%>% # group_by country
  summarise(PV1MATHmean = mean(PV1MATH), #calculate means
            ESCSmean = mean(ESCS, na.rm = TRUE))%>%
  mutate(text = ifelse(PV1MATHmean > 550, as.character(CNT), "")) 

# Use mutate to create a newc columns called text
# If PV1MATHmean is over 800 set text to equal CNT, otherwise set it to blank ("")

# Plot the data, putting mean math score on y and ESCS mean on the x
# Set the colour of points by the mean math score

ggplot(plotdata, aes(y=PV1MATHmean, x = ESCSmean, colour = PV1MATHmean))+
  geom_point()+
  geom_text_repel(aes(y = PV1MATHmean, x=ESCSmean, label = text))+
  xlab("Mean ESCS score")+
  ylab("Mean mathematics score")+
  ggtitle("Comparison of mean mathematics and mean ESCS score")+ 
  theme(legend.position = "none") # Hide the legend

```

:::

# Extension tasks

## Task 1 Binning data

Frequency plots are often used to show data following a normal distribution, To produce a frequency plot, we need to divide data into counts, each covering a range of data. This is called 'binning'. There is a choice to be made here, in how large or small to make the divisions ('bins') - this can affect how the data looks when graphed.

For example, to produce a frequency plot of science scores in the UK, we can first run a `summary` command on `PV1SCIE` to find the minimum and maximum score:

```{r}
#| eval: true
#| code-fold: false
#| echo: true

# Find the range of science score
summary(PISA_2022$PV1SCIE)
```

If we wanted to plot a frequency chart of heights, we might divide the range of scores (from 0-895.4) into bins of 20 points. To do this we can use the `cut(<field>,<breaks>)` function within the `mutate` command on a 'data.frame'. The `<field>` specifies the range (i.e. from around 0-900) and `<breaks>` the size of bins.

In the example below, we use `cut(PV1SCIE, breaks=seq(0,900,20))` to create a new vector (column in the table) with the science scores divided up into bins. The specification `breaks=seq(0,900,20))` sets how the data are divided up - into bins (i.e. groups of respondents) of scores starting at 0 and rising to 900 in steps of 20.

```{r}
#| eval: true
#| code-fold: false
#| echo: true

# Creates distribution of science scores

binnedsize <- PISA_2022 %>% # Creates a new data frame with binned data
  select(PV1SCIE, CNT)%>% # Select the columns needed
  filter(CNT == "United Kingdom")%>%
  mutate(BinnedPV1SCIE = cut(PV1SCIE, breaks=seq(0, 900, 20)))%>%
  na.omit() # Drop any NAs

# Plot the data as a bar graph
ggplot(binnedsize,
       aes(x = BinnedPV1SCIE)) +
  geom_bar(fill = "dark green") +
  theme(axis.text.x = element_text(angle = 90, vjust = 0.5, hjust = 1)) +
  labs(y = "Number of responses", x = "Scores range")

```

::: question
Create a graph of the binned counts of mathematics scores in Malta. Don't forget to run a summary command first to get a sense of the range in values.
:::

::: callout-tip
To find out the range of a vector, you can use the `range` function in the console - for example, to get a sense of the range of science scores for the whole data frame, we can type: `range(PISA_2022$PV1SCIE, na.rm=TRUE)`. Notice that, because there are `NA`s in the data, we need to tell the function to ignore them, using `na.rm=TRUE`.
:::

```{r}
#| eval: false
#| code-fold: true
#| code-summary: Answer
#| echo: true
# Creates distribution of schools by size
binnedsize <- PISA_2022 %>% # Creates a new data frame with binned data
  select(PV1MATH, CNT)%>% # Select the column I need
  filter(CNT == "Malta")%>%
  mutate(BinnedPV1MATH = cut(PV1MATH, breaks=seq(20,900,20)))%>%
  na.omit() # Drop any NAs
# Plot the data as a bar graph
ggplot(binnedsize,
       aes(x = BinnedPV1MATH)) +
  geom_bar(fill = "orange") +
  theme(axis.text.x = element_text(angle = 90, vjust = 0.5, hjust = 1)) +
  labs(y = "Number of schools", x = "Pupil range")

```

::: callout-tip
To find out a range of a vector, you can use the `range` function in the console - for example, to get a sense of the range of numbers of students on SEN support. I can type: `range(DfE_SEN_data$SEN.support)`
:::

::: question
Plot a binned `geom_bar` graph of the `HOMEPOS` scores in the UK and Belarus on the same axes.
:::

```{r}
#| eval: false
#| code-fold: true
#| code-summary: Answer
#| echo: true
# Creates distribution of schools by size
binnedwealth <- PISA_2022 %>% # Creates a new data frame with binned data
  select(HOMEPOS, CNT)%>% # Select the column I need
  filter(CNT == "Belarus"| CNT == "United Kingdom")%>%
  mutate(BinnedHOMEPOS = cut(HOMEPOS, breaks=seq(-8, 5, 0.25)))%>%
  na.omit() # Drop any NAs
# Plot the data as a bar graph
ggplot(binnedwealth,
       aes(x = BinnedHOMEPOS, fill = CNT)) +
  geom_bar() +
  theme(axis.text.x = element_text(angle = 90, vjust = 0.5, hjust = 1)) +
  labs(y = "Frequency", x = "Wealth")

```

## Task 2 Plotting data on maps

As well as graphs, R can also plot data onto maps. The `geom_map` function will plot a map of a region and you can either plot points (using `geom_point`) or fill regions by drawing a polygon of the shape of that region (using `geom_polygon`).

For example, imagine we created a data frame of the mean science scores of countries in the PISA data:

```{r}
#| eval: true
#| echo: true

# Pipe the total data frame and calculate the mean science scores
# To match with another data frame we will use, we will rename CNT to region
WorldSci <- PISA_2022 %>%
  select(CNT, PV1SCIE) %>%
  group_by(CNT) %>%
  summarise(meanSci = mean(PV1SCIE)) %>%
  rename(region = CNT)
```

In order to plot the data onto a map, colouring the countries by science scores, we need data which gives the coordinates of the edges of the countries. This data is available from the `map_data` function in `ggplot`. First we load the latitude and longitude data into a new data frame `world_data`. In the next steps we will then use `left_join` to combine it with the science data, but we need to tidy the data to match the PISA data, before we can join the two data sets.

```{r}
#| eval: true
#| echo: true
# Create a data frame of country latitude and longitude data
world_data <- map_data(map = "world")
```

One quirk of the two data frames (`world_data` and the PISA data) is that some countries have different names. For example, PISA uses 'United Kingdom', but the `rworldmap` package uses 'UK'. We can change the names of countries in the PISA `data.frame` to match those in `rworldmap`, using the `case_when` function with `mutate`. For example, if we want to change all `United Kingdom` entries to `UK` we use: `mutate(region = case_when(region == "United Kingdom" ~ "UK"`. The final part of the `case_when` sets the default - `.default = region` - that means if none of the matches are found, then `region` is set to `region` - i.e. it is left unchanged.

```{r}
#| eval: true
#| echo: true


# The names of two countries in the PISA data frame and world_data data frame don't match (UK/United Kingdom and US/United States). Change the level names in the PISA data to match the world_data

 WorldSci <- WorldSci %>%
   mutate(region = case_when(
                       region == "United Kingdom" ~ "UK",
                       region == "United States"  ~ "USA",
                       .default = region))
```

Now that the country names in the two data frames match, we can use `left_join` to combine them.

```{r}
#| eval: true
#| echo: true

# Add the country latitude and longitude data to the PISA scores

WorldSci <- left_join(WorldSci, world_data, by = "region")
```

To add labels, you can create a `data.frame`, with only the country names, and their longitudes and latitudes (the mean of the country longitude and latitude), to use as the labels. You can then add `geom_text_repel`to add the labels. The `repel` means the labels won't overlap. Given the number of countries, using `geom_text_repel` creates a warning that some labels won't fit.

```{r}

# Create a data frame of labels
# Group by region (which is the country names) and calculate a mean latitude and longitude for the country to position the label

Labels<-WorldSci%>%
  group_by(region)%>%
  summarise(meanSci = mean(meanSci), lat = mean(lat), long = mean(long))%>%
  na.omit()

```

To plot the data, we use `ggplot`, with the data frame to `WorldSci`, and the x and y variables set to `long` and `lat`, the longitudes and latitudes. We specify that we want to keep the grouping of the data frame (i.e. by country).

First we use `geom_map` to plot a blank map using the data - this will be the base for the highlighted countries. In the `aes` we give the longitudes and latitudes, and, as we want a blank map, set the `fill` to white and the line `colour` to black.

Finally, we use `geom_polygon` to draw coloured shapes, with the `fill` changing by the value of `meanSci`. To make the map look nice, I have used a pre-defined colour scale.

```{r}
#| eval: true
#| echo: true
# Use geom_map to plot the basic world map (fill is white, line colour is black)
# Use geom_polygon to plot the PISA data
# Add a colour scale
ggplot(data = WorldSci, aes(x = long, y = lat, group=group)) +
  geom_map(data = world_data, 
           map = world_data, 
           aes(map_id = region), 
           fill = "white", 
           colour = "black")+
  geom_polygon(aes(fill = meanSci)) +
  scale_fill_viridis_c(option = "viridis")
```

In ggplot, by default, the aesthetics are passed from one layer to the next. In our first `aes` call we set a grouping (`group=group`). However, when we come to the labeling function `geom_text_repel`, the data frame `Labels` is not grouped. We therefore need to use new aesthetics and want to ignore the original aesthetics (`aes(x=long, y=lat, group=group,label=region))`) we previously set.We do this with the command `inherit.aes = FALSE` (i.e. do not inherit the previous aesthetics). 

To neaten up the labels we use the `ggrepel` package to get the `geom_text_repel` function. You'll need to make sure that you have installed the `ggrepel` package, for a reminder on how to install packages see @sec-installation. This ensures that text labels don't overlap. Then in `geom_text_repel(data=Labels, inherit.aes = FALSE, aes(x=long, y=lat,label=region))` we specify the label `data.frame`, and in `aes` pass the positions of the labels (`x=long, y=lat`) and specify that the labels are in the `region` vector.

```{r}
#| eval: true
#| echo: true
#| warning: false

library(ggrepel)

ggplot(data = WorldSci, 
       aes(x = long, y = lat, group = group,label = region)) +
  geom_map(data = world_data,
           map = world_data, 
           aes(map_id = region), 
           fill = "white",
           colour = "black")+
  geom_polygon(aes(fill = meanSci)) +
  scale_fill_viridis_c(option = "viridis") +
  geom_text_repel(data = Labels,
                 inherit.aes = FALSE,
                 aes(x = long, y = lat, label = region))
  

```

::: question
Plot mean PISA mathematics scores on a world map, labelling the countries.
:::

```{r}
#| eval: false
#| code-fold: true
#| code-summary: "Answer"
#| echo: true
#| warning: false

# Create a data.frame of mathematics scores
WorldMath <- PISA_2022 %>%
  select(CNT, PV1MATH) %>%
  group_by(CNT) %>%
  summarise(meanmath = mean(PV1MATH)) %>%
  rename(region = CNT)
# The names of two countries in the PISA data frame and world_data data frame don't match (UK/United Kingdom and US/United States). Change the level names in the PISA data to match the world_data

WorldMath <- WorldMath %>%
   mutate(region = case_when(
                       region == "United Kingdom" ~ "UK",
                       region == "United States"  ~ "USA",
                       region == "Russian Federation" ~ "Russia",
                       .default = region))

# Create a data frame of country latitude and longitude data
world_data <- map_data(map = "world")

WorldMath <- left_join(WorldMath, world_data, by = "region")

# Use geom_map to plot the basic world map (fill is white, line colour is black)
# Use geom_polygon to plot the PISA data
# Add a colour scale

Labels<-WorldMath%>%
  group_by(region)%>%
  summarise(meanmath = mean(meanmath), lat = mean(lat), long = mean(long))%>%
  na.omit()

# Use geom_map to plot the basic world map (fill is white, line colour is black)
# Use geom_polygon to plot the PISA data
# Add a colour scale
ggplot(data = WorldMath, 
       aes(x = long, y = lat, group = group,label = region)) +
  geom_map(data = world_data,
           map = world_data, 
           aes(map_id = region), 
           fill = "white",
           colour = "black")+
  geom_polygon(aes(fill = meanmath)) +
  scale_fill_viridis_c(option = "viridis") +
  geom_text_repel(data = Labels,
                 inherit.aes = FALSE,
                 aes(x = long, y = lat, label = region))

```

## Task 3 Violin plots

Violin plots are a useful way of showing the distribution of scores across different items. They can make distributions easy to compare, at a glance, and to give a sense of the relative frequency of different scores. In `ggplot` you can use `geom_violin` to produce plots. For example, you can compare the distribution of science scores in Spain and Portugal by gender (note the relatively long tail at the bottom of #####):

```{r}
#| eval: true
#| echo: true
# Create a data frame of Spain and Portugal science scores including gender
PISAsubset <- PISA_2022%>%
  select(CNT, PV1SCIE, ST004D01T)%>%
  filter(CNT == "Spain" | CNT == "Portugal")

# Set up the country on the x axis, and score on the y, setting colour by gender
ggplot(PISAsubset, aes(x=CNT, y=PV1SCIE, fill=ST004D01T))+
  geom_violin()+
  xlab("Country")+
  ylab("Science Score")

```

You can rotate the plot using `+coord_flip()` and add black dots for the mean using: `stat_summary(fun = "mean", geom = "point", color = "black")`. You can also try `geom= "crossbar"`. Because we have two groups (male and female), we need to add `position = position_dodge(width=0.9)` to make the dots appear in the right place

```{r}
#| eval: true
#| echo: true
# Create a data frame of country science scores including gender

PISAsubset<-PISA_2022%>%
  select(CNT, PV1SCIE, ST004D01T)%>%
  filter(CNT == "United Kingdom" | CNT == "Ireland" |
         CNT == "Qatar" | CNT == "Brazil" | CNT == "Kazakhstan" |
           CNT == "Korea" | CNT == "Panama")

# Set up the country on the x axis, and score on the y, setting colour by gender, flip the axes and add points for the mean

ggplot(PISAsubset, aes(x = CNT, y = PV1SCIE, fill = ST004D01T))+
  geom_violin()+
  xlab("Country")+
  ylab("Science Score")+
  coord_flip()+
  stat_summary(fun = "mean", geom = "point", color = "black",
               position = position_dodge(width=0.9))


```

::: question
Plot violin plots of PISA `HOMEPOS` scores, by gender, for the UK, US, Sweden and Finland.
:::

```{r}
#| eval: false
#| code-fold: true
#| code-summary: "Answer"
#| echo: true
#| warning: false

# Create a data frame of WEALTH scores for the 4 countries including gender
PISAsubset<-PISA_2022%>%
  select(CNT, HOMEPOS, ST004D01T)%>%
  filter(CNT == "United Kingdom" | CNT == "Sweden" | 
           CNT == "Finland" | CNT == "United States")

# Set up the country on the x axis, and score on the y, setting colour by gender
ggplot(PISAsubset, aes(x=CNT, y=HOMEPOS, fill=ST004D01T))+
  geom_violin()+
  xlab("Country")+
  ylab("PISA Wealth measure")

```

# Useful resources

-   The '[From Data to Viz'](https://www.data-to-viz.com) blog has a helpful flowchart for choosing an appropriate type of graph for different forms of data. The'[Visual Vocabulary](https://ft-interactive.github.io/visual-vocabulary/)' site, produced by the Financial times is an alternative option for selecting an appropriate chart.
-   You can find a gallery of around 400 types of visualisation you can produce in R [at the R chart gallery](https://r-graph-gallery.com/all-graphs.html).
-   The Royal Statistical Society has produced a guide to making visual representations, [Best Practice for Data Representations](https://royal-statistical-society.github.io/datavisguide/), that shows you how to make readable, accessible data visualisations.
-   The classic text on representing data visually is Edward R. Tufte's *The Visual Display of Quantitative Information* [@tufte2001visual]. *The Visual Display* is a beautiful book with illustrations of best practice in graphing and other forms of data representation. Tufte sets out the following principles - a representation should:
    - show the data
    - induce the viewer to think about the substance rather than about methodology, graphic design, the technology of graphic production, or something else
    - avoid distorting what the data have to say
    - present many numbers in a small space
    - make large data sets coherent
    - encourage the eye to compare different pieces of data
    - reveal the data at several levels of detail, from a broad overview to the fine structure • serve a reasonably clear purpose: description, exploration, tabulation, or decoration
    - be closely integrated with the statistical and verbal descriptions of a data set. [@tufte2001visual, p. 13]

Lukasz Piwek has [created guidance](http://motioninsocial.com/tufte/) on creating Tufte-style representations using R packages.

-   Further useful guidance on graphing can be found at the [Friends Don't Let Friends Make Bad Graphs](https://github.com/cxli233/FriendsDontLetFriends#1-friends-dont-let-friends-make-bar-plots-for-means-separation) pages. For example, note the advice that: 'Friends Don't Let Friends Make Pie Chart' - the eye is not very good at interpreting the angles on sections of pie charts - differences between sectors of similar sizes can be hard to judge, so [their use has been criticized](https://www.data-to-viz.com/caveat/pie.html). A stacked bar chart is a better option.

-   Researchers have recently suggested journals avoid bar graphs - bar charts mask the distribution of points and can misrepresent the data [@weissgerber2015beyond]. Box plots or, if samples are large enough, violin plots are recommended.

-   See also [I’ve Stopped Using Box Plots. Should You?](https://nightingaledvs.com/ive-stopped-using-box-plots-should-you/)
