---
title: "06 T-tests"
editor: 
  markdown: 
    wrap: 72
echo: false
warning: false
#bibliography: ../references/mastemr.bib
---

```{r setup}
    options(max.print=15)

    library(here)
    source(here::here("MASTEMR_page_setup.R"))
```

# Pre-reading and pre-session tasks

## Optional Pre-reading

Before the session, you might find it useful to read chapter 7, introduction to t-tests, in
@geher2014straightforward [available
here](https://ebookcentral.proquest.com/lib/kcl/reader.action?docID=1692209&ppg=136).

# Hypothesis testing

Hypothesis testing is a form of statistical inference used to draw conclusions about population distributions or parameters (such as the mean or variance). Data from a simple random sample are used to test the plausibility of a hypothesis and the likelihood of it being true (or not).

When going about doing a hypothesis test we commonly choose what are called the null and alternative hypotheses. The null hypothesis usually refers to the hypothesis that there is no significant statistical difference in a set of observations, such as differences between expected values and observed values, differences between means, or differences between data and a distribution. The alternative hypothesis usually refers to the opposite of this, where there is a significant statistical difference in a set of observations, however, sometimes this can be of one direction, such as one mean being greater than another mean, instead of there being no difference.

We assume the null hypothesis is true when conducting a hypothesis test. The test itself then measures the plausibility that the null hypothesis is true and returns what is called a p-value (the probability the null hypthesis is true). If this value is relatively large then it is likely the null hypothesis is true. However, if this value is relatively small, then it is unlikely the null hypothesis is true and therefore likely it is false and that the alternative hypothesis is true instead.

Typically, we set a value such as 0.05 or 0.01 as the threshold to determine whether the null-hypothesis is true or not. So, if the null hypothesis is greater than this value then we accept that it is true, and if it is less than this value we reject that it is true and accept that the alternative hypothesis is true instead.

## Choosing a Hypothesis Test

When conducting a hypothesis test we need to choose the most appropriate test depending on the type of data we are working with, what we are trying to test and whether certain conditions are met or not. Here is a basic summary of what you may need to consider.

Type(s) of data:

-   Categorical (ordinal, nominal, binary);
-   Quantitative (continuous, discrete).

What you are trying to test:

-   Relationships between variables;
-   Comparison of means.

Distribution of data:

-   Normally distributed (parametric tests);
-   Not normally distributed (nonparametric tests).

When we consider each of the types of tests the conditions for the test will be stated, so it will be clear which test can be used when.

The flowchart below sets out some common statistical tests, and when they are used:


```{r }
#| eval: true
#| echo: true
#| code-fold: true
#| code-summary: "Show the code"
grViz("
      digraph Random{
      graph [layout = dot,
      overlap =T,
      bgcolor='white',
      splines=line]#controls l type setup
      node [shape = box,style='filled',
      fillcolor='skyblue',
      fontSize=20,fontcolor='darkgrey',
      fontname='Arial']
     a [label = 'What do you want to do?']
     b [label = 'Test a hypothesis - on what kind of data?']
     c [label = 'Examine a relationship']
     d [label = 'Continuous'];
     e [label = 'Normally Distribured/Parametric'];
     f [label = '1 Group'];
     g [label = 'One sample t-test', fillcolor='lightyellow'];
     h [label = 'Do a test of normality'];
     i [label = 'qqplot', fillcolor='lightyellow'];
     j [label = 'Tests on skewed distributions', fillcolor='lightyellow'];
     k [label = 'Discrete / Categorical'];
     l [label = '2 Groups'];
     m [label = 'Unpaired groups'];
     n [label = 'Expected counts more than 5\nin more than 75 per cent of cells'];
     o [label = 'Chi-squared', fillcolor='lightyellow'];
     p [label = 'Expected counts more than 5\n in less than 75 per cent of cells'];
     q [label = 'Fisher exact test', fillcolor='lightyellow'];
     r [label = 'Continuous variables'];
     s [label = 'Linear Regression', fillcolor='lightyellow'];
     t [label = '2 groups']
     u [label = 'Paired groups']
     v [label = 'Unpaired groups']
     w [label = 'Paired t-test', fillcolor='lightyellow']
     x [label = 'Unpaired t-test', fillcolor='lightyellow']
     y [label = '3 groups or more']
     z [label = 'anova', fillcolor='lightyellow']
     aa [label = 'Not normally distributed'] 
     a -> b
     a -> c
     b -> d
     d -> h
     h -> i
     i -> aa
     aa -> j
     i -> e
     e -> f
     f -> g
     b -> k
     k -> l
     l -> m
     m -> n
     n -> o
     m -> p
     p -> q
     c -> r
     r -> s
     e -> t
     t -> u
     t -> v
     u -> w
     v -> x
     e -> y
     y -> z

      }")
```

## Performing a hypothesis test - Fisher

Fisher, one of the statisticians who moved the field of hypothesis testing forward and formalised some of the procedures used for hypothesis testing, suggested using the following steps when conducting hypothesis testing.

1.  Select an appropriate test.

    Here, we need to consider the type(s) of data, what we are trying to test and whether the data is normally distributed or not, along with other conditions needed for the different tests.

2.  Set up the null and alternative hypotheses.

    This heavily depends on which test is being used, so more guidance will be given under each of the tests.

3.  Calculate the theoretical probability of the null hypothesis being true.

    This is where the test itself is used to calculate the probability of the null hypothesis being true (i.e. returning the p-value from the test).

4.  Assess the statistical significance of the result.

    This is where the p-value from step 3 is compared with a predetermined threshold, such as 0.05 or 0.01 to determine if the null hypothesis is true or false.

5.  Interpret the statistical significance of the results.

    Here, we take the result from step 4, so deciding whether to accept the null hypothesis or reject the null hypothesis and accept the alternative hypothesis and then what this means in the context of the problem being posed.

# T-tests

T-tests are statistical tests used to determine if the mean of a sample
is different to a presumed population mean, or if the means of two
different samples are different to one another.

Here are some examples in educational research of when t-tests might be
applied:

1.  A school wants to measure whether there is a statistical difference
    between Year 11's Progress 8 scores and the national average (which
    is approximately 0 on average in England). A t-test is used to
    determine if there is a statistical difference between the mean of
    the Year 11's progress 8 scores in the school and the national
    average.
2.  A researcher implements an intervention with a random sample of
    pupils from a school. A second random sample is also taken, which is
    called the control group. After a period of time of teaching, one
    with the intervention and without, the pupils are tested. A t-test
    is performed to see if there is a statistical difference between the
    mean of the control group and the mean of the intervention group.
3.  A teacher wants to see how well a class is able to recall knowledge
    learned 6 months after they are given a test. The pupils are given
    similar test questions before and after the period. A t-test is used
    to see if there is a statistical difference between test scores in
    the first test and in the test 6 months later.

## Conditions of t-tests

There are six conditions required for a t-test to be used:

a\) the population data is continuous;

b\) the sample size is small;

c\) the sample was randomly sampled;

d\) the variance of the data in each group is similar (homogeneity of
variance);

e\) the population data follows a normally distribution;

f\) the observations are independent of one another.

::: callout-note
Due to the robustness of the t-test some of the conditions can be
adjusted. For example,

-   Discrete interval data can be used (so long as there is sufficient
    granularity);
-   The sample data may be approximately normally distributed;
-   Variances can be unequal (this is covered by Welch's t-test);
-   Sample sizes can be large (this does not affect the validity of the
    test, but with a known population variance a z-test may be more
    appropriate).
:::

::: callout-note
A rule of thumb researchers use is that if the ratio of variances of the
larger sample to the smaller is less than 4, then a t-test can be used.
:::

## Hypothesis Testing and t-tests

When using the t-test, we are testing the null hypothesis that there is
no difference between the mean of a sample and, depending on the type of
t-test, either the hypothesised value of the population mean or the mean
of another sample. In other words, **if it is equal**.

```{r}
#| eval: true
#| echo: true
#| code-fold: true
#| code-summary: show the code

# Create random data sets of 10,000 boys' and girls' scores
girlscores <- as.data.frame(rnorm(n = 10000, mean = 100, sd = 10))

girlscores <- girlscores %>%
  mutate(gender="female") %>%
  rename(Scores="rnorm(n = 10000, mean = 100, sd = 10)")

boysscores <- as.data.frame(rnorm(n = 10000, mean = 105, sd = 10))

boysscores <- boysscores %>%
  mutate(gender="male") %>%
  rename(Scores="rnorm(n = 10000, mean = 105, sd = 10)")

# Plot the data
totalscores <- rbind(boysscores, girlscores)

ggplot(totalscores) +
  geom_histogram(data = boysscores,
                 aes(x = Scores, fill = gender, alpha=0.5), binwidth = 1) +
  geom_histogram(data=girlscores,
                 aes(x = Scores, fill = gender, alpha=0.5), binwidth = 1) +
  geom_vline(xintercept = mean(boysscores$Scores), colour="turquoise4", 
             linetype = 3) + # Add a vertical line for the mean scores
  geom_vline(xintercept = mean(girlscores$Scores), colour="red2", linetype=3) +
  geom_segment(aes(x = mean(boysscores$Scores), y=440, # Add arrow between means
                   xend = mean(girlscores$Scores),
                   yend = 440), 
               arrow = arrow(length = unit(0.01, "npc"), ends = "both")) +
  geom_text(x = 125, y = 440, label = "Difference in means") + # Add text label
  xlab("score") +
  ylab("Numer of students") +
  guides(alpha="none") # Remove alpha from the legend


```

The alternative hypothesis then depends on whether we are testing if the
mean of a sample is less than, greater than, or not equal to the
hypothesised value of the population mean or mean of another sample. In
other words, **if it less than or greater than, then it's a one-tailed
test** (since it is only considering one direction of the mean), and
**if it's not equal to, then it's a two-tailed test** (since it is
considering both sides of the mean).

```{r}
#| eval: true
#| code-fold: true
#| echo: true
library(gridExtra) # for grid.arrange

plot1 <- data.frame(y = rnorm(n = 1000, mean = 100, sd = 10), group = "A")
plot2 <- data.frame(y = rnorm(n = 1000, mean = 130, sd = 10), group = "B")
plot3 <- data.frame(y = rnorm(n = 1000, mean = 70,  sd = 10), group = "C")
plot4 <- data.frame(y = rnorm(n = 1000, mean = 100, sd = 10), group = "D")
plot5 <- plot3

plot5$group<-"E"
plot5$y <- plot5$y + 60
onetail <- rbind(plot1, plot2)
twotail <- rbind(plot3, plot4, plot5)

p1<-ggplot(onetail, aes(x = y, fill = group)) +
  geom_density(alpha = 0.5) +
  geom_segment(aes(x = mean(plot1$y), y = 0.045, # Add an arrow between means
                   xend = mean(plot2$y),
                   yend = 0.045), 
               arrow = arrow(length = unit(0.01, "npc"), ends = "both"))+
  geom_vline(xintercept = mean(plot1$y), linetype="dashed")+
  geom_vline(xintercept = mean(plot2$y), linetype="dashed")+
  ggtitle("One tailed tests look for differences in one direction")

p2 <- ggplot(twotail, aes(x = y, fill = group)) +
  geom_density(alpha = 0.5) +
  geom_segment(aes(x = mean(plot3$y), y = 0.043, # Add an arrow between means
                   xend = mean(plot4$y),
                   yend = 0.043), 
               arrow = arrow(length = unit(0.01, "npc"), ends = "both")) +
  geom_segment(aes(x = mean(plot4$y), y=0.045, # Add an arrow between means
                   xend = mean(plot5$y),
                   yend = 0.045), 
               arrow = arrow(length = unit(0.01, "npc"), ends = "both"))+
  geom_vline(xintercept = mean(plot3$y), linetype="dashed")+
  geom_vline(xintercept = mean(plot4$y), linetype="dashed")+
  geom_vline(xintercept = mean(plot5$y), linetype="dashed")+
  ggtitle("Two tailed tests look for differences in two directions")

grid.arrange(p1, p2)
```

To determine the outcome of the t-test we need to decide a sufficiently
low enough probability (p-value) that, if it's below this, then it is
unlikely that the null hypothesis is true and so we reject the null
hypothesis. Typically, this is set to 0.05, or 5%, but in other cases it
might be as high as 0.1 or as low as 0.01 or 0.001 depending on the
circumstances. Therefore, if the p-value is greater than 0.05 we accept
the null hypothesis and if it less than 0.05 we reject the null
hypothesis and accept the alternative hypothesis.

When considering the number of tails, we then need to consider the
p-value here also. If we set the overall probability to be 0.05 for the
threshold that we reject the null hypothesis, then for a **one-tailed**
test we compare our result with **0.05**. However, with a two-tailed
test, since we are looking at a **two-tailed** test, then the
probability of each of the tails is 0.025 and therefore we compare our
value with **0.025** instead. Fortunately, the `t.test()` function in R 
automatically scales the output so we don't need to make the adjustment 
ourselves - we can just compare to **0.05**. R automatically calculates the correct critical t-value or p-value corresponding to a single tail having an area of α.

Conventionally, we use 0.05 (1 in 20) as the cut off for statistical
significance (a convention that has been much critiqued e.g.
[@cohen1994earth; @baker2016statisticians]).

## Types of t-test

There are a two main types of t-tests:

-   One sample t-test - checks if the mean of a sample is different to a
    hypothesised value for the population mean.
-   Two sample t-test - checks if the mean of one sample is different to
    the mean of another sample. There are two main types:
-   Paired tests - compares the means of measurements from the same
    individual or object (e.g. in a pre- and post-test - the comparison
    of means and before and after scores compares the same student's
    scores in the first test with the second).
-   Unpaired t-tests - compares the means of two unrelated groups (for
    example, are the mean science scores of boys and girls in a school
    different).

For more information on t-tests, see [chapter
13](https://learningstatisticswithr.com/book/ttest.html), in Navaro's
Learning Statistics with R.

## One Sample t-tests

One sample t-tests are used to compare the mean of a sample of a data
set with a value. This value is usually a known population mean or a
theoretical population mean depending on the context of the test.

For a **two-tailed test**, the hypotheses are as follows

-   Null hypothesis: sample mean = population mean;
-   Alternative hypothesis: sample mean ≠ population mean.

For a **one-tailed test**, the hypotheses are as follows (depending on
the direction of the test)

Either,

-   Null hypothesis: sample mean = population mean;
-   Alternative hypothesis: sample mean \> population mean.

Or,

-   Null hypothesis: sample mean = population mean;
-   Alternative hypothesis: sample mean \< population mean.

**Example: Are Females' Reading Scores in the UK the Same or Different
on Average to the UK Average?**

Let's look at an example of where we might want to consider a sample, in
this case females' reading scores in the UK, with a presumed population
mean, the average reading score in the UK. This is a presumed mean,
since we don't know the true population mean here, but we do know the
average reading score from the PISA data, which is 494 (taken from PISA
[2022](https://www.oecd.org/publication/pisa-2022-results/country-notes/united-kingdom-9c15db47/)).

Following the steps for performing a hypothesis test, we will start by
selecting the appropriate test. Since we are comparing the mean of one
sample with a population mean, then we need to use a one sample t-test.

First, you will need to load the PISA data:

```{r}
#| eval: false
#| code-fold: false
#| echo: true

# Load PISA Data
PISA_2022 <- read_parquet(r"[add the link to your data file]")

```

To ensure we can use this test we need to make sure the conditions are
met. Since PISA tests are constructed in a way which means they are
normally distributed, then it is likely our sample is also normally
distributed. However, it is good practice to check the distribution
using a histogram (or qqplot) at the very least, and where appropriate,
a normality test. Due to the large sample size normality testing isn't
usually required, so we will just consider a histogram (or qqplot) here.

```{r}
#| eval: true
#| code-fold: false
#| echo: true

# Define PISA UK Girls Dataset
PISA_2022_Read_Girls <- PISA_2022 %>%
  select(CNT, PV1READ, ST004D01T) %>%                      # <2>
  filter(CNT == "United Kingdom" & ST004D01T == "Female")  # <3>

# Plot histogram
ggplot(data = PISA_2022_Read_Girls,                          # <4>
       aes(x = PV1READ)) +                                   # <5>
  geom_histogram(binwidth = 5, fill = "darkseagreen4")       # <6>
```
2.  `select` the variables of interest
3.  `filter` for the UK and for girls
4.  pass the data to `ggplot`
5.  put reading scores (`PV1READ`) on the x-axis
6.  use `geom_histogram` to create a histogram. We can define that the reading data is divided into bins of width 5 points (`binwidth = 5`) and set the colour of the bars (`fill = "darkseagreen4"`)

We can see from the histogram that the sample is approximately normally
distributed (as it follows a bell shaped curve).

Whilst examining the shape of the histogram can give some indication of normality it can be hard to judge deviations from a perfect normal curve by eye. An alternative is the quantile-quantile plot (or qqplot, for short). In this plot, the distribution is divided into equal sized intervals (the quantiles) and compared against the quantiles of a normal distribution. 

For example, in a normal distribution, the 95th percentile should come at a value equal to 1.645 standard deviations from the mean, the 50th percentile (the median) corresponds to 0 in a standards normal distribution, and so on. If the value of quantiles aligns with that expected for a normal distribution the plot will produce a straight line where y = x. Therefore a qqplot of a normal distribution should give a diagonal line (keep your eyes out for deviations from the straight line at the extremes of the line, suggesting some deviation from normal in the tails of the distribution). The judgement here is still subjective (it is hard to state how much deviation from a straight line is acceptable) but it is easier to compare to a straight line, that to a bell curve as in the histogram.

To perform a qqplot in R you can use the `qqpnorm()` function. Adding `qqline()` afterwards adds a coloured reference y = x line that makes judgement easier.

```{r}
#| eval: true
#| code-fold: false
#| echo: true

# Plot a qqplot
qqnorm(PISA_2022_Read_Girls$PV1READ)                 # <1>
qqline(PISA_2022_Read_Girls$PV1READ, col = "red")    # <2>

```


1. plot a quantile quantile plot of `PISA_2022_Read_Girls$PV1READ`
2. Add a reference line in red

The data here look a good fit with the normal distribution.

Next, we state the null and alternative hypotheses. Since we are
determining whether the average reading score is the same or different,
we are using a two-tailed test. This gives us the following hypotheses:

-   Null hypothesis: mean reading score for females in UK = mean reading
    score in UK
-   Alternative hypothesis: mean reading score for females in UK ≠ mean
    reading score in UK

Having set up the hypotheses, we now need to calculate the probability
of the null hypothesis being true. To do this, we use R's `t.test`
function and compare the scores for females in reading in the UK with
the OECD average score for reading, which is defined as 476 ([OECD
website](https://www.oecd-ilibrary.org/sites/9149c2f5-en/index.html?itemId=/content/component/9149c2f5-en#:~:text=linklink%20copied!-,Average%20performance%20in%20mathematics%2C%20reading%20and%20science,in%20science%20is%20485%20points.)).

```{r}
#| eval: true
#| code-fold: false
#| echo: true
OECD_ave_read_UK <-(476)
t.test(PISA_2022_Read_Girls$PV1READ, mu = OECD_ave_read_UK)

```

We can see from the result that we get a p-value of \< 2.2e-16, which is
very small. As this is lower than 0.05 we
reject the null hypothesis.

Therefore, we accept the alternative hypothesis that the average reading score
for females in the UK is different to the overall average reading score
in the UK. Looking at the average stated in the t-test, the average
score is higher for girls, at 500.23.

## Unpaired Two Sample t-tests

Unpaired two sample t-tests are used to compare the means of two,
unconnected samples. They are unconnected in the sense that the samples
taken are independent and are measurements of two unrelated groups.

For a **two-tailed test**, the hypotheses are as follows

-   Null hypothesis: sample mean of first group = sample mean of second
    group;
-   Alternative hypothesis: sample mean of first group ≠ sample mean of
    second group.

For a **one-tailed test**, the hypotheses are as follows (depending on
the direction of the test)

Either,

-   Null hypothesis: sample mean of first group = sample mean of second
    group;
-   Alternative hypothesis: sample mean of first group \> sample mean of
    second group.

Or,

-   Null hypothesis: sample mean of first group = sample mean of second
    group;
-   Alternative hypothesis: sample mean of first group \< sample mean of
    second group.

**Example: Are Males' and Females' Maths Scores in the USA the Same or
Different to One Another?**

Let's say we want to discern if the average maths score differ for males
and females in the USA. We want to test to see if there is a statistical
difference between the means.

Following the steps for performing a hypothesis test, we will start by
selecting the appropriate test. Since we are comparing the means of two,
unrelated samples (since they are the measurements of two different
groups that are not connected) then we use an unpaired t-test.

Again, we need to check if the groups are approximately normally
distributed so check the histogram of each group.

```{r}
#| eval: true
#| code-fold: false
#| echo: true
# Create a US maths subsets
USMATH<-PISA_2022 %>%
  select(CNT, PV1MATH,ST004D01T) %>%                      # <2>
  filter(CNT == "United States")                          # <3>

USMATHGIRLS <- USMATH %>%
  filter(ST004D01T == "Female")                           # <4>

USMATHBOYS <- USMATH %>%
  filter(ST004D01T == "Male")                             # <5>
# Plot histogram
# For girls
ggplot(data = USMATHGIRLS,                                # <6>
       aes(x = PV1MATH)) +                                # <7>
  geom_histogram(binwidth = 5, fill = "darkseagreen4")    # <8>

# For boys
ggplot(data=USMATHBOYS,
       aes(x=PV1MATH)) +
  geom_histogram(binwidth = 5, fill = "red")
```
2.  `select` the variables of interest, country, maths scores, and gender
3.  `filter` for the US
4.  create a new data frame `USMATHGIRLS` of the data for girls
5.  create a new data frame `USMATHBOYSS` of the data for boys
6.  pass the data to `ggplot`
7.  put maths scores (`PV1MATH`) on the x-axis
8.  luse `geom_histogram` to create a histogram. We can define that the reading data is divided into bins of width 5 points (`binwidth = 5`) and set the colour of the bars (`fill = "darkseagreen4"`)

Again, we can see both samples of males and females are approximately
normally distributed.

Next, we state the null and alternative hypotheses. Since we are
determining whether the average maths scores are the same or different,
we are using a two-tailed test. This gives us the following hypotheses:

-   Null hypothesis: mean maths score for males in US = mean maths score
    for females in US
-   Alternative hypothesis: mean maths score for males in US ≠ mean
    maths score for females in US

Next, we calculate the probability of the null hypothesis being true. As
before, we use R's `t.test` function and compare the mean scores of
males and females in the US.

```{r}
#| eval: true
#| code-fold: false
#| echo: true

# Plot a two-sided, unpaired t-test
t.test(PV1MATH ~ ST004D01T, 
       data = USMATH, 
       alternative = "two.sided")

```

We can see from the t-test result that the p-value returned is 8.086e-05.
As this is lower than 0.05 we reject the null
hypothesis.

As such, we accept the alternative hypothesis that the average scores
for males and females in the US are statistically different, with the
average score for males being higher than average score for females.

## Paired Two Sample t-tests

Paired two sample t-tests compare the mean of a sample to the mean of a
related sample. Usually, this is where there is a repeated measure taken
of the same individuals or objects, such as a pre and post test. It can
also be used for comparing a pair of measurements taken of the same
individuals or objects.

For a **two-tailed test**, the hypotheses are as follows

-   Null hypothesis: sample mean of first group = sample mean of second
    group;
-   Alternative hypothesis: sample mean of first group ≠ sample mean of
    second group.

For a **one-tailed test**, the hypotheses are as follows (depending on
the direction of the test)

Either,

-   Null hypothesis: sample mean of first group = sample mean of second
    group;
-   Alternative hypothesis: sample mean of first group \> sample mean of
    second group.

Or,

-   Null hypothesis: sample mean of first group = sample mean of second
    group;
-   Alternative hypothesis: sample mean of first group \< sample mean of
    second group.

**Example: Are there differences between science and maths scores on
average in the UK?**

Let's say we want to compare the science and maths scores of students in
the UK to see if there is a difference in their average score or not.

Following the steps for performing a hypothesis test, we will start by
selecting the appropriate test. Since we are comparing the means of two,
related samples (since they are two measurements of the same students,
therefore related) then we use a paired t-test.

Again, we need to check if the groups are approximately normally
distributed so check the histogram of each group.

```{r}
#| eval: true
#| code-fold: false
#| echo: true
# Create a US maths subsets
UKMATH<-PISA_2022 %>%
  select(CNT, PV1MATH,ST004D01T) %>%
  filter(CNT == "United Kingdom")

UKSCIENCE<-PISA_2022 %>%
  select(CNT, PV1SCIE, ST004D01T) %>%
  filter(CNT == "United Kingdom")

# Plot histogram
ggplot(data = UKMATH,
       aes(x = PV1MATH)) +
  geom_histogram(binwidth = 5, fill = "darkseagreen4")
ggplot(data = UKSCIENCE,
       aes(x = PV1SCIE)) +
  geom_histogram(binwidth = 5, fill = "red")
```

Again, we can see the two samples are approximately normally distributed so can
use the t-test.

Next, we state the null and alternative hypotheses. Since we are
determining whether the average maths score is the same or different to
the average science score, we are using a two-tailed test. This gives us
the following hypotheses:

-   Null hypothesis: mean maths score in UK = mean science score in UK
-   Alternative hypothesis: mean maths score in UK ≠ mean science score
    in UK

Having set up the hypotheses, we now need to calculate the probability
of the null hypothesis being true. Again, we use R's `t.test` function
and compare the average maths scores with the average science scores.

```{r}
#| eval: true
#| code-fold: false
#| echo: true

# Perform a two-sided, paired t-test
t.test(UKMATH$PV1MATH, UKSCIENCE$PV1SCIE, 
       paired = TRUE, alternative = "two.sided")
```

We return a p-value of < 2.2e-16, which is less than 0.05, meaning we reject the null hypothesis that there is no
difference between the average score in maths and science.

Therefore, we accept the alternative hypothesis that there is a
difference in scores in maths and science in the UK. This difference is
approximately 9.7 points.

::: callout-tip

In R, there are often multiple ways of doing the same tests. Note with
`t.test` there are two forms you can use, depending on the format of the
data you want to test.

First, in the example above, we had two data frames, for math scores,
and science scores, so we can pass two data frames to `t.test`.

```{r}
#| eval: true
#| code-fold: false
#| echo: true

# Carry out a two-sided, paired t-test
t.test(UKMATH$PV1MATH, UKSCIENCE$PV1SCIE, 
       paired = TRUE, alternative = "two.sided")
```

Alternatively, our data might be in one data frame, with a variable
indicating the two groups. For example, if we wanted to perform a test
of UK science scores, we might create a single data frame in which the
two groups are indicated by the column `ST004D01T`, gender. In that
case, we can use the `~` operator to do a `t.test` on the
`PISAUK$PV1SCIE` (i.e. the science scores) by (`~`) gender.

```{r}
#| eval: true
#| code-fold: false
#| echo: true

UKSci <- PISA_2022 %>%
  select(CNT, PV1SCIE, ST004D01T) %>%
  filter(CNT=="United Kingdom")

# Plot a two-sided, unpaired t-test
t.test(PV1SCIE ~ ST004D01T, 
       data = UKSci,
       alternative = "two.sided")
```

:::

## Finding an effect size for a t-test

Recent guidance suggests that when reporting the results of a t-test,
researchers should focus not only on the p-value, but also on the effect size. Especially with large sample sizes, it is possible to find a statistically significant difference between means, but the actual difference may be very small and not of practical significance. Hence it is recommended to report an effect size, which gives a measure of the magnitude of the difference between means.

One common measure of effect size for t-tests is Cohen's d. Cohen's d is calculated as the difference between the two means divided by the pooled standard deviation. 

To calculate Cohen's d in R, we can use the `cohen.d` function from the `effsize` package.

```{r}
#| eval: true
#| code-fold: false
#| echo: true
#| warning: false
#| message: false

library(effsize)
# Calculate Cohen's d for the difference in science scores by
# gender

cohen.d(data = UKSci, PV1SCIE ~ ST004D01T)

```
Cohen originally proposed a set of benchmarks for interpreting the magnitude of d: less than 0.2 is negligible, 0.2-0.5 is considered a small effect size, 0.5-0.8 a medium effect size, and over 0.8 a large effect size. However, these benchmarks should be interpreted with caution and in the context of the specific research area. For example, in education, where many factors influence outcomes, even a small effect size may be meaningful. @kraft2020interpreting suggest alternative, lower values for the effect size benchmarks in education research, with 0.05 as small, 0.05-0.2 as medium, and more than 0.2 as large. The cut offs are summarised in the table below.

| Effect Size | Cohen's Benchmarks | Kraft et al.'s Benchmarks for Education Research |
|-------------|--------------------|--------------------------------------------|
| Negligible  | < 0.2              | NA.                                        |
| Small       | 0.2 - 0.5          | < 0.05                                     |
| Medium      | 0.5 - 0.8          | 0.05 - <0.2                                |
| Large       | > 0.8              | >= 0.2                                     |


# Anova (multiple t-tests)

We have performed t-tests of the difference in mathematics scores
between boys and girls in the UK. We may now want to perform the same
t-test for all the countries in the sample. This can create a problem.
Remember that a p-value report the probability that an outcome is due to
chance. The more times we repeat a t-test, the more likely we are to see
an effect, even when none-exists.

To avoid potential artefacts from doing multiple t-tests, when we want
to compute the means of more than two groups, we use an anova (analysis
of variance) test.

To determine if there is a statistically significant difference between
groups we run an anova calculation, using the `aov` function in R. The
dependent variable is the one we are interested in explaining, the
independent variables are the factors we think might explain the
variance.

`aov(data, dependent_var ~ independent_var + independent_var + ...)`

For example, we used t-tests to look at differences in scores between a
pair of countries (e.g the UK and the US). We can now compare across all
the countires.

We pass the function the data we wish to focus on (`PV1MATH`) and then
indicate we wish to look for variation in the `PV1MATH`, by `CNT`. We
then summarise the result to get a table.

```{r}
#| eval: TRUE
#| code-fold: false
#| echo: true

ALLPISAMATH<-PISA_2022 %>%
  select(CNT, PV1MATH)                              # <1>

resaov <- aov(data = ALLPISAMATH, PV1MATH ~ CNT)    # <2>
sumresaov <- summary(resaov)                        # <3>
sumresaov                                           # <4>

```
1.  `select` the variables of interest, country, and maths scores
2.  use `aov` to perform the anova calculation. We set the data we are passing (`data = ALLPISAMATH`) and set that we want to determine the variance in maths scores by country (`PV1MATH ~ CNT`) - we put the output into a new variable `resaov`
3.  we produce a summary of the output `summary(resaov)`    
4.  print the result


looks to determine if there is a difference in the variance of means across
countries. After determining any variation of mean from those two variables, any
remaining variation is associated with residuals - you can think of this
as the unexplained variation that isn't associated with the vectors we
specified (i.e. the country).

First, look at the `Pr(>F)` column. This is a test of significance and
reports if the groups in that row show statistically significant
variation. In our case, country returns `Pr(>F) <2e-16 ***` - which is
below `0.05`. This suggests there are significant differences across
countries.

This is not terribly interesting so far(!) - but we can do two post-hoc
tests (i.e. tests after the first test, to find out the effect size and
between which countries differences exist)

## Eta-squared

Now knowing that there are significant differences between the two class
groups, we can next determine how much variation in the test score is
explained by class. To do this we calculate a variable called
eta-squared.

Eta-squared gives the proportion of variance of means explained by each variable.
The eta squared function is in the package `lsr` so we will install
`lsr`, and then use the result of our anova (`resaov`) to calculate the
eta squared variable using the function `etaSquared`. To report the
value of eta as a percentage we need to multiply the output of eta by
100.

Eta squared tells us the proportion of the total variance in means that is
explained by a variable (you can also think of it as an effect size). An
eta squared value of 1 indicates all the variance of a sample is explain
by some variable and 0 means the variable is not responsible for any of
the observed variance. We can multiply the eta-squared score by 100 to
get a value for the percentage of variance explained.

The percentage of variance of means explained is a useful figure. For example, it
has been reported that schools only account for 14% of the variance in
progress 8 scores, whilst family explains 43% (@wilkinson2018assessing).

```{r}
#| eval: TRUE
#| code-fold: false
#| echo: true

library(lsr)
eta <- as.data.frame(etaSquared(resaov))
eta <- eta*100
eta

```

The important column here is the `eta.sq` column - it tells us that the
country explains 28% of the variance in test scores. (The second column
contains information about partial eta-squared, which we won't go into,
but are used when the results of each measure are *not independent* i.e.
one result influences another).

::: callout-note
Interpretations of effect size vary. In some fields, 2% - 13% is read as small, 13% -26% medium and over 26% large [@shevlin2000applying; @cohen1988statistical]. In educational research, effect sizes are typically small, hence different categorisations are applied.


| Interpretation | Effect Size[@kraft2020interpreting] |Effect Size [@sanders2020effect]|
|----------------|------------------------------------|--------------------------------|
| Small          |              < 5%                  |              3% - 10%          |
| Medium         |             5% to 20%              |              10% - 20%         |
| Large          |              > 20%                 |              > 20%             |

:::

## Tukey's HSD

When an anova test reports that are some statistically significant
differences between groups, it does not imply there are statistically
significant differences between all subgroups. For example, if the anova
reports statistically significant differences by age, statistically
significant differences might exist between 11, 12 and 13-year old
students, but not between 13 and 14-year olds.

We can use an additional anova test, **Tukey's Honest Significant
Difference** test (or Tukey's HSD for short), to find out which pairs of
subgroups have statistically significant differences in means.

Consider the question: Are there statistically significant differences
in the variance of science scores of the UK, US, France and Germany? To
determine if such differences exist, we create a new subset for those
countries' science scores, and then run an anova test by country,
reporting the eta squared value.

```{r}
#| eval: TRUE
#| code-fold: false
#| echo: true
PISAMULTI <- PISA_2022 %>%
  select(CNT, PV1SCIE) %>%                                        # <1>
  filter(CNT == "United Kingdom" | CNT == "United States" |       # <2>
           CNT == "Germany" | CNT == "France")

resaov<-aov(PV1SCIE ~ CNT, data = PISAMULTI)                      # <3>
sumresaov<-summary(resaov)                                        # <4>
sumresaov

eta<-as.data.frame(etaSquared(resaov))                            # <5>
eta<-eta*100                                                      # <6>
eta                                                               # <7>
```
1.  `select` the variables of interest, country, and maths scores
2.  `filter` for countries of interest. Note a shorter way to do the filter is `filter(CNT %in% c("United Kingdom", "United States", "Germany", "France"))`
3.  use `aov` to perform the anova calculation. We set the data we are passing (`data = PISAMULTI`) and set that we want to determine the variance in science scores by country (`PV1SCIE ~ CNT`) - we put the output into a new variable `resaov`
4.  produce a summary of `resaov`
5.  use `etaSquared(resaov)` to find the eta squared value. We convert this to a data frame so we can perform the next steps
6.  to convert the eta squared value into a percentage of variance explained, we multiply by 100   
7.  print the result

The anova results tell us there are significant differences between the
countries, which account for 0.3 % of variance in scores. We can then run
a Tukey HSD test to determine which countries have significant
differences between mean scores.

```{r}
#| eval: TRUE
#| code-fold: false
#| echo: true
TukeyHSD(resaov)
```

From that we get a table with p values (p adj) for different pairs of
countries. Note that the p-values are below 0.05 for France-Germany, UK-France, US-UK and US-France, but greater than 0.05 for UK-Germany, US-Germany. So we can conclude there are significant differences in science scores between France-Germany, UK-France, US-UK and US-France but not between the UK and Germany, and the US and Germany.

# Seminar Tasks

## Task 1 - Choice of tests

Consider the following scenarios. Which (if any) of the t-tests would be
most of appropriate, including whether it would be a one or two tailed
test.

::: callout-note
One and two-tailed tests

Remember that a one-tailed test only looks for differences in one
direction from the mean (for example, that one sample has a higher mean
than the other). A two-tailed tests tests for the possibility of the
means of the two groups being higher or lower than each other.
:::

::: question
a)  An educational researcher wants to test a teaching approach with a
    group of 12 students to see if an intervention increases performance
    in their year 6 SATs results. Pupils are given a SATS paper before
    the intervention and again after. The results of these are
    normalised and then compared.

b)  A teacher wants to see if their pupils' GCSE computer science test
    scores are in line with the national average or not. The tests are
    out of 180 marks.

c)  The DfE want to compare uptake of STEM subjects pre and post
    pandemic to see if there has been any change in the percentage of
    students studying at least one STEM subject for A level.

d)  A school wants to compare salaries of staff between this academic
    year and last academic year to see whether there has been a
    significant increase.

e)  The same school as in d) wants to compare salaries of support staff
    and teaching staff to see if there is a significant difference.

f)  A researcher wants to find out how much time pupils spend looking at
    screens during lesson time, to see if there is a difference between
    year 7 and year 10 students. 10 pupils in year 7 and 10 pupils in
    year 10 are randomly selected across a school and observed
    throughout a school day in November. The times are recorded to the
    nearest minute.

g)  A business manager wants to check if the financial expenditure on
    STEM teachers per year is similar to the national average (based on
    average numbers of pupils on role)
:::

```{r}
#| eval: false
#| code-fold: true
#| code-summary: "Answer"
#| warning: false
#| echo: true

# Answers

# a)  An educational researcher wants to test a teaching approach with a group of 12 students to see if an intervention increases performance in their year 6 SATs results. Pupils are given a SATS paper before the intervention and again after. The results of these are normalised and then compared.

# Paired (because we compare pairs of results (before and after) for the same student) and two-tailed (because we are interested in the means being higher or lower)

# b)  A teacher wants to see if their pupils' GCSE computer science test scores are in line with the national average or not. The tests are out of 180 marks.

# Unpaired (the teacher compares their students with another pool) and two-tailed (because we are interested in the means being higher or lower)

# c)  The DfE want to compare uptake of STEM subjects pre and post pandemic to see if there has been any change in the percentage of students studying at least one STEM subject for A level.

# Unpaired (the students taking STEM subjects before and after the pandemic will be different individuals from different cohorts) and two-tailed (because we are interested in the means being higher or lower)

# d)  A school wants to compare salaries of staff between this academic year and last academic year to see whether there has been a significant increase.

# Possibly paired (if the staff body has remained the same), and two-tailed (because we are interested in the means being higher or lower)

# e)  The same school as in d) wants to compare salaries of support staff and teaching staff to see if there is a significant difference.

# Unpaired (because there are two different groups, teachers and support staff), and two-tailed (because we are interested in the mean salaries being higher or lower)

# f)  A researcher wants to find out how much time pupils spend looking at screens during lesson time, to see if there is a difference between year 7 and year 10 students. 10 pupils in year 7 and 10 pupils in year 10 are randomly selected across a school and observed  throughout a school day in November. The times are recorded to the nearest minute.

# Unpaired (because there are two different groups, Y7 and Y10), and two-tailed (because we are interested in the means of times being higher or lower)

# g)  A business manager wants to check if the financial expenditure on STEM teachers per year is similar to the national average (based on average numbers of pupils on role)

# Unpaired (because there are two different groups, the school and the national cohort), and two-tailed (because we are interested in the expenditure means being higher or lower)

```

## Task 2 - Performing t-tests

::: question

1. The OECD set the mean score for the achievement tests at 500. For students in the
    UK determine if the science scores (`PV1SCIE`) are statistically significant from     500.


```{r}
#| eval: false
#| warning: false
#| code-fold: true
#| code-summary: "Answer"
#| echo: true

# Are UK science scores statistically significantly different from 500?

# Select the gender (ST004D01T) and maths score columns (PV1SCIE)
# Filter the data to select UK responses

UK_sci <- PISA_2022 %>%
  select(CNT, ST004D01T, PV1SCIE) %>%
  filter(CNT == 'United Kingdom')

# perform a one sample t-test against a mean of 500

t.test(UK_sci$PV1SCIE, mu = 500)
# p-value < 2.2e-16 - true mean is not 500


```


2.  Use the PISA dataset and the gender `ST004D01T` and `PV1MATH` (Math
    score) variables to perform a two-sided, unpaired t-test to
    determine if boys and girls perform differently on the test in the
    UK. Before performing the test, check for the normality of the data,
    and check for equal variance. Find an effect size.

```{r}
#| eval: false
#| warning: false
#| code-fold: true
#| code-summary: "Answer"
#| echo: true

# Are there differences between the mean scores of UK boys and girls in PISA mathematics?

library(effsize) # for cohen's d

# Select the gender (ST004D01T) and maths score columns (PV1MATH)
# Filter the data to select UK responses

MaleUK <- PISA_2022 %>%
  select(CNT, ST004D01T, PV1MATH) %>%
  filter(CNT == 'United Kingdom',
         ST004D01T == 'Male')

FemaleUK <- PISA_2022 %>%
  select(CNT, ST004D01T, PV1MATH) %>%
  filter(CNT == 'United Kingdom',
         ST004D01T == 'Female')

# The conditions to do a t-test include that the data are normally distributed
# and there is homogeneity (similarity) of the variances (the squared standard deviations)
# Let us check the conditions are met by calculating first if the data sets are normally
# distributed using a qq plot

qqnorm(MaleUK$PV1MATH)
qqline(MaleUK$PV1MATH, col = "red")

qqnorm(FemaleUK$PV1MATH)
qqline(FemaleUK$PV1MATH, col = "red")

# The plots produces relatively straight lines so the distributions can be assumed to be normal
#
# We will then check the variances of the two data sets

VarM <- var(MaleUK$PV1MATH, na.rm = TRUE)
VarF <- var(FemaleUK$PV1MATH, na.rm = TRUE)
VarM / VarF

# The variance ratio is close to 1 (1.1)
# So our two conditions are met and can we can perform the t-test


# Plot a two-sided, unpaired t-test
t.test(MaleUK$PV1MATH, FemaleUK$PV1MATH, 
       paired = FALSE, alternative = "two.sided")

# The p-value is <0.05 (4.061e-08) suggesting there are statistically
# differences between boys and girls

# Calculate the effect size using Cohen's d
cohen.d(MaleUK$PV1MATH, FemaleUK$PV1MATH)
# The effect size is 0.144 which is negligible in Cohen's terms (<0.2) but medium in Kraft and colleagues; terms (0.05-0.2)



```
:::

::: question
3.  Choose a different country of interest to compare performance in
    males and females in that country

```{r}
#| eval: false
#| warning: false
#| code-fold: true
#| code-summary: "Answer"
#| echo: true
# Are there differences between the mean scores of boys and girls in Japan PISA mathematics?
#
# Select the gender (ST004D01T) and math score columns (PV1MATH)
# Filter the data to select UK responses

MaleJapan <- PISA_2022 %>%
  select(CNT, ST004D01T, PV1MATH) %>%
  filter(CNT == 'Japan', ST004D01T == 'Male')

FemaleJapan <- PISA_2022 %>%
  select(CNT, ST004D01T, PV1MATH) %>%
  filter(CNT == 'Japan', ST004D01T == 'Female')

# The conditions to do a t-test include that the data are normally distributed
qqnorm(MaleJapan$PV1MATH)
qqline(MaleJapan$PV1MATH, col = "red")

qqnorm(FemaleJapan$PV1MATH)
qqline(FemaleJapan$PV1MATH, col = "red")


# The plots produces relatively straight lines so the distributions can be # assumed to be normal
#
# We will then check the variances of the two data sets

VarM<-var(MaleJapan$PV1MATH, na.rm = TRUE)
VarF<-var(FemaleJapan$PV1MATH, na.rm = TRUE)
VarM/VarF

# The variance ratio is close to 1 (1.26)
# So our two conditions are met and can we can perform the t-test

# Create a Japan Math subset
CHIMATH <- PISA_2022 %>%
  select(CNT, PV1MATH, ST004D01T) %>%
  filter(CNT == "Japan")
# Plot a two-sided, unpaired t-test
t.test(MaleJapan$PV1MATH, FemaleJapan$PV1MATH,  
       paired = FALSE, alternative = "two.sided")

# The p-value is <0.05 (0.0004128) suggesting there are statistically
# differences between boys and girls in mathematics in Japan

# Calculate the effect size using Cohen's d

cohen.d(MaleJapan$PV1MATH, FemaleJapan$PV1MATH)

# The effect size is 0.09 which is negligible in Cohen's terms (<0.2) and medium in Kraft and colleagues' terms (<0.05)


```
:::

::: question
4.  Now try to compare the performance of males and females in Maths in
    all the `OECD` countries in the PISA data. You may need to think
    about how to do this using R without repeating the test for each
    country.

```{r}
#| eval: false
#| code-fold: true
#| code-summary: "Answer"
#| echo: true

MaleOECD <- PISA_2022 %>%
  select(OECD, ST004D01T, PV1MATH) %>%
  filter(OECD == 'Yes', ST004D01T == 'Male')

FemaleOECD <- PISA_2022 %>%
  select(OECD, ST004D01T, PV1MATH) %>%
  filter(OECD == 'Yes', ST004D01T == 'Female')

qqnorm(MaleOECD$PV1MATH)
qqline(MaleOECD$PV1MATH, col = "red")

qqnorm(FemaleOECD$PV1MATH)
qqline(FemaleOECD$PV1MATH, col = "red")


# The plots produces relatively straight lines so the distributions can be assumed to be normal

# We will then check the variances of the two data sets

VarM <- var(MaleOECD$PV1MATH, na.rm = TRUE)
VarF <- var(FemaleOECD$PV1MATH, na.rm = TRUE)
VarM / VarF

# The variance ratio is close to 1 (1.18)
# So our two conditions are met and can we can perform the t-test

# The p-value is <0.05 (<2.2e-16) suggesting there are statistically
# differences between boys and girls

OECD <- PISA_2022 %>%
  select(OECD, ST004D01T, PV1MATH) %>%
  filter(OECD == 'Yes')

t.test(MaleOECD$PV1MATH, FemaleOECD$PV1MATH, 
       paired = FALSE, alternative = "two.sided")

# Calculate the effect size using Cohen's d

cohen.d(MaleOECD$PV1MATH, FemaleOECD$PV1MATH)

# The effect size is 0.10 which is negligible in Cohen's terms (<0.2) and medium in Kraft and colleagues' terms (>0.05 and <0.2)


```
:::

::: question
4.  Are there statistically significant differences in mean `HOMEPOS`
    scores of boys and girls in the UK?

```{r}
#| eval: false
#| code-fold: true
#| code-summary: "Answer"
#| echo: true

MaleUK <- PISA_2022 %>%
  select(ST004D01T, HOMEPOS) %>%
  filter(ST004D01T == 'Male') %>%
  na.omit()

FemaleUK <- PISA_2022 %>%
  select(ST004D01T, HOMEPOS) %>%
  filter(ST004D01T == 'Female') %>%
  na.omit()

qqnorm(MaleUK$HOMEPOS)
qqline(MaleUK$HOMEPOS, col = "red")
qqnorm(FemaleUK$HOMEPOS)
qqline(FemaleUK$HOMEPOS, col = "red")

# The plots here are somewhat skewed which might suggest not performing the test

# We will then check the variances of the two data sets

VarM <- var(MaleUK$HOMEPOS, na.rm = TRUE)
VarF <- var(FemaleUK$HOMEPOS, na.rm = TRUE)
VarM / VarF

# The variance ratio is close to 1 (1.1)
# So our two conditions are met and can we can perform the t-test

HOMEPOS <- PISA_2022 %>%
  select(HOMEPOS, ST004D01T, CNT) %>%
  filter(CNT == 'United Kingdom')

t.test(MaleUK$HOMEPOS, FemaleUK$HOMEPOS, 
       paired = FALSE, alternative = "two.sided")

# p-value < 2.2e-16 so there are statistically significant differences between boys' and girls' reported wealth in the UK

# Calculate the effect size using Cohen's d
cohen.d(MaleUK$HOMEPOS, FemaleUK$HOMEPOS)
# The effect size is 0.003 which is negligible in Cohen's terms (<0.2) and small in Kraft and colleagues' terms (<0.05)

```
:::

## Task 3 - Multiple t-tests with anova

::: question
For the UK, France, Thailand, and the US determine if there are
statistically significant differences in variation in mathematics
scores.

```{r}
#| code-fold: true
#| code-summary: answer
#| eval: true
#| echo: true

# create a dataframe for the four countries and their mathematics scores
PISAMULTI <- PISA_2022 %>%
  select(CNT, PV1MATH)%>%
  filter(CNT == "United Kingdom" | CNT == "Thailand" | CNT == "United States" |
         CNT == "France")

# perform the anova
resaov <- aov(PV1MATH ~ CNT, data = PISAMULTI)

# get a summary of results and print
sumresaov <- summary(resaov)
sumresaov

# calculate the eta squared effect size
eta <- as.data.frame(etaSquared(resaov))
eta <- eta*100
eta

# perform a post-hoc test
TukeyHSD(resaov)
# There are statistically significant differences in mathematics scores for all pairs of countries

```
:::

::: question
For the UK, are there statistically significant differences in reading
score by students who speak different languages (LANGN)

```{r}
#| code-fold: true
#| code-summary: answer
#| eval: true
#| echo: true


# Create a data frame with Reading and school type variables, filte for the uK
UKLangRead<-PISA_2022 %>%
  select(CNT, PV1READ, LANGN) %>%
  filter(CNT == "United Kingdom")

# Perform an anova test for reading scores, school type

resaov <- aov(PV1READ ~ LANGN, data = UKLangRead)
sumresaov <- summary(resaov)
sumresaov

# Calculate eta squared

eta <- as.data.frame(etaSquared(resaov))
eta <- eta*100
eta

# Perform the post-hoc Tukey test to determine where signficiant differences lie

TukeyHSD(resaov)

# There are statistically significant differences between Pre-vocational and general, and vocational and pre-vocational, but not between vocational and general.

```
:::

::: question
For the UK, use `ifelse` on the `HOMEPOS` variable to create categories for
four quartiles (divide the data into top; upper-mid; lower-mid; and
bottom). Use anova to determine between which groups there are
statistically significant differences in science scores

::: callout-tip Hint: You can use
`quantiles <- quantile(UKSci$PV1SCIE, prob=c(.25,.5,.75), na.rm=TRUE)` to
give the values of PV1SCIE that account for 25%, 50% and 75% of the
responses.
:::

```{r}
#| code-fold: true
#| code-summary: answer
#| eval: true
#| echo: true

# Create a data frame of UK science scores and Wealth data
# Drop levels to remove other countries
UKSci <- PISA_2022 %>%
  select(CNT, PV1SCIE, HOMEPOS) %>%
  filter(CNT == "United Kingdom") %>%
  na.omit() %>%
  droplevels() %>%
  select(PV1SCIE, HOMEPOS)

# Find the cut of points for 25%, 50% and 75% (quantiles) of HOMEPOS

quantiles <- quantile(UKSci$HOMEPOS, prob = c(.25,.5,.75), na.rm=TRUE)

# Note the results are stored in quantiles[1] (25%), quantiles[2] 
# (50%), quantiles[3] (75%)

# Mutate the HOMEPOS column, using ifelse, to substitute the labels, "Bottom",
# "Mid", "upper-mid" and "top" for the quantiles.

UKSci<-UKSci%>%
  mutate(HOMEPOS = ifelse(HOMEPOS<quantiles[1], "Bottom", 
                        ifelse(HOMEPOS < quantiles[2], "Lower-mid",
                               ifelse(HOMEPOS < quantiles[3],
                                      "Upper-mid","Top"))))

# Perform the avova for science scores by wealth, categories into 4 groups

resaov<-aov(PV1SCIE ~ HOMEPOS, data = UKSci)
sumresaov<-summary(resaov)
sumresaov

# p is <2e-16 so significant differences exist between wealth groups on science
# scores

# perform the post hoc Tukey test to find out where statistically signficant differences exist

TukeyHSD(resaov)

# There are statistically significant differences between all groups by wealth

```

:::

# Extension and additional reading

## Recreating a paper in R

::: question
Looking at @cook2014visiphilia, consider the following

<!-- -->

1.  Is it possible to replicate the results using the PISA 2018 data
    set? Make a graph showing the difference in male and female
    `PV1MATH` results for each country `CNT`. To do this we are going to
    have to:

-   work out the `mean()` maths score `PV1MATH` for each country `CNT`
    and gender `ST004D01T` grouping, call this `meanmath` and create a
    new data frame to store this
-   from this data frame, create two new data frames, one for males only
    and one for females only
-   `rename()` (see @sec-renaming) the `meanmath` score in each data
    frame to `male_mean` and `female_mean`
-   bind the data frames together using using column bind function
    `cbind(<male_df>, <female_df>)` and store this in a new dataframe
    called `Mathgendergap`. **NOTE:** cbind only accepts tables with
    different names, so you'll need to select `select(CNT, male_mean)`
    from the male dataframe and `select(female_mean)` from the female
    dataframe.
-   use `mutate` (see @sec-mutate) to calculate the difference in male
    and female mean maths scores for each country
-   plot the results for each country

```{r}
#| eval: false
#| warning: false
#| code-fold: true
#| code-summary: "creating difference dataset"
#| echo: true
# A relatively simple recreation (without significance testing)
Mathgendergap <- PISA_2022 %>%
  select(CNT, PV1MATH, ST004D01T) %>%
  group_by(CNT, ST004D01T) %>%
  summarise(meanmath=mean(PV1MATH)) %>%
  ungroup()

# alternative, using column binding, cbind():
Mathgendergap<-  cbind(Mathgendergap %>% 
                           filter(ST004D01T == "Male") %>%
                           rename(male_mean = meanmath) %>%
                             select(CNT, male_mean),
                        Mathgendergap %>%
                           filter(ST004D01T == "Female") %>%
                             rename(female_mean = meanmath) %>%
                             select(female_mean))

# alternatively you can use pivot_wider
# Mathgendergap <- pivot_wider(Mathgendergap, names_from = ST004D01T, values_from = meanmath)

Mathgendergap <- Mathgendergap %>%
  mutate(difference = female_mean - male_mean) %>%
  arrange(desc(difference))
```

```{r}
#| eval: false
#| warning: false
#| code-fold: true
#| code-summary: "the graph"

ggplot(Mathgendergap,
       aes(x = reorder(CNT, difference), y = difference, fill = difference))+
  geom_col() +
  geom_hline(yintercept = 0, lty=2) +  # add a line on 0
  coord_flip() +                      # rotate the graph
  theme(axis.text.x = element_text(angle = 90, vjust = 0.5, hjust = 1, size = 5))
```

The above code doesn't include the t-test results for each country, to
do this we need to run some more complex code, you can see how it works
below:

```{r}
#| eval: true
#| warning: false
#| echo: true
#| code-fold: true
#| code-summary: "code"

# A fuller recreation with t-tests
# conduct a ttest across countries on a specified column
library(broom)

ttest_by_country <- function(data, column = PV1MATH){

  # work out which countries have full 30+ datasets for this ttest
  countries <- data %>% ungroup() %>%
    filter(!is.na({{column}})) %>%    # {{column}} allows you to change the field of focus
    select(CNT, ST004D01T, {{column}}) %>%
    group_by(CNT) %>%
    filter(n() > 30) %>%
    pull(CNT) %>%  # the pull command returns the column as a vector, not a table
    unique()
 
  # list the countries that don't meet that criteria
  message("dropping: ", setdiff(unique(data$CNT), countries), " as too few entries for ttest")
 
  # reduce the dataset to only those countries with 30+ entries
  data <- data %>%
    filter(CNT %in% countries)

  # for each country in this new dataset perform a set of calculations
  test_result <- map_df(unique(data$CNT),
                        function(x){
   
                            # make a subset of the data just for that country
                            df <- data %>% filter(CNT == x)
                           
                            # get the results pull({{column}}) for females and males as two separate vectors
                            f_data <- df %>% filter(ST004D01T == "Female") %>% pull({{column}})
                            m_data <- df %>% filter(ST004D01T == "Male") %>% pull({{column}})
                           
                            # tell us the number of results
                            message(x, " f:", length(f_data), " m:", length(m_data))
                           
                            # work out the means of each vector
                            f_mean <- mean(f_data)
                            m_mean <- mean(m_data)
                           
                            t.test(m_data, f_data) %>%  # conduct a ttest on the male and female results
                              tidy() %>%      # convert the ttest result into a dataframe
                              mutate(CNT = x,           # add columns to record the country
                                     f_mean = f_mean,   # the mean female grade
                                     m_mean = m_mean,   # the mean male grade
                                     gender_diff = m_mean - f_mean,  # the difference between the two
                                     prop_male = length(m_data) / (length(m_data) + length(f_data)))
                                     # and the proportion who are male in the dataset
    })
  return(test_result)
}

plot_ttest_by_country <- function(data, column = "PV1MATH"){

  ggplot(data %>% mutate(sig = p.value < 0.05),
         aes(x=reorder(CNT, gender_diff), y=gender_diff, colour=sig))+
    geom_point(aes(size = prop_male)) +  
    geom_hline(yintercept = 0, lty=2) +  # add a line on 0
    coord_flip() +     # rotate the graph
    xlab("country") +
    ylab("mean(male - female)") +
    ggtitle(paste("gender differences for:", column)) +
    theme(axis.text.x = element_text(angle = 90, vjust = 0.5,
                                     hjust=1, size=5))
}

# run the first function using 2018 data and the PV1MATH column
ttest_results <- ttest_by_country(PISA_2022, PV1MATH)

#plot the results
plot_ttest_by_country(ttest_results, "PV1MATH")

# Alternatively, you could run the following:
# run a ttest for each country
# ttest_results <- PISA_2022 %>%
#     filter(!is.na(PV1MATH)) %>% # Vietnam no results?!
#     select(CNT, ST004D01T, PV1MATH) %>%
#     group_by(CNT) %>%
#     nest(data = c(ST004D01T, PV1MATH)) %>% #create a dataframe of gender results for each country
#     summarise(tt = map(data, function(df){ # apply a ttest to each country
#       t.test(df %>% filter(ST004D01T == "Female") %$% PV1MATH,
#              df %>% filter(ST004D01T == "Male") %$% PV1MATH) %>%
#         tidy() # convert results into a dataframe
#     })) %>%
#     unnest(tt)
```

2.  What issues are there with using a t-test for the context given in
    the paper?
3.  How do your findings from question 3 and the 2018 dataset compare
    with those in the paper? Are there any differences or disagreements
    with your findings?
4.  How could the paper be improved?
:::

# Doing t-tests in R

<iframe width="560" height="315" src="https://www.youtube.com/embed/KdhAcLPq5B4?start=3" title="YouTube video player" frameborder="0" allow="accelerometer; autoplay; clipboard-write; encrypted-media; gyroscope; picture-in-picture" allowfullscreen>

</iframe>

You can find the code from the video below:

```{r}
#| label: ttest video code
#| eval: false
#| code-fold: true
#| code-summary: "Show the Code"
#| echo: true

# Introduction to t-tests in R
#
# Download data from /Users/k1765032/Library/CloudStorage/GoogleDrive-richardandrewbrock@gmail.com/.shortcut-targets-by-id/1c3CkaEBOICzepArDfjQUP34W2BYhFjM4/PISR/Data/PISA/subset/Students_2018_RBDP_none_levels.rds
# You want the file: Students_2018_RBDP_none_levels.rds
# and place in your own file system
# change loc to load the data directly. Loading into R might take a few minutes
install.packages("nortest")

library(tidyverse)
library(nortest)
loc <- "/Users/k1765032/Library/CloudStorage/GoogleDrive-richardandrewbrock@gmail.com/.shortcut-targets-by-id/1c3CkaEBOICzepArDfjQUP34W2BYhFjM4/PISR/Data/PISA/subset/Students_2022_RBDP_none_levels.rds"
PISA_2022 <- read_rds(loc)

# Are there differences between the mean scores of UK boys and girls in PISA mathematics?
#
# Select the gender (ST004D01T) and math score columns (PV1MATH)
# Filter the data to select UK responses

MaleUK <- PISA_2022 %>%
  select(CNT, ST004D01T, PV1MATH) %>%
  filter(CNT == 'United Kingdom') %>%
  filter(ST004D01T == 'Male')

FemaleUK <- PISA_2022 %>%
  select(CNT, ST004D01T, PV1MATH) %>%
  filter(CNT == 'United Kingdom') %>%
  filter(ST004D01T == 'Female')

# The conditions to do a t-test include that the data are normally distributed
# and there is homogeneity (similarity) of the variances (the squared standard deviations)
# Let us check the conditions are met by calculating first if the data sets are normally
# distributed using the Pearson test of normality from the nortest package

pearson.test(as.numeric(MaleUK$PV1MATH))
pearson.test(as.numeric(FemaleUK$PV1MATH))

# The p-values are over 0.05 so both distriburtions are normal
# Pearson chi-square normality test
#
# data:  as.numeric(MaleUK$PV1MATH)
# P = 75.714, p-value = 0.1936
# Pearson chi-square normality test
#
# data:  as.numeric(FemaleUK$PV1MATH)
# P = 74.06, p-value = 0.2589
#
# We will then check the variances of the two data sets

VarM <- var(MaleUK$PV1MATH)
VarF <- var(FemaleUK$PV1MATH)
VarM/VarF

# The variance ratio is close to 1 (1.1)
# So our two conditions are met and can we can perform the t-test

t.test(MaleUK$PV1MATH, FemaleUK$PV1MATH)

# The p-value is <0.05 (4.061e-08) suggesting there are statistically
# differences between boys and girls
```
