---
title: "07 Hypothesis testing and Chi-square tests"

---

```{r setup}
    options(max.print=15)

    library(here)
    source(here::here("MASTEMR_page_setup.R"))
```



## Optional Pre-reading

Before the session, you may find it useful to read chapter 6, the basic elements of hypothesis testing, in @geher2014straightforward ([available here](https://ebookcentral.proquest.com/lib/kcl/reader.action?docID=1692209&ppg=136)).

## Pre-session task - Loading the data

We will continue to use the PISA_2022 dataset, make sure it is loaded.

```{r}
#| eval: false
#| code-fold: false
#| code-summary: "Show the code"
#| echo: true
# Load PISA data

library(arrow)
library(tidyverse)

PISA_2022 <- read_parquet(r"[<folder>PISA_2022_student_subset.parquet]")

```


# Chi-square tests

Chi squared ($\chi^2$) tests are *non-parametric* tests, this means that the test isn't expecting the underlying data to be distributed in a certain way. Chi-squared determines how well the frequency distribution for a sample fits the population distribution and will let you know when things aren't distributed as expected. For example you might expect girls and boys to have the same coloured dogs, a chi squared test can tell you whether the null hypothesis, that there is no difference between the colours of dogs owned by girls and boys, is true or not.

In more mathematical terms, chi squared examines differences between the categories of an independent variable with respect to a dependent variable measured on a nominal (or categorical) scale. A nominal scale has values that aren't ordered, or continuous, for example gender or favourite flavour of ice cream.

## Conditions of Chi-Square Tests

Four assumptions need to be met in order to use a chi-square test:

a\) The data (both variables) should be categorical (for ordinal data, see the section on Kruskal Wallis tests below);

b\) All observations are independent;

c\) Cells in the contingency table (see below) are mutually exclusive;

d\) Expected values in each cell in the contingency table should be five or greater for more than 80% of cells.

See [Section 12.5](https://learningstatisticswithr.com/book/chisquare.html#chisqassumptions) in Navaro's Learning Statistics with R

## Types of Chi-Square Tests

Chi-square tests can be categorised in two groups:

-   *A test of goodness of fit* - this is a form of hypothesis test which determines whether a sample fits a wider population. For example, does the pattern of exam results in one school fit the national distribution?
-   *A test of independence* - allows inference to be made about whether two categorical variables in a population are related. For example, are there differences in the uptake of careers by gender?

For more information on chi-square tests, see [chapter 12](https://learningstatisticswithr.com/book/chisquare.html), in Navaro's Learning Statistics with R.

# Performing chi square tests

## Creating contingency tables

Chi-square calculations depend on contingency tables. A contingency table is a table that shows the frequency counts for two variables. We can use the `xtabs` function to create contingency tables in R.

For example, imagine we want to create a contingency table for the number of boys and girls in the UK and US in the PISA sample. First we create a subset of the `PISA_2022` `data.frame` including country and gender, and `filter` for the two countries. We use the `xtabs` function to create the table. We pass the subset data (`UKUSgender`) to `xtabs` and indicate the columns and rows we want `~CNT + ST004D01T`

```{r}
#| eval: true
#| code-fold: false
#| code-summary: "Show the code"
#| echo: true
# Example contingency table

# First create a data frame with the gender data for the UK and US 

UKUSgender <- PISA_2022 %>%
  select(CNT, ST004D01T) %>%                                   # <1>
  filter(CNT == "United Kingdom" | CNT == "United States") %>% # <2>
  droplevels()                                                 # <3>   

# The use xtabs to create a contingency table by ('~') country (CNT) and gender (ST004D01T)

ContTable <- xtabs(data = UKUSgender, ~ CNT + ST004D01T)        # <4> 
ContTable

```

1.  `select` the needed columns gender (`ST004D01T`) and country (`CNT`)
2.  `filter` for countries of interest, the UK and the US
3.  `droplevels()` to avoid empty country levels (those not selecting) cluttering the table at 0 counts
4.  use `xtabs` to create a contingency table comparing `CNT` with `ST004D01T`

# Chi-square goodness of fit tests

If we want to determine if a sample categorical data matches the pattern of a whole population we can use a Chi-square goodness of fit test. The test is a form of hypothesis testing.

Hypothesis testing is one type of statistical analysis. A researcher states an assumption that they want to test.

For example, they might want to examine whether the distribution of boys and girls in the UK matches the expected distribution of 50:50.

A researcher typically proposes a null hypothesis - that is that there is no difference in groups. Whilst this is typical practice, and we will follow it in this course, researchers have pointed out that it is rare for the null hypothesis to be true, which impacts the validity of the test (see: @cohen1994earth) . Nonetheless, we will adopt the practice here, as it is a widely used approach.

Our null hypothesis then is:

*There is no difference in the distribution of boys and girls in the UK and a random sample of 50/50 girls and boys*.

Notice the 'goodness of fit' element - we are checking if some categorical data from a sample population (the UK) fits an expected pattern.

The outcome of a hypothesis test is typically reported by stating the value of some test (the test statistic, in this case, the Chi-squared statistic) which is used to calculate a significance level (or p-value). The p-value is the probability of obtaining test results at least as extreme as the result actually observed, under the assumption that the null hypothesis is correct.

This assumption has been critiqued, but in many research traditions, if the p-value is less than 0.05 (p\>0.05) the result is taken to be statistically significant.

In the case of a Chi-squared goodness of fit test:

• If the p-value is ***greater than 0.05*** (p\>0.05) ***we accept the null hypothesis*** that the sample has been drawn from the wider population.

• If the p-value ***is less than 0.05*** (p\<0.05) ***we reject the null hypothesis*** that the sample has been drawn from the wider population.

::: callout-tip
It is important that care is taken when interpreting p-values. A p-value of below 0.05 does not mean the null hypothesis is false. Monya Baker provided a helpful summary of how to think of p-values:

"A P value of 0.05 does not mean that there is a 95% chance that a given hypothesis is correct. Instead, it signifies that if the null hypothesis is true, and all other assumptions made are valid, there is a 5% chance of obtaining a result at least as extreme as the one observed. And a P value cannot indicate the importance of a finding; for instance, a drug can have a statistically significant effect on patients' blood glucose levels without having a therapeutic effect."

See @baker2016statisticians for further discussion of how to interpret p-values.
:::

## An example: Does the disritbution of male and female students in the UK fit the expected pattern (50:50)?

To perform the goodness of fit test, we make a subset `dataframe` of the UK data including the `ST004D01T` (gender variable).

```{r}
#| eval: true
#| code-fold: false
#| code-summary: "Show the code"
#| echo: true
# Perform a chi-square goodness of fit test on categorical data related to gender in the UK

# Create a data frame of UK genders

UKPISAgender <- PISA_2022 %>%
  select(CNT, ST004D01T) %>%                                   # <1>
  filter(CNT == "United Kingdom") %>%                          # <2>
  droplevels()                                                 # <3>

# Use cross tabs to create a contingency table of the dataframe, by ST004D01T (gender)

GenderContTable <- xtabs(data = UKPISAgender, ~ ST004D01T)   # <4>

# Perfom the chisq test, comparing against the expected probabilities of 50:50 (e.g. 1/2 to 1/2)

chisq.test(GenderContTable, p=c(1/2, 1/2))                      # <5>


```
1.  `select` the needed columns gender (`ST004D01T`) and country (`CNT`)
2.  `filter` for country of interest, the UK
3.  `droplevels()` to avoid empty country levels (those not selecting) cluttering the table at 0 counts
4.  use `xtabs` to create a contingency table comparing `CNT` with `ST004D01T`
5.  perform the chi-sq test of goodness of fit, comparing against an expected probability of 50:50

The outcome of the chi squared test returns a p-value = 0.1181. This is greater than 0.05, suggesting we accept the null hypothesis, and the numbers of boys and girls in the UK sample matches a 50:50 distribution.

## An example: Gender distribution in the PISA sample

::: question
a)  Should we accept or reject the hypothesis that the populations of boys and girls in the United States, Japan and Korea are 50:50?

```{r}
#| eval: false
#| code-fold: true
#| code-summary: "Show the code"
#| echo: true
# Perform a chi-square goodness of fit test on categorical data related 
# to gender in the US, Japan and China

USPISAgender<-PISA_2022 %>%
  select(CNT, ST004D01T) %>%                                  # <1>
  filter(CNT == "United States") %>%                          # <2>
  droplevels()                                                # <3>

# produce the contingency table by country and gender

GenderContTable <- xtabs(data = USPISAgender, ~ ST004D01T) # <4>

# Perform the chisq test, comparing against the expected probabilities of 50:50 (e.g. 1/2 to 1/2)

chisq.test(GenderContTable, p = c(1/2, 1/2))                    # <5>

# In the US, p= 0.2535, accept null hypothesis of equality, population of M:F is 50:50 

JPNPISAgender <- PISA_2022 %>%
  select(CNT, ST004D01T) %>% # Select gender and country variables
  filter(CNT == "Japan") %>% # Filter for the Japan
  droplevels() # To prevent the levels for other countries confusing the table

# produce the contingency table by country and gender

GenderContTable <- xtabs(data = JPNPISAgender,~ ST004D01T)

# Perfom the chisq test, comparing against the expected probabilities of 50:50 (e.g. 1/2 to 1/2)

chisq.test(GenderContTable, p = c(1/2, 1/2))

# In Japan,  p= 0.5271, accept null hypothesis of equality, population of M:F differs from 50:50

KoreaPISAgender <- PISA_2022 %>%
  select(CNT, ST004D01T) %>% # Select gender and country variables
  filter(CNT == "Korea") %>% # Filter for Korea
  droplevels() # To prevent the levels for other countries confusing the table


# produce the contingency table by country and gender

GenderFreqTable <- xtabs(data=KoreaPISAgender, ~ ST004D01T)

# Perform the chisq test, comparing against the expected probabilities of 50:50 (e.g. 1/2 to 1/2)

chisq.test(GenderFreqTable, p = c(1/2, 1/2))
# In Korea, p=0.0147, reject null hypothesis of equality.


```

1.  `select` the needed columns gender (`ST004D01T`) and country (`CNT`)
2.  `filter` for country of interest, the US
3.  `droplevels()` to avoid empty country levels (those not selecting) cluttering the table at 0 counts
4.  use `xtabs` to create a contingency table comparing `CNT` with `ST004D01T`
5.  perform the chi-sq test of goodness of fit, comparing against an expected probability of 50:50

:::

::: callout-tip
R uses standard form: an output of p= 3.724e-06, represents, p=3.724x10^-6^, or p = 0.0000003724.
:::

## Effect sizes for chi-square tests

Chi-square tests can tell us if two categorical variables are related, but they do not tell us how strong the relationship is. To do this we can calculate an effect size. For chi-square tests, a common effect size is Cramer's V. For the example above, we can calculate Cramer's V using the `cramersV` function in the `lsr` package. Recall the test of goodnesss of fit for Does the disritbution of male and female students in South Korea fit the expected pattern (50:50)?

```{r}
#| eval: false
#| code-fold: false
#| code-summary: "Show the code"
#| echo: true

KoreaPISAgender <- PISA_2022 %>%
  select(CNT, ST004D01T) %>%                                   
  filter(CNT == "Korea") %>%                          
  droplevels()                                                 

GenderContTable <- xtabs(data = KoreaPISAgender, ~ST004D01T)   

chisq.test(GenderContTable, p=c(1/2, 1/2))  

# find the effect size using Cramer's V
library(lsr)
cramersV(GenderContTable)  
# Cramer's V = 0.03, a very small effect size

```

::: callout-tip
Interpreting Cramer's V

How to interpret Cramer's V depends on the degrees of freedom (df). The degrees of freedom is calculated as: df = (r - 1)(c - 1)

where r is the number of rows and c is the number of columns in the contingency table.

For example, a 2x2 table, df = 1. If there is only one row in the table, but several columns, df = number of columns - 1.

The interpretation of Cramer's V for different degrees of freedom is as follows:

| Degrees of Freedom (df) | Small Effect | Medium Effect | Large Effect |
|-------------------------|--------------|---------------|--------------|
| 1                       | 0.1          | 0.3           | 0.5          |
| 2-4                     | 0.07         | 0.21          | 0.35         |
| 5-9                     | 0.05         | 0.15          | 0.25         |
| 10+                     | 0.04         | 0.13          | 0.20         |

:::

# Chi-square test of independence

Goodness of fit tests can be useful, but they rely on knowing the expected distribution (for example, assuming a 50:50 distribution of boys and girls).

An alternative ways of using a Chi-square test, is the test of independence. This approach determines whether two categorical variables in a sample are related.

For example, a categorical variable in the data set is item - `WB178Q01HA` - Overall, did you feel that you accomplished something yesterday?. Which can be responded to with 'yes' or 'no' (or `<NA>`).

We might want to see if students in the Netherlands respond to this question in the same way as other countries, for example Mexico and New Zealand. That is, are students in Netherlands just as likely to feel they have accomplished something as their peers in Mexico and New Zealand.

A simple first attempt is to create frequency tables for the Netherlands and the rest of the sample and examine the responses.

```{r}
#| eval: true
#| code-fold: false
#| code-summary: "Show the code"
#| echo: true
# Produce tables of counts of accomplishment for the Netherlands and Mexico and New Zealand

# Create a data frame of accomplishment data for Mexico and New Zealand
MexNZAccomp <- PISA_2022 %>%
  select(CNT, WB178Q01HA) %>%                                 # <2>
  filter(CNT == "Mexico"| CNT == "New Zealand") %>%
  droplevels() %>%                                            # <3>
  na.omit()                                                   # <4>

# Create the contingency table for the Mexico and NX and print it

MexNZAccompAccompTable <- xtabs(data = MexNZAccomp, ~ WB178Q01HA)    # <5>
print(MexNZAccompAccompTable)

# Create the data frame and contingency table for the Netherlands and print it

NethAccomp <- PISA_2022 %>%
  select(CNT, WB178Q01HA) %>% 
  filter(CNT == "Netherlands") %>% 
  droplevels() 

NethAccompTable <- xtabs(data=NethAccomp, ~ WB178Q01HA)
print(NethAccompTable)

```

1.  select the needed columns accomplishment (`WB178Q01HA`) and country (`CNT`)
2.  `droplevels()` to avoid empty country levels (those not selecting) cluttering the table at 0 counts
3.  remove na values
4.  use `xtabs` to create a contingency table for `WB178Q01HA`

From the data, it is hard to tell if the Netherlands data is different from the overall pattern. To make things easier, we can use `mutate` to add a percentage column to aid comparison.

```{r}
#| eval: false
#| code-fold: false
#| code-summary: "Show the code"
#| echo: true
# Produce tables of counts of having accomplished for the Netherlands and for New Zealand and Mexico

# Create a data frame of accomplishment

Accomp <- PISA_2022 %>%
  select(WB178Q01HA, CNT) %>%                         # <1>
  filter(CNT == "Mexico"| CNT == "New Zealand") %>%
  droplevels()                                        # <2>

# turn the data frame into a contingency table - and then convert to a data frame for easier manipulation

AccompFreqTable <- xtabs(data = Accomp, ~ WB178Q01HA)  # <3>
AccompFreqTable <- as.data.frame(AccompFreqTable)      # <4>

# Sum to find the total count

Total <- sum(AccompFreqTable$Freq)                      # <5> 

# Use mutate to add a new column, Perc, which is Freq/Total*100 to give the percentage

AccompFreqTable <- AccompFreqTable%>%
  mutate(Perc = (Freq * 100) / Total)                # <6>
print(AccompFreqTable)

# Repeat for the Netherlands
NethAccompFreqTable <- PISA_2022 %>%
  select(CNT, WB178Q01HA) %>% # Select accomplishment and country variables
  filter(CNT == "Netherlands") %>% # Filter for the Netherlands
  droplevels() # To prevent the levels for other countries confusing the table

NethAccompFreqTable <- xtabs(data = NethAccompFreqTable, ~ WB178Q01HA)

NethAccompFreqTable <- as.data.frame(NethAccompFreqTable)

Total <- sum(NethAccompFreqTable$Freq) # Find the total count
NethAccompFreqTable <- NethAccompFreqTable%>%
  mutate(Perc = (Freq * 100)/Total)  # Mutate the table to calculate  percentage

print(NethAccompFreqTable)

```

1.  `select` the needed columns accomplishment (`WB178Q01HA`) 
2.  `droplevels()` to avoid empty country levels (those not selecting) cluttering the table at 0 counts
3.  use `xtabs` to create a contingency table for `WB178Q01HA`
4.  convert the result of `xtabs` into a data frame for easier manipulation
5.  calculate the total frequency, for use in calculating the percentage
6.  use `mutate` to add a new percentage column given by the calculation of the frequency divided by the total multiplied by 100 `(Perc = (Freq * 100) / Total)`



## Plotting the chi-square relationships

The numbers in the contingency table are hard to interpret - it is challenging to see how far out the numbers for each row are from each other. Alternatively, we can visualise the data from the contingency table by building a mosaic plot, a form of stacked bar chart. Mosaic plots can be a useful visualisation before running a chi-squared test.

To create a mosaic plot, you are going to need to install and load the `ggmosaic` package. The code below will install the package:

```{r}
#| warning: false
#| echo: true
#| eval: false

install.packages("pak")
library(pak)
pak::pak("haleyjeppson/ggmosaic")
library(ggmosaic) # load the ggmosaic package

```

Imagine we want to plot the 'accomplishment' data (`WB178Q01HA`) from the previous section for the Netherlands, Mexico and New Zealand.

To create the mosaic plot we use ggplot, as we used for previous graphs. As before, we first pass the data (in this case `Accomp`) to ggplot. Then, to create the graph, `geom_mosaic` is used. `geom_mosaic does` not have a direct mapping of input to x and y variable so we need to pass it what we want plotted on the y-axis (`WB178Q01HA`) and x-axis (`CNT`) within the product function (`product(WB178Q01HA, CNT)`). We can also specify how we want the rectangles to be coloured (in our case, by `CNT`).

```{r}
#| warning: false
#| echo: true
library(ggmosaic) # load the ggmosaic package

# Create a data frame of accomplishment data for the 3 countries
Accomp <- PISA_2022 %>%
  select(WB178Q01HA, CNT) %>%
  filter(CNT %in% c("Netherlands", "Mexico", "New Zealand")) %>%
  droplevels()

# plot results
# Note that with geom_mosaic you pass the x and y variables using: aes(x = product(WB178Q01HA, CNT))

ggplot(data = Accomp) +
  geom_mosaic(aes(x = product(WB178Q01HA, CNT), fill = WB178Q01HA)) + # <1>
  ylab("Feeling of accomplishment") + # <2>
  xlab("Country") +
  theme(axis.text.x = element_text(angle = 90, vjust = 0.5, hjust=1)) # <3>


```

1.  set the two variables to be plotted `CNT` and `WB178Q01HA` and pass the data to be plotted `(data = Accomp)` 
2.  set the title
3. rotate the text on the axes

Note that in the mosaic plot the width of the bars represents the number of students in the sample for each country.

## Running Chi-square tests of independence

The mosaic plot suggests that students' feeling of accomplishment is different in Mexico, the Netherlands and New Zealand. Simply looking at the data does not tell us if the distributions are different - a Chi-square tests of independence can report the significance level, which can help us make a judgement.

The null hypothesis in a test of independence is that the categorical variables are not related. So in the case of comparing Mexico and the Netherlands the null hypothesis is: 'There is no difference between students' sense of accomplishment in Mexico and the Netherlands'.

```{r}
#| eval: true
#| code-fold: false
#| warning: false
#| echo: true
# Produce tables of counts of having accomplished something for Mexico and the Netherlands

AccompNethMex <- PISA_2022 %>%
  select(WB178Q01HA, CNT) %>%                                    # <1>
  filter(CNT == "Netherlands"| CNT == "Mexico") %>%              # <2>
  droplevels()                                                   # <3>

# Produce the contingency table for the NZ and Mexico

AccompNethMexFreqTable <- xtabs(~ WB178Q01HA + CNT, data = AccompNethMex)  # <4>

# Print the table
print(AccompNethMexFreqTable)

# Perform Chisq test between NZ and Mexico

chisq.test(AccompNethMexFreqTable)                                   # <5>

# p-value < 0.01427, less than 0.05, so there is a statistically significant differences in feelings of accomplishment between NZ and Mexico, the null hypothesis is rejected


```
1.  select the accomplishment (`WB178Q01HA`) and country variable 
2.  filter for New Zealand and Mexico
3.  drop levels to keep the table clean
4.  use `xtabs` to create the contingency table for accomplishment (`WB178Q01HA`)  by country
5.  perform the chi.sq test

The test here returns a p-value=0.01427. This is less than 0.05 so suggests we do not accept the null hypothesis. We assume then there is a difference between feelings of accomplishment in New Zealand and Mexico.

As with the goodness of fit test, we can calculate Cramer's V to determine the effect size of the relationship.

```{r}
#| eval: true
#| code-fold: false
#| warning: false
#| echo: true
# Perform Cramer's V to find the effect size of the relationship between country and feeling of accomplishment

library(lsr)
cramersV(AccompNethMexFreqTable)  
# Cramer's V = 0.02, a very small effect size
```


# Testing ordinal data - the Kruskal Wallis test

An assumption of a chi-square test is that the data are categorical. Some of the items in the PISA are a type of categorical data which come in naturally ordered sequence - ordinal data. For example, gender is a categorical variable with no preferred order to responses: female or male. By contrast, the answer to a question: How many books do you have in your home? 0-10; 11-100; 101-200; More than 200, is ordinal data.

Though there is some debate among statisticians, if testing ordinal data, it is recommended you use an alternative to the chi-square test, the Kruskal Wallis test, which functions in a similar manner. It is called using the `kruskal.test` function. Unlike the chi square test, you pass it the raw data, rather than a contingency table.

For example,`ST251Q01JA` asks students to report the number of cars in their home, ordinal data. To carry out the Kruskal Wallis on differences in car ownership in the UK by gender we create a `data.frame` of responses to `ST012Q02TA` in the UK by gender. Unlike the chi square test, there is no need to create a contingency table and we just pass the data frame to `kruskal.test`, specifying we want to compare number of cars (`ST251Q01JA`) to gender (`ST004D01T`): `kruskal.test(data=CarsUKGender, ST251Q01JA ~ ST004D01T)`

```{r}
#| eval: true
#| code-fold: false
#| code-summary: "Show the code"
#| echo: true

# Create a data frame including data on cars and gender for the UK

CarsUKGender <- PISA_2022%>%
  select(CNT, ST004D01T, ST251Q01JA ) %>%                 # <1> 
  filter(CNT == "United Kingdom") %>%                     # <2> 
  select(ST004D01T, ST251Q01JA ) %>%                       # <3> 
  droplevels()                                            # <4> 

kruskal.test(data = CarsUKGender, ST251Q01JA ~ ST004D01T) # <5> 

# The p-value is more than 0.05 (p-value=0.8736), therefore we accept the null hypothesis that the number of cars is the same for boys and girls
```

1.  choose country, no of cars and gender 
2.  filter for the UK
3.  drop country variable now filtering is done
4.  remove other countries which exist as factors (`WB178Q01HA`)  by country
5.  perform the kruskal test

## Effect sizes with ordinal data

As with chi-square tests, it is possible to calculate an effect size for the Kruskal Wallis test. A common effect size is eta squared, which can be calculated using the `kruskal_effsize` function in the `rstatix` package.

```{r}
#| eval: true
#| code-fold: false
#| code-summary: "Show the code"
#| echo: true
# Perform eta squared to find the effect size of the relationship

library(rstatix)
kruskal_effsize(CarsUKGender, ST251Q01JA ~ ST004D01T)  
# eta squared = 0.000008, a very small effect size
```

## Post-hoc tests after a Kruskal Wallis test

Just like with the Tukley test after an ANOVA, if you find a significant result with a Kruskal Wallis test, you can perform post-hoc tests to determine which groups differ from each other. A common post-hoc test for the Kruskal Wallis is the Dunn test, which can be performed using the `dunn_test` function in the `rstatix` package.

```{r}
#| eval: true
#| code-fold: false
#| code-summary: "Show the code"
#| echo: true
# Perform a Dunn test to find which groups differ from each other
# install.packages("rstatix")
library(rstatix)

Cars_countries <- PISA_2022%>%
  select(CNT, ST251Q01JA ) %>%                 
  filter(CNT %in% c("United Kingdom", "United States", "Peru", "Cambodia")) %>%                     
  droplevels()        

dunn_test(Cars_countries, ST251Q01JA ~ CNT, p.adjust.method = "bonferroni")
# There are statistically significant associations in car ownership between the UK and Peru, the UK and Cambodia and the US, but not between the Cambodia and Peru


```


# Seminar Tasks

## Task 1 - Creating contingency tables

::: question
a)  Create a contingency table for UK, Germany and France levels of maternal education (`ST005Q01JA`). In which countries are most mothers (in total) educated to post school level?

::: callout-tip
The responses to `ST005Q01JA` are:

| **ISCED Level**                        | **Description**                       |
|-----------------------------------|------------------------------------|
| \<ISCED level 3.4\>                    | Post-secondary non-Tertiary Education |
| \<ISCED level 3.3\>                    | Upper Secondary Education             |
| \<ISCED level 2\>                      | Lower Secondary Education             |
| ISCED level 1\>                        | Primary Education                     |
| She did not complete \<ISCED level 1\> |                                       |
:::

```{r}
#| eval: true
#| code-fold: true
#| code-summary: "Show the code"
#| echo: true
# Create contingency table of mother's level of education
# Create a data frame of mother's level of education

MatEd<-PISA_2022%>%
  select(CNT, ST005Q01JA)%>% # Select maternal ed and country variables
  filter(CNT == "United Kingdom"| CNT=="France"|CNT=="Germany")%>% 
          # Filter for CNT
  droplevels() # To prevent the levels for other countries confusing the table

# Turn data frame into a contingency table

ContTab<-xtabs(~ST005Q01JA + CNT, data = MatEd)
ContTab

# The country with the highest number of level 3A mothers is the UK
```

b)  `ST261Q04JA` asks if, in the last 3 months, students' transport difficulties stopped students getting to school. Create a contingency table by gender for this variable for students in the UK. Do more girls or boys have transport difficulties?

```{r}
#| eval: true
#| code-fold: true
#| code-summary: "Show the code"
#| echo: true
# Create contingency table of having transport difficulties
# First, create a data frame of having a desk in the UK

DeskUK <- PISA_2022 %>%
  select(CNT, ST261Q04JA, ST004D01T) %>% # Select transport & country variables
  filter(CNT == "United Kingdom") %>% # Filter for the UK
  droplevels() # To prevent the levels for other countries confusing the table

# Convert the data frame to a contingency table

ContTab <- xtabs(~ST261Q04JA + ST004D01T, data = DeskUK)

ContTab

# More boys have transport problems

```

c)  `ST250Q05JA` asks if students have access to the internet. In which country in the data frame do students report the highest levels of access to the internet?

```{r}
#| eval: false
#| echo: true
#| code-fold: true
#| code-summary: "Show the code"

# The country with respondents reporting most internet access is Spain

PISA_2022 %>%
  select(CNT, ST250Q05JA) %>% # Select internet and country variables
  filter(ST250Q05JA == "Yes") %>% 
  group_by(CNT) %>%
  count() %>%
  ungroup() %>%
  filter(n == max(n))

# alternatively, you could use:
#  arrange(desc(n))

```
:::

::: callout-tip
To sort a table, the easiest way is to convert it to a data frame and then use the `arrange` function. The default order for arrange is ascending, adding `desc` switches to descending.

```{r}
#| eval: true
#| code-fold: false
#| echo: true

# Arranging a table

PISA_2022 %>%
  select(CNT, ST250Q05JA) %>% # Select internet and country variables
  filter(ST250Q05JA == "Yes") %>% 
  group_by(CNT) %>%
  count() %>%
  arrange(desc(n))

```
:::

## Task 2 - Goodness of fit test. Are the responses of the survey in proportion to populations?

::: question
The populations of three countries in the sample are:

| Country | Population  | Ratio |
|---------|-------------|-------|
| US      | 332 million | 0.69  |
| Germany | 83 million  | 0.17  |
| UK      | 67 million  | 0.14  |

Are the number of responses in the sample a good fit for the overall populations?

Use a goodness of fit implementation of Chi square with the null hypothesis that the proportion of students in the sample of PISA match that of the overall population.

Find a Cramer's V effect size for the test.

**Hint**: As we only have one variable here, numbers in each country, we don't use `xtabs` to count - instead we can use `group_by(CNT)` and `count()` (don't forget to use `drop_levels()` . For example, to create counts of the number of entries in the PISA data for France, Spain and Italy, you can use the code below:

```{r}
#| echo: true

# Create a data frame of the entries for Spain, France and Italy

PISAFRSPIT<-PISA_2022%>%
  select(CNT)%>%
  filter(CNT == "France" | CNT== "Spain" | CNT == "Italy")%>%
  group_by(CNT)%>%
  count()%>%
  droplevels()

PISAFRSPIT
```
:::

```{r}
#| eval: false
#| code-fold: true
#| code-summary: "Show the code"
#| echo: true

# Create a data frame of entries in the data for the UK, US and Germany

Subset <- PISA_2022 %>%
  select(CNT) %>%
  filter(CNT == "United Kingdom"|CNT == "United States"|CNT == "Germany") %>%
  group_by(CNT)%>%
  count()%>%
  droplevels()

# perfom the test 
chisq.test(Subset$n, p = c(0.17, 0.14, 0.69))

# P is < 2.2e-16 indicating that
# The UK has many more responses that expected by proportion of its size.

# Find the effect size using Cramer's V
library(lsr)
cramersV(Subset$n, p = c(0.17, 0.14, 0.69))
# Cramer's V = 0.54, a large

```

## Task 3 - Goodness of fit test: Birth month distribution

::: question
Perform a goodness of fit test to determine if the birth months (`ST003D02T`) of respondents are distributed as expected in

a)  the whole sample;
b)  in the UK. Use ggplot to plot a column graph of both data sets (the world and the UK).
c)  What might explain any patterns you see.

Hint - you can use the `rep` function to save you typing. For example, if you want to set a variable to be 0.1 ten times, you use `variable<-c(rep(0.1, times=10))`
:::

```{r}
#| eval: false
#| code-fold: true
#| code-summary: "Show the code"
#| echo: true

# Create a data frame of counts of birth months
Worldmonth <- PISA_2022 %>%
  select(ST003D02T) %>%
  group_by(ST003D02T) %>%
  droplevels() %>%
  na.omit()

# Create a contingency table of birth months

WorldmonthFreqTable <- xtabs(data=Worldmonth, ~ ST003D02T)

# Create an expected variable - hint rep repeats a value a specified number of times - to reduce typing

Expected <- c(rep(1/12, times=12))

chisq.test(WorldmonthFreqTable, p = Expected)

# p-value =  2.2e-16, don't accept the null hypothesis.
# The world data does not follow the expected distribution or birth months

# find the effect size 
cramersV(WorldmonthFreqTable)
# there are 11 degrees of freedom
# Cramer's V = 0.014, not even a small effect size

UKmonth <- PISA_2022 %>%
  select(ST003D02T, CNT) %>%
  filter(CNT=="United Kingdom") %>%
  group_by(ST003D02T) %>%
  droplevels() %>%
  na.omit()

# Create a contingency table of birth months in the UK
UKmonthtable <- xtabs(data=UKmonth, ~ ST003D02T)

# Set the expected values
Expected <- c(rep(1/12, times=12))

# Perform the test

chisq.test(UKmonthtable, p = Expected)
# p-value = 0.0002224, don't accept the null hypothesis.
# The world data does not follow the expected distribution or birth months

# find the effect size
cramersV(UKmonthtable)
# there are 11 degrees of freedom
# Cramer's V = 0.015, not even a small effect size

ggplot(data = Worldmonth,
       aes(x=ST003D02T, y = n, fill = ST003D02T)) +
  geom_col() +
  ggtitle("World birth months")

ggplot(data = UKmonth,
       aes(x=ST003D02T,y = n, fill = ST003D02T)) +
  geom_col() +
  ggtitle("UK birth months")

```

## Task 4 - Country differences - hypothesis testing

::: question
Perform a hypothesis test to determine:

a)  `ST250Q01JA` asks students if they have a room of there own. Test the hypothesis that there is no difference in having a room between students in the UK, US, New Zealand and Australia.

Plot a mosaic plot of the proportions

```{r}
#| eval: false
#| code-fold: true
#| code-summary: "Show the code"
#| echo: true

# Create a data frame of having a room (ST250Q01JA) for the 4 countries

Room <- PISA_2022 %>%
  select(CNT, ST250Q01JA) %>% # choose country and room
  filter(CNT == "United Kingdom"|CNT == "United States"|
           CNT == "New Zealand"|CNT == "Australia") %>% # filter by four countries
  droplevels() # Remove other countries which exist as factors

# Create a contingency table
RoomCont <- xtabs(data=Room, ~ ST250Q01JA + CNT)

chisq.test(RoomCont) # Perform the test

# The p-value is less than 0.05 (p-value < 2.2e-16), therefore we reject the null hypothesis that the distributions of rooms are the same in the countries

# find the effect size
cramersV(RoomCont)
# the contingency table has 4 rows and 2 columns, so df = (4-1)(2-1) = 3
# Cramer's V = 0.06, not quite reaching the 0.07 threshold for a small effect size

# Create a geom mosaic plot of LANGN by CNT

ggplot(data = Room) +
  geom_mosaic(aes(x = product(ST250Q01JA, CNT), fill = ST250Q01JA)) +
  ylab("Number of rooms at home") +
  xlab("Country") +
  theme(axis.text.x = element_text(angle = 90, vjust = 0.5, hjust=1))# Rotate x-axis labels




```
:::

::: question
b)  Are there significant differences between Japan, Greece and the UK in `ST250Q02JA`, A computer (laptop, desktop, or tablet) that you can use for school work? (Yes or No). Produce a mosaic plot.

```{r}
#| eval: false
#| code-fold: true
#| code-summary: "Show the code"
#| echo: true

# Create a data frame of ST250Q02JA (computers at home) for the 3 countries of interest

Comp <- PISA_2022 %>%
  select(CNT, ST250Q02JA) %>% # choose country and use of computer
  filter(CNT %in% c("United Kingdom", "Japan", "Greece")) %>% 
  # filter by countries
  droplevels() # Remove other countries which exist as factors

# Produce a contingency table
ContComp <- xtabs(data=Comp, ~ CNT + ST250Q02JA)

chisq.test(ContComp) # Perform the test
# The p-value is less than 0.05 (p-value < 2.2e-16), therefore we reject the null hypothesis that the distributions are the same

# find the effect size
cramersV(ContComp)
# the contingency table has 3 rows and 2 columns, so df = (3-1)(2-1) = 2
# Cramer's V = 0.127, a small effect size

# Produce a mosaic plot by CNT

ggplot(data = Comp)+
  geom_mosaic(aes(x = product(ST250Q02JA, CNT), fill = ST250Q02JA))+
  ylab("Number of computers at home")+
  xlab("Country")+
  theme(axis.text.x = element_text(angle = 90, vjust = 0.5, hjust=1))# Rotate x-axis labels



```
:::

::: question
Perform a hypothesis test to determine if `ST007Q01JA` - the highest level of schooling completed by respondents' fathers, is different in the UK, US, France and Germany.

Plot a mosaic plot of the proportions.

Follow up question: Are the proportions of paternal education different in the three European countries (UK, France and Germany)?

::: callout-tip
Hint: assume the null hypothesis is that fathers have the same level of qualifications in the four countries.
:::

```{r}
#| eval: false
#| code-fold: true
#| code-summary: "Show the code"
#| echo: true

# Create a data frame of paternal education ST007Q01TA for the 3 countries

PatEdTypes<-PISA_2022 %>%
  select(CNT, ST007Q01JA) %>% # choose country and type of school
  filter(CNT =="United Kingdom"|CNT == "United States"|
           CNT== "France"|CNT == "Germany")%>% # filter by four countries
  droplevels() # Remove other countries which exist as factors

# Create a contingency table

Conttable <- xtabs(data=PatEdTypes, ~ CNT + ST007Q01JA)

# Perform the chi.sq test

chisq.test(Conttable) # Perform the test
# The p-value is less than 0.05 (p-value < 2.2e-16), 
# therefore we reject the null hypothesis that the distributions are the same

# find the effect size
cramersV(Conttable)
# the contingency table has 1 rows and 4 columns, so df = (4-1)(5-1) = 12
# Cramer's V = 0.35, a large effect size

# Create a mosaic plot

ggplot(data = PatEdTypes)+
  geom_mosaic(aes(x = product(ST007Q01JA, CNT), fill = ST007Q01JA))+
  ylab("Paternal Education")+
  xlab("Country")+
  theme(axis.text.x = element_text(angle = 90, vjust = 0.5, hjust=1))# Rotate x-axis labels



# Follow up Question
# Create a data frame of paternal education ST007Q01JA for the 3 countries
PatEdTypes<-PISA_2022%>%
  select(CNT, ST007Q01JA)%>% # choose country and type of school
  filter(CNT == "United Kingdom" | CNT == "France" | CNT == "Germany")%>% 
  droplevels() # Remove other countries which exist as factors

Conttable<-xtabs(data = PatEdTypes, ~CNT + ST007Q01JA)
chisq.test(Conttable) # Perform the test
# The p-value is less than 0.05 (p-value < 2.2e-16), 
# therefore we reject the null hypothesis that the distributions are the same

# find the effect size
cramersV(Conttable)
# the contingency table has 1 rows and 3 columns, so df = (3-1)(5-1) = 8
# Cramer's V = 0.39, a large effect size

# Create a geom_mosaic

ggplot(data = PatEdTypes)+
  geom_mosaic(aes(x = product(ST007Q01JA, CNT), fill = ST007Q01JA))+
  ylab("Paternal Education")+
  xlab("Country")+
  theme(axis.text.x = element_text(angle = 90, vjust = 0.5, hjust=1))# Rotate x-axis labels



```
:::

# Testing ordinal data with the Kruskal-Wallis test

The chi square test assumes the data is categorical - that is, there is no defined order to the data (for example, type of school (vocational, technical, general)). Ordinal data have a defined rank - for example, categories of the number of books in a home (e.g. 0-5, 6-10, 11-15, etc.) have a numerical order.

When testing hypotheses related to ordinal data, an alternative test to the chi square test is used, the Kruskal Wallis test. This works in the same way as the chi square tests, but uses a data frame that includes ordinal data.

For example, item `ST254Q03JA` asks how many laptops participants have in their homes. The data here will will be ordinal (the response values are: `1`,`2`,`3`,`4`,`5`). Therefore, rather than performing a Chi Square test we use the `kruskal.test` function. To perform a `kruskal.test` you set the data frame you want to test `kruskal.test(data=UKUSphones, ST254Q03JA ~ CNT)` in our case, `UKUSphoness` and indicate the variables you want to compare `ST254Q03JA ~ CNT` in our case difference in the possession of phones `ST254Q03JA` by `~` country `CNT`.

There is no need to create a contingency table for the Kruskal Wallis test

```{r}
#| echo: true
# Create a data frame for US and UK phones
UKUSphones <- PISA_2022 %>%
  select(CNT, ST254Q03JA) %>% # choose country and no of phones
  filter(CNT == "United Kingdom" | CNT == "United States") %>% 
  # filter for the UK and US
  droplevels() # Remove other countries which exist as factors


kruskal.test(data=UKUSphones, ST254Q03JA ~ CNT) # Perform the test
# The p-value is less than 0.05 (p-value= 2.2e-16), therefore we reject the null hypothesis that the number of phones is the same for respondents in the UK and the US

# find the effect size
library(rstatix)
kruskal_effsize(UKUSphones, ST254Q03JA ~ CNT)
# eta squared = 0.0056, a small effect size

```

The test is interpreted in the same way as a chi square test. The null hypothesis is that there are no differences in the ordinal data of the groups being tested. If the p-value is over 0.05 the null hypothesis is accepted, if it is below, the null hypothesis is rejected.

## Task 5 - Kruskal Wallis Tests

::: question
Are there differences in the number of televisions in the home `ST254Q01JA` in the UK and the United States? Create a mosaic plot

```{r}
#| eval: true
#| code-fold: true
#| code-summary: "Show the code"
#| echo: true

UKUSTV<-PISA_2022 %>%
  select(CNT, ST254Q01JA ) %>% # choose country and no of TVs
  filter(CNT == "United Kingdom"| CNT == "United States") %>% 
  # filter for the UK and US
  droplevels() # Remove other countries which exist as factors

# Perform the test

kruskal.test(data = UKUSTV,ST254Q01JA ~ CNT) 

# The p-value is less than 0.05 (p-value=2.2e-16), therefore we reject the null hypothesis that the number of books is the same for respondents in the UK and the US

# find the effect size

kruskal_effsize(UKUSTV, ST254Q01JA ~ CNT)
# eta squared = 0.0006, a small effect size

# Create a geom_mosaic

ggplot(data = UKUSTV)+
  geom_mosaic(aes(x = product(ST254Q01JA, CNT), fill = ST254Q01JA))+
  ylab("Number of TVs by country")+
  xlab("Country")+
  theme(axis.text.x = element_text(angle = 90, vjust = 0.5, hjust=1))# Rotate x-axis labels



```

In the UK, are there differences for boys and girls for the number of instruments in the home (`ST251Q06JA`)? Create a mosaic plot.

```{r}
#| eval: true
#| code-fold: true
#| code-summary: "Show the code"
#| warning: false
#| echo: true

# Create a data frame of instruments in the UK including gender

UKInstruments<-PISA_2022%>%
  select(CNT, ST251Q06JA, ST004D01T ) %>% 
  # choose CNT, no of instruments, gender
  filter(CNT == "United Kingdom") %>% # filter for the UK
  select(ST251Q06JA, ST004D01T) %>% 
  #drop the country variable now filtering is done
  droplevels() # Remove other countries which exist as factors


# Perform the test

kruskal.test(data = UKInstruments, ST251Q06JA ~ ST004D01T) 
# The p-value is less than 0.05 (p-value = 6.146e-07), therefore we reject the null hypothesis. Girls and boys have different access to instruments in the home.

# find the effect size

kruskal_effsize(UKInstruments, ST251Q06JA ~ ST004D01T)
# eta squared = 0.0011, a small effect size


# Create a geom_mosaic

ggplot(data = UKInstruments)+
  geom_mosaic(aes(x = product(ST251Q06JA, ST004D01T), fill = ST251Q06JA))+
  ylab("Number of instruments in the home for UK young people")+
  xlab("Gender")+
  theme(axis.text.x = element_text(angle = 90, vjust = 0.5, hjust=1))# Rotate x-axis labels
```

In the UK, are there differences for boys and girls working for pay in the UK (`WORKPAY`)? Plot the data as a mosaic plot.

```{r}
#| eval: true
#| code-fold: true
#| code-summary: "Show the code"
#| warning: false
#| echo: true

# Create a data frame of working in the Uk with gender

UKWork<-PISA_2022 %>%
  select(CNT, WORKPAY, ST004D01T) %>% # choose CNT, WORKPAY, gender
  filter(CNT == "United Kingdom") %>% # filter for the UK
  select(WORKPAY, ST004D01T)%>% 
  #drop the country variable now filtering is done
  droplevels() # Remove other countries which exist as factors

# Perform the test

kruskal.test(UKWork, WORKPAY ~ ST004D01T) # Perform the test

# The p-value is less than 0.05 (p-value < 2.2e-16), therefore we reject the null hypothesis. Girls and boys have different levels of work

# find the effect size
kruskal_effsize(UKWork, WORKPAY ~ ST004D01T)
# eta squared = 0.0003, a small effect size

# Create a geom_mosaic

ggplot(data = UKWork)+
  geom_mosaic(aes(x = product(WORKPAY, ST004D01T), fill = WORKPAY))+
  ylab("Frequency of work")+
  xlab("Gender")+
  theme(axis.text.x = element_text(angle = 90, vjust = 0.5, hjust= 1))+
  theme(axis.text.y = element_text(size=rel(0.5))) # Reduce y-axis font size
  
# Rotate x-axis labels
```
:::


## Useful resources

# Doing Chi-Square tests in R

<iframe width="560" height="315" src="https://www.youtube.com/embed/qnZX0G1M29E" title="YouTube video player" frameborder="0" allow="accelerometer; autoplay; clipboard-write; encrypted-media; gyroscope; picture-in-picture" allowfullscreen>

</iframe>

You can find the code used in the video below

```{r}
#| eval: false
#| warning: true
#| echo: true

# Introduction to Chi-square
#
# Download data from /Users/k1765032/Library/CloudStorage/GoogleDrive-richardandrewbrock@gmail.com/.shortcut-targets-by-id/1c3CkaEBOICzepArDfjQUP34W2BYhFjM4/PISR/Data/PISA/subset/Students_2018_RBDP_none_levels.rds
# You want the file: Students_2018_RBDP_none_levels.rds
# and place in your own file system
# change loc to load the data directly. Loading into R might take a few minutes

loc <- "https://drive.google.com/open?id=14pL2Bz677Kk5_nn9BTmEuuUGY9S09bDb&authuser=richardandrewbrock%40gmail.com&usp=drive_fs"
PISA_2018 <- read_rds(loc)

# Are there differences between how often students change school?
# ST004D01T is the gender variable (Male, Female)
# SCCHANGE is a categorical variable (No change / One change / Two or more changes)

chidata <- PISA_2018 %>%
  select(CNT,ST004D01T,SCCHANGE) %>%
  filter(CNT=="United Kingdom")

chidata<-chidata[-c(1)]
chidata<-drop_na(chidata)

 chidata <- PISA_2018 %>%
   filter(CNT=="United Kingdom")
   select(ST004D01T,SCCHANGE) %>% 
   drop_na()
# Above is the approiach I took in the video
# An alternative, Pete suggests, which is more elegant, is below
# Note he drops the country varibale, within the piped section
# using: elect(-CNT)
#    
# chidata <- PISA_2018 %>%
#   select(CNT,ST004D01T,SCCHANGE) %>%
#   filter(CNT=="United Kingdom") %>%
#   select(-CNT) %>% 
#   drop_na()

# run the test
chisq.test(chidata$ST004D01T, chidata$SCCHANGE)
  

```
