---
title: "Self Study Tasks"
---

```{r setup}
    options(max.print=15)

    library(here)
    source(here::here("MASTEMR_page_setup.R"))
```

The pages below set out a series of graded challenges that you can use
to test your R and statistical skills. Sample code that solves each
problem is included so you can compare your solution with ours. Don't
worry if you solve something in a different way, there will be multiple
solutions to the same task. The tasks are all set on the PISA_2022 data
set:
[PISA_2022](/Users/k1765032/Library/CloudStorage/GoogleDrive-richardandrewbrock@gmail.com/.shortcut-targets-by-id/1c3CkaEBOICzepArDfjQUP34W2BYhFjM4/PISR/Data/PISA/2022/PISA_student_2022.parquet)

To load the data, use the code below:

```{r}
#| eval: false
#| warning: false
install.packages("arrow") # if you haven't already
library(arrow)
library(tidyverse)
PISA_2022 <- read_parquet("<location of the downloaded PISA_2022.parquet file>")
```

## Task 1 Practice creating a summary table #1

::: question
Create a table that summarises the mean PISA science scores by country.
You will need to use the `group_by`, `summarise` and `mean` functions.

```{r}
#| echo: true
#| eval: true
#| code-fold: true
#| code-summary: "Show the code"
#| warning: false

PISAsummary <- PISA_2022 %>%  # Pipe the overall frame to a summary data.frame
  select(CNT, PV1SCIE) %>%  # Select the two required columns
  group_by(CNT) %>%        # Group the entries by country
  summarise(meansci = mean(PV1SCIE)) # calculate means for each country

print(PISAsummary)

```

**Extension**: use the `signif` function to give the responses to three
significant figures
:::

## Task 2 Practice creating a summary table (including percentages) #2

::: question
Use the table function to create a summary of numbers of speakers of
different languages (LANGN) recorded in the data frame for the UK. Use
the mutate function to turn these into percentages (you will need to
calculate a total)

```{r}
#| echo: true
#| eval: true
#| code-fold: true
#| code-summary: "Show the code"
#| warning: false

UKPISA<-PISA_2022%>%
  select(CNT,LANGN)%>%               # Select the country school type 
  filter(CNT == "United Kingdom")%>%  # filter for the UK
  select(LANGN) %>%                  # Just select the language (removing country)
  droplevels()                                   

UKPISA<-xtabs(data=UKPISA, ~ LANGN)  # Create a summary of counts
                                    # To manipulate the table it is
UKPISA<-as.data.frame(UKPISA)       # easier to convert it to a 
                                    # a data.frame

UKPISA<-mutate(UKPISA, per = Freq / sum(Freq)*100)
UKPISA  
# If you want to sort the data (arrange descending by the percentage vector)
UKPISA<-UKPISA%>%
  arrange(desc(per))
UKPISA
```
:::

## Task 3 Practice pivoting a table

::: question
Convert a table of UK Science, Maths and Reading scores, extracted from
the main data set, into the long format R prefers. In the long format,
each score becomes a single so each student will have three entries.

```{r}
#| echo: true
#| eval: true
#| code-fold: true
#| code-summary: "Show the code"
#| warning: false
# Create a data frame in wide format, with three columns for each student's scores (math, reading and science)
UKScores<-PISA_2022%>%
  select(CNT,PV1MATH, PV1READ, PV1SCIE)%>%
  filter(CNT == "United Kingdom")%>%
  select(PV1MATH, PV1READ, PV1SCIE)
# Use pivot longer to turn the three columns into one. First, pass pivotlonger the dataframe to be converted, then the three columns
# to convert into one, the name of the new longer column and the
# name of the new scores column

UKScores<-pivot_longer(UKScores, cols = c('PV1MATH', 'PV1READ', 'PV1SCIE'),
                       names_to = 'Subject', values_to = 'Score' )

```
:::

## Task 4 Graphing Practice #1 A Bar Chart

::: question
Draw a bar chart of the mean mathematics scores for Germany, the UK, the
US and China

```{r}
#| echo: true
#| eval: true
#| code-fold: true
#| code-summary: "Show the code"
#| warning: false

Plotdata<-PISA_2022%>%
  select(CNT, PV1MATH)%>%
  filter(CNT == "United Kingdom"| CNT == "United States"|
           CNT == "Germany"| CNT == "B-S-J-Z (China)")%>%
  group_by(CNT)%>%
  summarise(mean = mean(PV1MATH))

ggplot(Plotdata,               # Pass the data to be plotted to ggplot
       aes(x = CNT, y = mean))+    # set the x and y varibale
  geom_col(fill = "red")         # plot a column graph and fill in red

```
:::

## Task 5 Graphing Practice #2 A Bar Chart with two series

::: question
Draw a bar chart of the mean mathematics scores for Germany, the UK, the
US and Korea for boys and girls

```{r}
#| echo: true
#| eval: true
#| code-fold: true
#| code-summary: "Show the code"
#| warning: false

Plotdata<-PISA_2022%>%
  select(CNT, PV1MATH, ST004D01T)%>%
  filter(CNT == "United Kingdom"|CNT=="United States"|
           CNT == "Germany"|CNT == "Korea")%>%
  group_by(CNT, ST004D01T)%>%
  summarise(mean = mean(PV1MATH))

ggplot(Plotdata,
       aes(x = CNT, y=mean, fill = ST004D01T))+ # Setting the fill to the gender
                                            # variable gives two series
  geom_col(position = position_dodge())     # position_dodge here means the
                                            # means the bars are plotted                                                # side by side

```
:::

## Task 6 Graphing Practice #3 A scatter plot

::: question
Plot a graph of science scores against mathematics scores for students
in the UK

```{r}
#| echo: true
#| eval: true
#| code-fold: true
#| code-summary: "Show the code"
#| warning: false

Plotdata<-PISA_2022%>%              # Create a new data frame to be plotted
  select(CNT, PV1MATH, PV1SCIE)%>%  # Choose the country, and scores vectors
  filter(CNT == "United Kingdom")    # Filter for only Uk results

ggplot(Plotdata,                  # Pass the data to be plotted to ggplot
       aes(x = PV1MATH, y = PV1SCIE))+ # Define the x and y variable
      geom_point(size = 0.1, alpha = 0.2, colour="red")+ 
                                  # Use geom-point to create a scatter                                      # graph and set the size of the point 
                                    # alpha (i.e transparency)
      labs(x = "Math Score", y = "Science score") # Add clearer labels
```
:::

## Task 7 Graphing Practice #4 A scatter plot with multiple series

::: question
Plot a graph of science scores against mathematics scores for students
in the UK, with data split into two series for boys and girls

```{r}
#| echo: true
#| eval: true
#| code-fold: true
#| code-summary: "Show the code"
#| warning: false

Plotdata<-PISA_2022%>%              # Create a new dataframe to be plotted
  select(CNT, PV1MATH, PV1SCIE, ST004D01T)%>%  
  filter(CNT == "United Kingdom")    # Filter for only Uk results

ggplot(Plotdata,                  
       aes(x = PV1MATH, y = PV1SCIE, colour = ST004D01T))+ 
      geom_point(size = 0.1, alpha = 0.2)+ 
                          # As above, but set colour by the gender varibale
      labs(x = "Math Score", y = "Science score") 

```
:::

## Task 8 Graphing Practice #4 A scatter plot with varying size points

::: question
Plot a graph of mean science scores against mean mathematics scores for
all the countries in the data set. Vary the point size by the number of
students per country.

```{r}
#| echo: true
#| eval: true
#| code-fold: true
#| code-summary: "Show the code"
#| warning: false

Plotdata<-PISA_2022%>%
  select(CNT, PV1MATH, PV1SCIE) %>%
  group_by(CNT) %>%
  summarise(meansci = mean(PV1SCIE), meanmath=mean(PV1MATH), total=n())

  # Summarise finds mean scores by countries and n() is used to sum
  # the number of students in each country

ggplot(Plotdata,
       aes(x = meansci, y = meanmath, size = total, colour = "red"))+
  # The size aesthetic is set to the total entries value computed
  # for the data set
  geom_point()+
  labs(x = "Mean science score", y = "Mean math score")


```
:::

## Task 9 Graphing Practice #5 A mosaic plot

::: question
Plot a mosaic plot of the number of students who speak (use `LANGN`)
French and Spanish in the whole data set

```{r}
#| echo: true
#| eval: true
#| code-fold: true
#| code-summary: "Show the code"
#| warning: false

Lang<-PISA_2022 %>%
  select(ST004D01T, LANGN) %>%
  filter(LANGN == "French" | LANGN == "Spanish") %>%
  na.omit() %>%
  droplevels()

library(ggmosaic)
ggplot(Lang)+
  geom_mosaic(aes(x = product(ST004D01T, LANGN), fill = LANGN))
         
```
:::

## Task 10 T-test practice #1

::: question
Using the PISA 2022 data set, determine if there are statistically
significant differences between the science, reading and mathematics
scores of the UK and the US.

```{r}
#| echo: true
#| eval: true
#| code-fold: true
#| code-summary: "Show the code"
#| warning: false
# Create data frames with the score results for UK and US
UKscores<-PISA_2022%>%
  select(CNT,PV1MATH,PV1READ, PV1SCIE)%>%
  filter(CNT == "United Kingdom")

USscores<-PISA_2022%>%
  select(CNT,PV1MATH,PV1READ, PV1SCIE)%>%
  filter(CNT == "United States")

# Perform the t-test with maths results

t.test(UKscores$PV1MATH, USscores$PV1MATH)
# p-value is < 2.2e-16 so significant differences exist for maths

t.test(UKscores$PV1READ, USscores$PV1READ)
# p-value = 1.339e-10 - statistically significant differences exist for reading between Uk and US

t.test(UKscores$PV1SCIE, USscores$PV1SCIE)
# p-value = 0.00119  significant differences exist for science between the UK and US

```
:::

## Task 11 T-test practice #2

::: question
Divide the UK population into two groups, those that have internet
access at home (`ST250Q05JA`) and those who do not. Are there
statistically significant differences in the means of their reading,
science and mathematics scores?

```{r}
#| echo: true
#| eval: true
#| code-fold: true
#| code-summary: "Show the code"
#| warning: false
# Create data frames with the score results for UK in two
# groups, has internet and no internet, based on ST011Q06TA

UKHasIntscores<-PISA_2022%>%
  select(CNT,PV1MATH,PV1READ, PV1SCIE, ST250Q05JA)%>%
  filter(CNT=="United Kingdom" & ST250Q05JA == "Yes")

UKNoIntscores<-PISA_2022%>%
  select(CNT,PV1MATH,PV1READ, PV1SCIE, ST250Q05JA)%>%
  filter(CNT=="United Kingdom" & ST250Q05JA == "No")

# Perform the t-test with maths results
t.test(UKHasIntscores$PV1MATH, UKNoIntscores$PV1MATH)

# p-value is < 2.2e-16 so no significant differences for maths scores from

# those with and without internet

t.test(UKHasIntscores$PV1READ, UKNoIntscores$PV1READ)

# p-value = 2.547e-16 so no signficant differences for reading scores from

# those with and without internet

t.test(UKHasIntscores$PV1SCIE, UKNoIntscores$PV1SCIE)

# p-value = 7.135e-15 so no signficant differences for science scores from
# those with and without internet
```
:::

## Task 12 T-test practice #3

::: question
Using the PISA 2022 data set, are the mean mathematics scores of US boys
and girls different to a statistically significant degree?

```{r}
#| echo: true
#| eval: true
#| code-fold: true
#| code-summary: "Show the code"
#| warning: false
# Create a data frame of US boys math scores

USboys <- PISA_2022 %>%
  select(CNT, PV1MATH, ST004D01T)%>%
  filter(CNT == "United States")

# Create a dataframe of US girls math scores

USgirls <- PISA_2022 %>%
  select(CNT, PV1MATH, ST004D01T) %>%
  filter(CNT == "United States")

# Perform the t-test, using $PVMATH to indicate which column of the data frame to use

t.test(USboys$PV1MATH, USgirls$PV1MATH)

# The p-value is 1 which is over 0.05 suggesting we accept the null hypothesis, there are no  statistically significant difference in US girls and boys math scores
  
```
:::

## Task 13 T-test practice #3

::: question
Are the mean science scores of all students in the US and the UK
different to a statistically significant degree?

```{r}
#| echo: true
#| eval: true
#| code-fold: true
#| code-summary: "Show the code"
#| warning: false

# Create a data frame of US science scores

USSci<-PISA_2022 %>%
  select(CNT, PV1SCIE)%>%
  filter(CNT == "United States")

# Create a data frame of UK science scores

UKSci<-PISA_2022 %>%
  select(CNT, PV1SCIE)%>%
  filter(CNT == "United Kingdom")

# Perform the t-test, using $PV1SCIE to indicate which column of the dataframe to use

t.test(USSci$PV1SCIE, UKSci$PV1SCIE)

# The p-value is 0.00119, less than 0.05, so we reject the null hypothesis, there are statistically significant differences between US and UK science scores  
```
:::

## Task 14 Kruskal Wallis practice #1

::: question
Are there statistically significant differences in the proportion of
boys and girls who Working in household/take care of family members
before or after school (`WORKHOME`) for the whole dataset? Note the
responses are:

-   No work in household or care of family members
-   1 time of working in household or caring for family members per week
-   2 times of working in household or caring for family members per
    week
-   3 times of working in household or caring for family members per
    week
-   4 times of working in household or caring for family members per
    week
-   5 times of working in household or caring for family members per
    week
-   6 times of working in household or caring for family members per
    week
-   7 times of working in household or caring for family members per
    week
-   8 times of working in household or caring for family members per
    week
-   9 times of working in household or caring for family members per
    week
-   10 or more times of working in household or caring for family
    members per week

```{r}
#| echo: true
#| eval: false
#| code-fold: true
#| code-summary: "Show the code"
#| warning: false
# Create a data frame of including gender and working or caring

workcare <- PISA_2022 %>%
  select(WORKHOME, ST004D01T) %>%
  filter(!is.na(WORKHOME))

# As the data are ordinal, use a kurskal wallis test

kruskal.test(data=workcare, WORKHOME ~ ST004D01T )

# p-value < 2.2e-16 so there are statistically significant differences between genders

# plot the results

workcare<-workcare %>%
  droplevels()%>%
  na.omit()

ggplot(data = workcare)+
   geom_mosaic(aes(x=product(WORKHOME, ST004D01T), fill=WORKHOME))+
  scale_y_discrete(label=abbreviate)

```
:::

## Task 15 Chi-square practice #1

::: question
Are there statistically significant differences, in the US, in the
languages spoken (`LANGN`) by boys and girls?

```{r}
#| echo: true
#| eval: true
#| code-fold: true
#| code-summary: "Show the code"
#| warning: false
# Create a data frame of languages spoken in the US, including gender

USLang <- PISA_2022 %>%
  filter(CNT == "United States") %>%
  select(LANGN, ST004D01T) %>%
  na.omit() %>%
  droplevels()

# Create a contingency table

Contab <- xtabs(data=USLang, ~ LANGN + ST004D01T)

# Run the chi.sq test

chisq.test(Contab)

# The output p-value is 0.04665 which is less than 0.05. So reject the null hypothesis. There is a difference in language by gender
```
:::

## Task 16 Chi-square practice #3

::: question
Are there statistically significant differences in numbers of students
missing school for more than 3 months because they were bored
(`ST261Q01JA`) between the UK and US

```{r}
#| echo: true
#| eval: true
#| code-fold: true
#| code-summary: "Show the code"
#| warning: false
# ST261Q01JA - Why miss school for 3+ months: I was bored.
# Create a data frame for the two countries

Bored <- PISA_2022 %>%
  select(CNT, ST261Q01JA) %>%
  filter(CNT == "United Kingdom" | CNT == "United States") %>%
  droplevels() %>%
  na.omit()

# Create a contingency table

Contab <- xtabs(data=Bored, ~ CNT + ST261Q01JA)

# Do the chi squared test

chisq.test(Contab)

# p-value is less than 0.05 (2.537e-07), so reject the null hypotheses - there are statistically significant differences in boredom in the UK and the US

```
:::

## Task 18 Anova practice #1

::: question
Are there statistically significant differences in mathematics scores of students in
France, Germany, Spain, the UK and Italy? Find between which pairs of
countries there are statistically significant differences in mathematics scores.

```{r}
#| echo: true
#| eval: true
#| code-fold: true
#| code-summary: "Show the code"
#| warning: false

# Create a data frame of the required countries

EuroPISA <- PISA_2022 %>%
  select(CNT, PV1MATH)%>%
  filter(CNT %in% c("Spain", "France", "United Kingdom", "Italy", "Germany"))

# Perform the anova

resaov <- aov(data = EuroPISA, PV1MATH ~ CNT)
summary(resaov)

# Yes, statistically significant differences exist between the countries Pr(>F) <2e-16 ***
# Perform a Tukey HSD test

TukeyHSD(resaov)

# Significant differences p<0.05 exist for all countries except: Spain-Germany; Italy-Germany, UK-Spain.
```
:::

## Task 19 Anova practice #2

::: question
For the UK PISA 2022 data set, which variable out of `HOMEPOS`,
`ST004D01T`, `OCOD1` (Mother's occupation), `OCOD2` (Father's
occupation), `ST250Q05JA` (having a link to the internet), and highest
level of parental education (`HISCED`) accounts for the most variation
in science score? What percentage of variance is explained by each
variable?

! This is a big calculation so will take some time to compute !

```{r}
#| echo: true
#| eval: true
#| code-fold: true
#| code-summary: "Show the code"
#| warning: false

# Create a data frame for the UK
UKPISA_2022 <- PISA_2022 %>%
  filter(CNT == "United Kingdom")

# Perform the anova calculation with science score as the dependent variable

resaov <- aov(data=UKPISA_2022, 
              PV1SCIE ~ HOMEPOS + ST004D01T + OCOD1 + OCOD2 + ST250Q05JA + HISCED)

# Print the output
summary(resaov)

# Calculate the value of eta and multiple by a 100 to get the % of variance explained
eta <- etaSquared(resaov)
eta <- 100*eta
eta <- as.data.frame(eta)
eta
# The variable that explains most variation in science scores is father's occupation OCOD2 (7.7%), then home possession OCOD1 (5.9%), then wealth HOMEPOS (5.1%)
```
:::
