---
title: "Questions and Answers"
---

```{r setup}
    options(max.print=15)

    library(here)
    source(here::here("MASTEMR_page_setup.R"))
```

# Questions about R {#sec-QANDA}

## Why doesn't my `select`/`filter` statement work?

When you are loading packages, sometimes different packages have the same function names in them, and the functions themselves will do very different things. For example, there is a `select` function in the `tidyverse`, but also *another* `select` function in the package `MASS` that does something very different. If we load the tidyverse before loading MASS, then the MASS version of select is the one that will be used?!

```{r}
#| echo: false
#| eval: false
#| warning: false
# if("package:broom" %in% search()) detach("package:broom", unload=TRUE)
# if("package:tidyr" %in% search()) detach("package:tidyr", unload=TRUE)
# if("package:dbplyr" %in% search()) detach("package:dbplyr", unload=TRUE)
# if("package:gt" %in% search()) detach("package:gt", unload=TRUE)
# if("package:dplyr" %in% search()) detach("package:dplyr", unload=TRUE)
# 
# if("package:multcomp" %in% search()) detach("package:multcomp", unload=TRUE)
# if("package:TH.data" %in% search()) detach("package:TH.data", unload=TRUE)
# if("package:MASS" %in% search()) detach("package:MASS", unload=TRUE)
# if("package:see" %in% search()) detach("package:see", unload=TRUE)
# if("package:stats" %in% search()) detach("package:stats", unload=TRUE)
# remove all packages
while(!is.null(sessionInfo()$loadedOnly)){
  lapply(names(sessionInfo()$loadedOnly), require, character.only = TRUE)
  invisible(lapply(paste0('package:', names(sessionInfo()$otherPkgs)), detach, character.only=TRUE, unload=TRUE, force=TRUE))    
}

```

```{r}
#| error: true
#| eval: false
#| echo: true
library(tidyverse)
library(MASS)

diamonds %>% select(carat, cut, color)
```

To get around this make sure that you load the `tidyverse` *after* `MASS`, to be safe you should always load the `tidyverse` last.

```{r}
#| error: true
#| echo: true
library(MASS) 
library(tidyverse)

diamonds %>% select(carat, cut, color)
```

you can also specify the package that select comes from (in this case from a package within the `tidyverse` called `dplyr`):

```{r}
#| error: true
#| echo: true
diamonds %>% dplyr::select(carat, cut, color)
```

Finally, there is a package that helps your deal with conflicts called `conflicted`. At the top of your script you can define which version of a function you prefer, this will be the version that will always be used. If you want to specify another version of a function, you can still use other versions of the function using the `::` notation:

```{r}
#| eval: false
#| echo: true
library(conflicted)
conflicts_prefer(dplyr::select)
conflicts_prefer(dplyr::mutate)
conflicts_prefer(dplyr::summarise)
conflicts_prefer(dplyr::filter)

df %>% select(name, age) # uses select from dplyr
df %>% MASS::select(name) # uses select from MASS
```

To find out if you have any conflicts, you can run `conflict_scout()`.

## How can I unload packages

If you are finding yourself with a conflict as mentioned above and want to `unload` packages, then you need to run the following code:

```{r}
#| eval: false
#| echo: true
# adapted from: @mmfrgmpds https://stackoverflow.com/questions/7505547/detach-all-packages-while-working-in-r
while(!is.null(sessionInfo()$loadedOnly)){
  lapply(names(sessionInfo()$loadedOnly), require, character.only = TRUE)
  invisible(lapply(paste0('package:', names(sessionInfo()$otherPkgs)), detach, character.only=TRUE, unload=TRUE, force=TRUE))    
}
```

## Why am I getting the error 'could not find function "%>%"'

The pipe operator `%>%` is loaded when you load the tidyverse package - make sure you have installed tidyverse and loaded it

```{r}
#| eval: false
#| echo: true
install.packages("tidyverse")  # install
library(tidyverse)             # load
```

## I am getting the error: '"Error: Mapping should be created with `aes()` or `aes_()`."' when using ggplot

This may be caused by having a bracket after the geom rather than before it

```{r}
#| eval: false
#| echo: true
data<-data.frame(x=5:1,
                 y=10:6)
ggplot(data, aes(x, y) +           # Reproduce error message
         geom_point())

```

```{r}
#| eval: false
#| echo: true
data<-data.frame(x=5:1,
                 y=10:6)
ggplot(data, aes(x, y)) +           # Fixed error by moving bracket
         geom_point()

```

## My axis labels are too long

If you want to use long axis titles you may find they overrun the space available

```{r}
#| eval: true
#| echo: true

# Make a sample data frame
data<-data.frame(x=5:1,
                 y=10:6)

# plot the data
ggplot(data, aes(x=x, y=y)) +           
         geom_col()+
  ylab("A very long description for the y-axis label that will overflow and not look very nice")

```

To insert a line break in the axis label, add `\n` to the text where you want line breaks.

```{r}
#| eval: true
#| echo: true

# Make a sample data frame
data<-data.frame(x=5:1,
                 y=10:6)

# plot the data
ggplot(data, aes(x=x, y=y)) +           
         geom_col()+
  ylab("A very long description \n for the y-axis label that will \n overflow and not look very nice")

```

## How can I find out the full item labels in the PISA data frame?

The code below will give you a list of all the item labels

```{r}
#| eval: false
#| echo: true

# You may want to set the maximun print output to see all the labels
options(max.print = 1300)

lapply(PISA_2022, attr, "label")

```

## How do I turn a data frame into a form I can put into an assignment or paper?

You can use the `gt` package to convert data frames to aesthetically pleasing outputs.

```{r}
#| eval: true
#| echo: true

# load the gt package - you will need to run install.packages("gt") the first time
# install.packages("gt")
library(gt)

# Make the data frame you want you want to output
science_mean_scores <- PISA_2022 %>%
  group_by(CNT) %>%
  summarise(mean_science_score = mean(PV1SCIE, na.rm = TRUE),
            sd_science = sd(PV1SCIE, na.rm =TRUE))

# Use gt to produce an output - you can copy and paste the table from the
# viewer window to your report

gt(science_mean_scores)

# You can see more options on formating the table here: https://gt.rstudio.com
# For example to add a heading

gt(science_mean_scores) %>% tab_header(
    title = "Mean science scores",
    subtitle = "PISA 2022 data")

# Or to change the number of decimal places and a column name

gt(science_mean_scores) %>% tab_header(
  title = "Mean science scores",
  subtitle = "PISA 2022 data") %>%
  fmt_number(columns = c(mean_science_score , sd_science), decimals = 1) %>%
  cols_label("CNT" = md("**Country**"))

## If you have an output from a linear model, there is an additional step to do before using gt
# You need to use the broom package, which has the `tidy` function which gets the output of `lm`
# Into a suitable format for turing into a table
# install.packages("broom")

library(broom)

# Create example dataframe

UK_PISA <- PISA_2022 %>%
  select(PV1SCIE, PV1MATH, CNT) %>%
  filter(CNT == "United Kingdom")

# Run the model

uk_mod <- lm(data = UK_PISA, PV1SCIE ~ PV1MATH)

# tidy the model

uk_mod_tidy <- tidy(uk_mod)

# pass to gt to produce nice output

gt(uk_mod_tidy) %>% 
  fmt_number(columns=c("p.value"), 
              decimals = 3) %>%
  fmt_number(columns=c("estimate", "std.error", "statistic"), 
              decimals = 2)


```

## How do I report the results of tests in APA format?

When you have run some tests, the outputs (for example the degrees of freedom, test statistic and p-value) should be formatted in your paper in line with the citation convention of the journal or assessment (for example, KCL assignments follow the American Psychological Association (APA) citation style). You can find a general guide to presenting your results in APA style here: APA [numbers and statistics style guide](https://apastyle.apa.org/instructional-aids/numbers-statistics-guide.pdf).

For example, to report a chi-square result:

```{r}
#| eval: true
#| echo: true

# Create example dataframe

UK_gender_differences <- PISA_2022 %>%
  select(CNT, ST004D01T) %>%
  filter(CNT == "United Kingdom") %>% 
  droplevels()

cont_tab <- xtabs(data = UK_gender_differences, ~ ST004D01T)

# Run a chi sqaure goodness of fit test

chisq.test(cont_tab, p = c(0.5, 0.5))

```
A chi-square test of goodness-of-fit was conducted to compare the observed frequencies in the contingency table to the expected frequencies (equal numbers of boys and girls). The result was not statistically significant, χ²(1, N = 12973) = 2.44, p = .12, indicating no significant difference between the observed and expected frequencies.


To report a t-test of the mathematics scores of girls and boys in the UK:

```{r}
#| eval: true
#| echo: true

# Create example dataframe

UK_PISA <- PISA_2022 %>%
  select(PV1MATH, CNT, ST004D01T) %>%
  filter(CNT == "United Kingdom")

# Run a t-test

t.test(data = UK_PISA, PV1MATH ~ ST004D01T)

```
The output of that test would be reported as follows:

The results of a Welch's t-test indicated a statistically significant difference in mean mathematics scores between UK females (M = 475.61) and UK males (M = 489.29), t(12,942) = -8.22, p < .001, 95% CI [-16.95, -10.42].

Alternatively, consider the output of an anova to look at the difference in wealth scores between Finland, Norway, and Sweden:

```{r}
#| eval: true
#| echo: true

# Create example dataframe

wealth_scores <- PISA_2022 %>%
  select(HOMEPOS, CNT) %>%
  filter(CNT == "Finland" | CNT == "Norway" | CNT == "Sweden")

# Run an anova and summarise

aov_out <- aov(data = wealth_scores, HOMEPOS ~ CNT)
summary(aov_out)
TukeyHSD(aov_out)

```

A one-way ANOVA was conducted to compare levels of home possessions (HOMEPOS) across three countries (CNT: Finland, Norway, and Sweden). The analysis revealed a statistically significant effect of country on home possessions, F(2, 22,335) = 356.7, p < .001, η² = .031.

Post hoc Tukey's HSD tests indicated the following pairwise differences:

Norway had significantly higher HOMEPOS scores than Finland (M difference = 0.38, 95% CI [0.35, 0.42], p < .001).
Sweden had significantly higher HOMEPOS scores than Finland (M difference = 0.16, 95% CI [0.13, 0.20], p < .001).
Sweden had significantly lower HOMEPOS scores than Norway (M difference = -0.22, 95% CI [-0.26, -0.18], p < .001).
A total of 584 observations were excluded due to missing data.

Alternatively, you can use the `report` function in the `easystats` package to give a summary of an output.

```{r}
#| eval: true
#| echo: true

library(easystats)

# Get a summary of an anova

report(aov_out)

# And of a linear model

mod1 <- lm(data = wealth_scores, HOMEPOS ~ CNT)

report(mod1)

```

The `report` function will work on other test outputs, for example, on a t-test:

```{r}
#| eval: true
#| echo: true
#| warning: false

# Create example data frame

UK_PISA <- PISA_2022 %>%
  select(PV1MATH, CNT, ST004D01T) %>%
  filter(CNT == "United Kingdom") %>%
  droplevels()

# Use report on a t-test

report(t.test(data = UK_PISA, PV1MATH ~ ST004D01T))


```



To produce tables in APA format, you can use the `apa.aov.table` (for anovas) or `apa.reg.table` (for linear model outputs) functions from the `apaTables` package. You will need to install the package first using `install.packages("apaTables")`.

```{r}
#| eval: true
#| echo: true

library(apaTables)

# Get a summary table from an anova

apa.aov.table(aov_out)

# And of a linear model

apa.reg.table(mod1)

```
You can get more aesthetically pleasing versions of both tables using the `gt` function in the `gt` package. Before passing the outputs of `avo` of `lm` to `gt`, you will need to use the `tidy` fucntion from the `broom` package to tidy the output.

```{r}
#| eval: true
#| echo: true

library(gt)
library(broom)

# Tidy the outputs

tidied_aov <- tidy(aov_out)
tidied_mod1 <- tidy(mod1)


# Produced a nicely formatted table from an anova

gt(tidied_aov)

# And of a linear model

gt(tidied_mod1)

```

## How do I add the results from statistical tests to ggplot?

There are a number of helpful packages that can add data to charts. First `ggpubr` lets you add data about the regression line to your plot using the `stat_regline_equation()` function. You will need to tweak the coordinates (`label.x` and `label.y`) to appear appropiately on your chart.

```{r}
#| eval: true
#| echo: true

# Load the ggpbur library (you will need to run install.packages("ggpubr") the first time you use it)
library(ggpubr)

# Create example dataframe

UK_wealth_reading <- PISA_2022 %>%
  select(CNT, PV1READ, HOMEPOS) %>%
  filter(CNT == "United Kingdom") 

# Plot a scatter plot with the regression equation

ggplot(UK_wealth_reading, aes(x = HOMEPOS, y = PV1READ)) +
  geom_point(colour = "lightgreen", size = 0.1) +
  geom_smooth(method = "lm") +
  stat_regline_equation(label.x = -5, label.y = 600)  # Add regression equation


```

You can also use `stat_regline_equation` to add additional information like the p-value of the model and the R^2^ value:

```{r}
#| eval: true
#| echo: true

# Load the ggpbur library (you will need to run install.packages("ggpubr") the first time you use it)
library(ggpubr)

# Create example dataframe

UK_wealth_reading <- PISA_2022 %>%
  select(CNT, PV1READ, HOMEPOS) %>%
  filter(CNT == "United Kingdom") 

# Plot a scatter plot with the regression equation

ggplot(UK_wealth_reading, aes(x = HOMEPOS, y = PV1READ)) +
  geom_point(colour = "lightgreen", size = 0.1) +
  geom_smooth(method = "lm", formula = y ~ x) + # Ensure formula is specified here too
  stat_regline_equation(
    aes(label = paste(..eq.label.., ..adj.rr.label.., sep = "~~~~")), # Indicate you want the equation and R2 value and the seperator text
    label.x = -10, label.y = 600,
    formula = y ~ x)


```

Another powerful package for adding data to graphs is `ggstatsplot`. For example, it can anotate a plot with anova data. For example, if you want compare the science scores of the UK, Japan and the US:

```{r}
#| eval: true
#| echo: true

# Load the ggstatplot library (you will need to run install.packages("ggstatsplot") the first time you use it)
library(ggstatsplot)

# Create example dataframe

country_sci_data <- PISA_2022 %>%
  select(CNT, PV1SCIE) %>%
  filter(CNT == "United Kingdom" | CNT == "Japan" | CNT == "United States")

# Use ggstatplot to create an annotated plot with an anova result

ggbetweenstats(
  data  = country_sci_data,
  x     = CNT,
  y     = PV1SCIE,
  title = "Science scores in the UK, US and Japan",
  type  = "parametric" # Forces the test to be ANOVA
)



```

## How do I turn a data frame into a form I can put into an assignment or paper?

You can use the `gt` package to convert data frames to aesthetically pleasing outputs.

```{r}
#| eval: true
#| echo: true

# load the gt package - you will need to run install.packages("gt") the first time
# install.packages("gt")
library(gt)

# Make the data frame you want you want to output
science_mean_scores <- PISA_2022 %>%
  group_by(CNT) %>%
  summarise(mean_science_score = mean(PV1SCIE, na.rm = TRUE),
            sd_science = sd(PV1SCIE, na.rm =TRUE)) %>%
  filter(CNT == "United Kingdom" | CNT == "United States" | CNT == "Germany")

# Use gt to produce an output - you can copy and paste the table from the
# viewer window to your report

gt(science_mean_scores)

# You can see more options on formatting the table here: https://gt.rstudio.com
# For example to add a heading

gt(science_mean_scores) %>% tab_header(
    title = "Mean science scores",
    subtitle = "PISA 2022 data")

# Or to change the number of decimal places and a column name

gt(science_mean_scores) %>% tab_header(
  title = "Mean science scores",
  subtitle = "PISA 2022 data") %>%
  fmt_number(columns = c(mean_science_score , sd_science), decimals = 1) %>%
  cols_label("CNT" = md("**Country**")) # Here the **Country** makes Country bold
```

```{r}
#| eval: true
#| echo: true
## If you have an output from a linear model, there is an additional step to do before using gt
# You need to use the broom package, which has the `tidy` function which gets the output of `lm`
# Into a suitable format for turing into a table
# install.packages("broom")

library(broom)

# Create example dataframe

UK_PISA <- PISA_2022 %>%
  select(PV1SCIE, PV1MATH, CNT) %>%
  filter(CNT == "United Kingdom")

# Run the model

uk_mod <- lm(data = UK_PISA, PV1SCIE ~ PV1MATH)

# tidy the model

uk_mod_tidy <- tidy(uk_mod)

# pass to gt to produce nice output

gt(uk_mod_tidy)


```
## How can I change the colours of a plot if I don't want the default colours?

ggplot will use a default palette - you can sometimes identify plots in papers as being produced by ggplot from the distinctive green and pink colour! If you want to move away from these, you can use the `scale_fill_manual` and `scale_colour_manual` functions.

For example, if you have a bar graph of the number of boys and girls in the UK, you can change the fill colours using  `scale_fill_manual` and defining the colours for the categories (in this case `Male` and `Female`).

```{r}
#| eval: true
#| echo: true

UK_PISA <- PISA_2022 %>%
  select(ST004D01T, CNT) %>%
  filter(CNT == "United Kingdom")

ggplot(UK_PISA, aes(x = ST004D01T, fill = ST004D01T)) +
  geom_bar() +
  scale_fill_manual(values = c("Male" = "orange", "Female" = "lightblue"))  # manually set the values for male and female

```
You need to match the scale manual command to the aesthetic you want to change. For example, when using `geom_point` you use the `colour` aesthetic to change the colour of points. In that case, you use `scale_colour_manual`.

```{r}
#| eval: true
#| echo: true

UK_PISA <- PISA_2022 %>%
  select(ST004D01T, CNT, PV1SCIE, PV1MATH) %>%
  filter(CNT == "United Kingdom")

ggplot(UK_PISA, aes(x = PV1MATH, y = PV1SCIE, colour = ST004D01T)) +
  geom_point(size = 0.1, alpha = 0.6) +
  scale_colour_manual(values = c("Male" = "lightgreen", "Female" = "red")) # manually set the colour values for male and female

```


