---
title: "05 Introduction to PISA"
# bibliography: references.bib
echo: false
---

```{r setup}
#| eval: true
#| warning: false
#| echo: false

# only load this code if testing this file independently of overall project
if(!exists("test")){
  warning("RUNNING LOCAL FILE LOAD")
  
  library(glue)
  proj_folder <- "C:/Users/Peter/Google Drive/Kings/Code/MASTEMR/"
  proj_folder <- r"(/Users/k1765032/Google Drive/My Drive/MASTEMR/)"
  
  node <- Sys.info()
  if(node["nodename"] == "KCLJ7J6LFC74J"){ #Richard's PC
    proj_folder <- r"(/Users/k1765032/Google Drive/My Drive/MASTEMR/)"
  }else if(node["nodename"] == "KCL3Y7NKR3"){ #Pete's NEW Laptop
    proj_folder <- r"(G:\My Drive\Kings\Code\MASTEMR\)"
  }else if(node["nodename"] == "PETES_PC"){ #Pete's NEW Laptop
    proj_folder <- r"(C:\Users\Peter\Google Drive\Kings\Code\MASTEMR\)"
  }

  source(glue("{proj_folder}MASTEMR_setup.R"))
}
```

# Introduction to PISA

## Pre-session tasks

### Pre-reading

Please read through 2018 technical report: [PISA technical report](https://www.oecd.org/pisa/data/pisa2018technicalreport/PISA2018-TecReport-Ch-01-Programme-for-International-Student-Assessment-An-Overview.pdf)

### Getting set up

Remember to load the PISA 2018 data set

```{r}
#| eval: false
#| echo: true
library(arrow)
library(tidyverse)

PISA_2018 <- read_parquet(r"[<folder>PISA_2018_student_subset.parquet]")
PISA_2018Full <- read_parquet("/Users/k1765032/Library/CloudStorage/GoogleDrive-richardandrewbrock@gmail.com/.shortcut-targets-by-id/1c3CkaEBOICzepArDfjQUP34W2BYhFjM4/PISR/Data/PISA/2018/PISA_2018_student.parquet")
# Paste link from your own version of /PISR/Data/PISA/2018/PISA_2018_student.parquet


```

## The PISA assessments

Though it has been noted that interest, as measured by publications, interest in International large-scale assessments (ILSAs) peaked around 2012 [@jerrim2023peak], ILSAs remain a useful comparative tool for detecting patterns in educational achievement over time and across countries.

The OECD Programme for International Student Assessment is an international, survey of 15-year-old students [@OECD2018technical]. The survey includes items that measure achievement in mathematics, reading and science, various competencies (for example, digital skills and problem solving) and participants' perceptions of their school and home. In addition to the student survey which is the focus of this paper, teacher and parent surveys collect additional data.

The first PISA survey took place in 2000 (this round of data collection was followed by the inclusion of additional countries in 2001) [@OECD2018technical]. The assessment has taken place every three years (2000, 2003, 2006, 2009, 2012, 2015, 2018 and 2022). The 2022 survey was delayed due to the COVID pandemic [@OECD2021adjusted]. The number of participating countries has risen from 43, in 2000, to 690,000 in 2022. The number of participating students is shown in Table 1.

Table 1: Number of students completing PISA assessment by year

| Year | Number completing assessment |
|------|------------------------------|
| 2000 | 265,000                      |
| 2003 | 275,000                      |
| 2006 | 400,000                      |
| 2009 | 470,000                      |
| 2012 | 510,000                      |
| 2015 | 540,000                      |
| 2018 | 600,000                      |
| 2022 | 690,000                      |

The PISA survey uses a probabilistic, stratified clustered survey design [@jerrim2017weight].

To ensure that countries make equal contributions to regression models when they have different response rates, senate weights are applied to data [@jerrim2017weight].

A number of limitations of the PISA methodology have been noted. Sampling issues, including sample representatives, non-response rates and population coverage [@zieger2022conditioning; @rutkowski2016call; @gillis2016pisa; @hopmann2007pisa].

@anders2021canada and @jerrim2021pisa have shown that assumptions for imputing values for non-participating students used to construct the sample may have significant impacts on achievement scores.

Three sources of error are worth considering [@OECD2019results]. First, sampling error, uncertainty in the degree to which results from the sample generalise to wider populations - in 2018, the OECD average sampling error was 0.4 of a PISA point score. Second, measurement error, uncertainty in the extent to which test items measure proficiency. In 2018, the measurement error was around 0.8 of a point in mathematics and science and 0.5 of a score point in reading. Third, the link error is the uncertainty in comparison between scores in different years. For comparisons of Science scores between 2018 and 2015, the link error is 1.5 points.

From 2015, the majority of countries have used computer based test, following a switch from paper assessments [@jerrim2016pisa]. A randomised control trial of the two methods of delivery, conducted by the OECD prior to the switch indicated a difference in score between those taking paper and computer tests. The OECD introduced an adjustment to compensate for this effect, but the difference is not entirely removed by the correction [@jerrim2016pisa, @jerrim2018pisa]. However, Jerrim notes that 'in terms of cross-country rankings, there remains a high degree of consistency...the vast majority of countries are simply 'shifted' by a uniform amount' (pp. 508-509).

See also @feskens2019differential - change in scores due to changed mode. E.g. 'The reported decrease in mean science scores in 2015 was larger than the decrease in mathematics.4 This might be due to the introduction of new interactive items, which only took place for the science domain.' [@feskens2019differential, p. 245]

## A reminder about summarising data, graphing and categorising

### Summarising data

Recall you can use `group_by` and `summarise` to group individual student measures and find means and standard deviations for countries. For example, to find the mean wealth scores for the countries, and rank in descending order, we first select the variables of interest `CNT` and `WEALTH`, then `group_by` `CNT` and `summarise` to get the mean. As there are some NA values, we need to include `na.rm=TRUE` to tell `summarise` to ignore the missing values. Finally, we `arrange` in descending order by the new variable we create `meanwealth`. We can do the same and add a calculation to get the standard deviation.

```{r}
#| echo: true
PISA2018WealthRank <- PISA_2018 %>%
  select(CNT, WEALTH) %>% # Select variables of interest
  group_by(CNT) %>% # group by country
  summarise(meanwealth = mean(WEALTH, na.rm = TRUE)) %>% 
                    # summarise  country data to find the mean WEALTH
  arrange(desc(meanwealth)) # arrange in descending order based on the meanwealth score

# With standard deviations

PISA2018WealthRank <- PISA_2018 %>%
  select(CNT, WEALTH) %>% # Select variables of interest
  group_by(CNT) %>% # group by country
  summarise(meanwealth = mean(WEALTH, na.rm = TRUE),  
            # summarise  country data to find the mean WEALTH
            sdwealth=sd(WEALTH, na.rm = TRUE)) %>% # and standard deviation
  arrange(desc(meanwealth)) 
# arrange in descending order based on the meanwealth score

```

### Bar charts

Recall you can use `geom_bar` to plot a bar graph. For example, if we wanted to plot the `PISA2018WealthRank` dataframe we just created, we pass the data to `ggplot`. Recall that if you are passing `geom_bar` the exact values you want to plot, rather than making it count (for example, by including the original dataset with all student entries), you need to specify `geom_bar(stat='identity')`

I have added `+theme(axis.text.x = element_text(angle = 90, vjust = 0.5, hjust=1))` which rotates the text on the x-axis.

```{r}
#| echo: true
ggplot(PISA2018WealthRank, aes(x = CNT, y = meanwealth))+
  geom_bar(stat = 'identity')+
  theme(axis.text.x = element_text(angle = 90, vjust = 0.5, hjust = 1))
```

We can improve this plot by reordering the x-axis to rank the countries - we switch `x=CNT` to `x=reorder(CNT, -meanwealth)` that is we reorder the x axis based on descending (indicated by the minus sign `-meanwealth`) meanwealth.

```{r}
#| echo: true
ggplot(PISA2018WealthRank, aes(x=reorder(CNT, -meanwealth), y = meanwealth)) +
  geom_bar(stat='identity') +
  theme(axis.text.x = element_text(angle = 90, vjust = 0.5, hjust = 1))
```

If you like, you can add colour, tidy up the axis labels, and give a title:

```{r}
#| echo: true
ggplot(PISA2018WealthRank, aes(x = reorder(CNT, -meanwealth), y = meanwealth))+
  geom_bar(stat='identity', fill = "skyblue")+
  theme(axis.text.x = element_text(angle = 90, vjust = 0.5, hjust = 1))+
  ggtitle("Countries ranked by WEALTH")+
  xlab("Country")+
  ylab("Mean WEALTH")
```

### Scatter plots

To plot a scatter plot, recall we use `geom_point`. For example, to plot reading scores against mathematics scores in China (note China's CNT value: `B-S-J-Z (China)` which indicates the data ) we: a) create a data set of reading and science scores after filtering for China; b) pass the data to `ggplot`; c) use '`aes` to specify the x and y variables and d) plot with geom_point().

```{r}
#| echo: true
Chinaplot <- PISA_2018 %>%
  select(CNT, PV1READ, PV1SCIE) %>%
  filter(CNT == "B-S-J-Z (China)")

ggplot(Chinaplot, aes(x = PV1READ, y = PV1SCIE)) +
  geom_point()
  
```

That graph is quite dense, so we can use the `alpha` function to make the points slightly transparent, `size` to make them smaller, and set their colour. I will also tidy up the axis names and add a line.

```{r}
#| echo: true
Chinaplot <- PISA_2018 %>%
  select(CNT, PV1READ, PV1SCIE) %>%
  filter(CNT == "B-S-J-Z (China)")

ggplot(Chinaplot, aes(x = PV1READ, y = PV1SCIE)) +
  geom_point(alpha = 0.6, size=0.1, colour = "red") +
  xlab("Reading score") +
  ylab("Science score") +
  geom_smooth(method = "lm", colour = "black")
  
```

### Density plots

An alternative type of plot, you have seen before, is the density plot, which is a kind of continuous histogram. The density plot can be useful for visualising the achievement scores of students. For example, the mathematics scores of girls and boys (recall the gender variable is `ST004D01T`) in the US. Notice, for the plot, I use `aes` to set my x variable, and then specify that the plot should fill by gender (`fill=ST004D01T`). Finally, in `geom_density(alpha=0.6)` I set the alpha to 0.6 to make the fill areas partially transparent.

```{r}
#| echo: true
USMathplot <- PISA_2018 %>%
  select(CNT, PV1MATH, ST004D01T) %>%
  filter(CNT == "United States")

ggplot(USMathplot, aes(x = PV1MATH, fill = ST004D01T)) +
  geom_density(alpha = 0.6)
```

### Facet wrapping - producing the same graph for multiple countries.

A powerful feature of ggplot is being able to produce the same graph for multiple values of a variable, for example, for mutliple countries. For example, we may want to produce the density graph of `PV1MATH` score by gender, for several countries in the data set. To do that, we produce a data set of `PV1MATH` scores, and gender (`ST004D01T`) and filter for four countries (Philippines, UK, Bulgaria and Germany). We use the same code as above to plot the graphs but add `+facet_wrap(.~CNT)` - `facet_wrap` tells ggplot to produce a multipanel plot and `.~CNT` means do the same as above, but vary across countries.

```{r}
#| echo: true
Mathplot <- PISA_2018 %>%
  select(CNT, PV1MATH, ST004D01T) %>%
  filter(CNT == "Philippines"|CNT == "United Kingdom"|CNT == "Bulgaria" |
           CNT == "Germany")

ggplot(Mathplot, aes(x = PV1MATH, fill = ST004D01T)) +
  geom_density(alpha = 0.6) +
  facet_wrap(. ~ CNT)
```

### Categorising responses

A useful analytical choice is to categorise some a numerical variable into ordinal classes. For example, rather than treating `WEALTH` as a continuous scale, you might want to split into high and low wealth groups (for example, those above and below the mean value). To do this, first calculate the mean `mean(WEALTH)`. Then we add a new vector, which we will call `wealthclass` using the `mutate` function. We set the value of this `wealthclass` using `ifelse`. If `WEALTH` is more than the mean score, we set `wealthclass` to `HIGH`, and if it is anything else, we set it to `LOW`. We do that using `wealthclass = ifelse(WEALTH > MeanUKwealth, "High", "Low")`. Note in `ifelse` the first value is returned if the identity is true (i.e. if `WEALTH > MeanUKwealth` `wealthclass` is set to `High`). If the value if not true, the second value is set (e.g. if `WEALTH` is not `> MeanUKwealth` then `wealthclass` is set to `LOW`)

For example, create a dataset of UK participants `WEALTH` sorted into `HIGH` and `LOW` categories.

TODO: @Richard, see the updated code below

```{r}
#| echo: true
UKPISA2018 <- PISA_2018 %>%
  select(CNT, WEALTH) %>%
  filter(CNT == "United Kingdom") %>%
  mutate(wealthclass =  ifelse(WEALTH > mean(WEALTH, na.rm=TRUE), 
                               "High", 
                               "Low"))
```

## Seminar activities

### Task 1 Discussion activity

### Task 2 Create a ranked list

Create a ranked list of countries by their mean science scores (`PV1SCIE`). What are the top five countries for science? Do the same for reading (`PV1READ`). What patterns do you notice? Why might a researcher be critical of such rankings \[Extension: Include the standard deviation of each country (hint: use the `sd` function) - can you detect any patterns?\]

```{r}
#| echo: true
#| eval: true
#| code-fold: true
#| code-summary: "Show the answer"

PISA2018SciRank<-PISA_2018%>%
  select(CNT, PV1SCIE)%>% # Select variables of interest
  group_by(CNT)%>% # group by country
  summarise(meansci = mean(PV1SCIE))%>% 
     # summarise  country data to find the mean Sci score
  arrange(desc(meansci)) # arrange in descending order based on the meansci score
print(PISA2018SciRank)


PISA2018WealthRank<-PISA_2018%>%
  select(CNT, WEALTH)%>% # Select variables of interest
  group_by(CNT)%>% # group by country
  summarise(meanwel = mean(WEALTH, na.rm=TRUE))%>% 
     # summarise  country data to find the mean Sci score
  arrange(desc(meanwel)) # arrange in descending order based on the meansci score
print(PISA2018WealthRank)

PISA201ReadRank<-PISA_2018%>%
  select(CNT, PV1READ)%>% # Select variables of interest
  group_by(CNT)%>% # group by country
  summarise(meanread = mean(PV1READ))%>% 
     # summarise  country data to find  mean Read score
  arrange(desc(meanread)) # arrange in descending order based on the meansci score

# With standard deviations

PISA2018SciRank<-PISA_2018%>%
  select(CNT, PV1SCIE)%>% # Select variables of interest
  group_by(CNT)%>% # group by country
  summarise(meansci = mean(PV1SCIE), 
            sdsci = sd(PV1SCIE))%>% 
  # summarise  country data to find the mean Sci score
  arrange(desc(meansci)) # arrange in descending order based on the meansci score

PISA201ReadRank<-PISA_2018%>%
  select(CNT, PV1READ)%>% # Select variables of interest
  group_by(CNT)%>% # group by country
  summarise(meanread = mean(PV1READ),
            sdread = sd(PV1READ))%>% 
  # summarise  country data to find  mean Read score
  arrange(desc(meanread)) 
  # arrange in descending order based on the meansci score

```

### Task 2.5 Plot distributions of scores

```{r}
WealthcompPISA<-PISA_2018 %>%
  select(CNT, ESCS, WEALTH, ST004D01T)%>%
  filter(CNT == "Japan" | CNT == "United Kingdom" | CNT == "Colombia" | CNT == "Sweden")

ggplot(WealthcompPISA, aes(x = ESCS, y = WEALTH, colour=ST004D01T))+
  geom_point(size=0.1, alpha=0.5)+
  facet_wrap(.~CNT)


```

### Task 3 Plot distributions of scores

::: question
-   Use geom_density to plot distributions Plot the distribution of Japanese and UK mathematics scores - what patterns do you notice?
:::

```{r}
#| echo: true
#| eval: true
#| code-fold: true
#| code-summary: "Show the answer"

JPUKMath<-PISA_2018 %>%
  select(CNT, PV1MATH) %>%
  filter(CNT == "United Kingdom"|CNT == "Japan")

ggplot(data = JPUKMath,
       aes(x = PV1MATH, fill = CNT)) +
  geom_density(alpha = 0.5)

```

### Task 4 Plot distributions of scores by gender

::: question
-   Examine gender differences: Plot the distributions of mathematics achievement in the UK by gender.
:::

```{r}
#| echo: true
#| eval: true
#| code-fold: true
#| code-summary: "Show the answer"

UKMathGender <- PISA_2018 %>%
  select(CNT, PV1MATH, ST004D01T) %>%
  filter(CNT == "United Kingdom")

ggplot(data = UKMathGender,
       aes(x = PV1MATH, fill = ST004D01T)) +
  geom_density(alpha = 0.5)

```

### Task 5 Facet wrap by country

::: question
Plot density graphs of gender differences in mathematics scores in the UK, US, Japan, Korea and Finland. Hint use `facet_wrap(.~CNT)`
:::

```{r}
#| echo: true
#| eval: true
#| code-fold: true
#| code-summary: "Show the answer"

MathGender <- PISA_2018 %>%
  select(CNT, PV1MATH, ST004D01T) %>%
  filter(CNT == "United Kingdom"|CNT == "United States"|CNT == "Japan"
         | CNT=="Korea"|CNT == "Finland")

ggplot(data = MathGender,
       aes(x = PV1MATH, fill = ST004D01T)) +
  geom_density(alpha = 0.5) +
  facet_wrap(.~CNT)

```

### Task 6 Plot a scatter graph

::: question
Plot a scatter graph of mean mathematics achievement (y-axis) by mean wealth (x-axis) with each country as a single point.
:::

```{r}
#| echo: true
#| eval: true
#| code-fold: true
#| code-summary: "Show the answer"

Wealthdata <- PISA_2018 %>%
  select(CNT, WEALTH, PV1MATH) %>%
  group_by(CNT) %>%
  summarise(MeanWealth=mean(WEALTH, na.rm = TRUE),
            MeanMath=mean(PV1MATH, na.rm = TRUE))

ggplot(data = Wealthdata,
       aes(x = MeanWealth, y = MeanMath)) +
  geom_point(alpha = 0.5) +
  xlab("Wealth") +
  ylab("Mathematics score")

```

Highlight outliers (any score of over 550) in a different colour

```{r}
#| echo: true
#| eval: true
#| code-fold: true
#| code-summary: "Show the answer"

Wealthdata <- PISA_2018 %>%
  select(CNT, WEALTH, PV1MATH) %>%
  group_by(CNT) %>%
  summarise(MeanWealth = mean(WEALTH, na.rm = TRUE),
            MeanMath = mean(PV1MATH, na.rm = TRUE)) %>%
  mutate(label=ifelse(MeanMath > 550, "Red", "Blue"))

ggplot(data = Wealthdata,
       aes(x = MeanWealth, y = MeanMath, colour = label)) +
  geom_point() +
  xlab("Wealth") +
  ylab("Mathematics score")

```

Add the country names as a label to the outliers

```{r}
#| echo: true
#| eval: true
#| code-fold: true
#| code-summary: "Show the answer"

Wealthdata <- PISA_2018 %>%
  select(CNT, WEALTH, PV1MATH) %>%
  group_by(CNT) %>%
  summarise(MeanWealth = mean(WEALTH, na.rm = TRUE),
            MeanMath = mean(PV1MATH, na.rm = TRUE)) %>%
  mutate(label = ifelse(MeanMath>550, "Red", "Blue")) %>%
  mutate(labelname = ifelse(MeanMath>550, as.character(CNT), NA))
  

ggplot(data = Wealthdata,
       aes(x = MeanWealth, y = MeanMath, colour = label)) +
  geom_point() +
  geom_text(aes(label = labelname), 
            colour = "black", 
            check_overlap = TRUE) +
  xlab("Wealth") +
  ylab("Mathematics score")

```

### Task 7 Plot Likert responses using facet wrapping

::: question
Examine Likert responses by country using facet plot.

For `ST125Q01NA` - How old were you when you started early childhood education? Plot responses, first, for the whole data set, then facet plot for the UK, Germany, Belgium, Austria, France, Poland, Estonia, Finland and Italy.
:::

```{r}
#| echo: true
#| eval: true
#| code-fold: true
#| code-summary: "Show the answer"

Culturedata<-PISA_2018 %>%
  select(CNT, ST125Q01NA) %>%
  group_by(CNT)
  
ggplot(data = Culturedata,
       aes(x = ST125Q01NA)) +
  geom_bar() +
  xlab("How old were you when you started early childhood education?") +
  theme(axis.text.x = element_text(angle = 90, vjust = 0.5, hjust = 1))

```

Then use faceting to split the plots by country

```{r}
#| echo: true
#| eval: true
#| code-fold: true
#| code-summary: "Show the answer"

Culturedata <- PISA_2018 %>%
  select(CNT, ST125Q01NA) %>%
  filter(CNT == "United Kingdom"|CNT == "Germany" | CNT == "Belgium"
         | CNT == "Austria"| CNT == "France" | CNT == "Poland"
         | CNT == "Estonia" | CNT=="Finland"| CNT=="Italy")
  
ggplot(data = Culturedata,
       aes(x = ST125Q01NA, fill = CNT))+
  geom_bar()+
  xlab("How old were you when you started early childhood education?") +
  theme(axis.text.x = element_text(angle = 90, vjust = 0.5, hjust = 1)) +
  facet_wrap(. ~ CNT)

```

### Task 8 Categorise `WEALTH` scores

::: question
Categorising Variables

Split the `WEALTH` variable for the UK and Germany into the following groups:

| WEALTH       | Name of category |
|--------------|------------------|
| \>1          | Very High        |
| 0\>WEALTH\<1 | High             |
| 0\<          | Low              |

Plot bar graphs of participants in these categories for both countries
:::

```{r loading_data_wealth}
#| echo: true
#| eval: true
#| code-fold: true
#| code-summary: "Show the answer"

Wealth <- PISA_2018 %>%
  select(CNT, WEALTH) %>%
  filter(CNT == "United Kingdom" | CNT == "Germany") %>%
  mutate(WTHCTG = ifelse(WEALTH > 0, 
                       ifelse(WEALTH > 1,
                             "Very high",
                             "High"), 
                       "Low"))%>%
  group_by(CNT) %>%
  droplevels()

ggplot(data = Wealth, 
       aes(x = WTHCTG, fill = WTHCTG))+
  geom_bar()+
  facet_wrap(.~CNT)+
  xlab("Wealth grouping")

```

:::

### Task 9 Compare the association between mathematics and science PV values across three diverse countries

::: question
Plot scatter plots of science versus mathematics achievement in United Kingdom, Qatar and Brazil
:::

```{r loading_data_sciMaths}
#| echo: true
#| eval: true
#| code-fold: true
#| code-summary: "Show the answer"
# All students
SciMaths <- PISA_2018 %>%
  select(CNT, PV1MATH, PV1SCIE, ST004D01T) %>%
  filter(CNT == "Colombia" | CNT == "New Zealand" | CNT == "Qatar"| CNT=="Israel") %>%
  droplevels()

ggplot(data = SciMaths, 
       aes(x = PV1MATH, y = PV1SCIE, colour=ST004D01T))+
  geom_point(size=0.1, alpha=0.5)+
  facet_wrap(.~CNT)


# Low achieving 
SciMaths <- PISA_2018 %>%
  select(CNT, PV1MATH, PV1SCIE, ST004D01T) %>%
  filter(CNT == "Colombia" | CNT == "New Zealand" | CNT == "Qatar"| CNT=="Israel") %>%
  filter(PV1MATH<400)%>%
  filter(PV1SCIE<400)%>%
  droplevels()

ggplot(data = SciMaths, 
       aes(x = PV1MATH, y = PV1SCIE, colour=ST004D01T))+
  geom_point(size=0.1, alpha=0.5)+
  facet_wrap(.~CNT)

```

- What differences can you see between the countries?

:::

## Proposed new activities

### Task 1 Discussion activity

\[Add ideas for discussion tasks\]

### Task 2 Create a ranked list

::: question
Create a ranked list of countries by their mean wealth (`WEALTH`). What are the top five countries by WEALTH? What do you notice about the patterns of `WEALTH`? Is there anything unusual? Table 16.12 in this [document](https://www.google.com/url?sa=t&rct=j&q=&esrc=s&source=web&cd=&ved=2ahUKEwjDncfcnLeCAxWyTkEAHdj3CkoQFnoECBEQAQ&url=https%3A%2F%2Fwww.oecd.org%2Fpisa%2Fdata%2Fpisa2018technicalreport%2FPISA2018_Technical_Report_chapter-16_Background_Questionnaires.xlsx&usg=AOvVaw3jKeWRTlYCvLt1TCHVeiKa&opi=89978449) shows the items that are used to develop the `WEALTH` variable - can you use them to explain the ranking? \[Notice the difference weightings of items by countries\]

Do the same for reading (`ESCS`). What patterns do you notice? Why might a researcher be critical of such rankings. You can see how the `ESCS` variable is generated on this [page](https://www.oecd-ilibrary.org//sites/0a428b07-en/index.html?itemId=/content/component/0a428b07-en#s131).

Extension - add the standard deviations of the `WEALTH` and `ESCS` variables to the tables.
:::

::: callout-note
Don't forget to add `na.rm=TRUE` to the mean function to ignore `NA`s.
:::

```{r}
#| echo: true
#| eval: true
#| code-fold: true
#| code-summary: "Show the answer"
# For wealth
PISA2018WelRank<-PISA_2018%>%
  select(CNT, WEALTH)%>% # Select variables of interest
  group_by(CNT)%>% # group by country
  summarise(meanwel = mean(WEALTH, na.rm=TRUE))%>% 
     # summarise  country data to find the mean wealth score
  arrange(desc(meanwel)) # arrange in descending order based on the meanwel score
print(PISA2018WelRank)

# For ESCS
PISA2018ESCSRank<-PISA_2018%>%
  select(CNT, ESCS)%>% # Select variables of interest
  group_by(CNT)%>% # group by country
  summarise(meanescs = mean(ESCS, na.rm=TRUE))%>% 
     # summarise  country data to find  mean ESCS score
  arrange(desc(meanescs)) # arrange in descending order based on the meanescs score
print(PISA2018ESCSRank)


# With standard deviations
PISA2018WelRank<-PISA_2018%>%
  select(CNT, WEALTH)%>% # Select variables of interest
  group_by(CNT)%>% # group by country
  summarise(meanwel = mean(WEALTH, na.rm=TRUE),
            sdwel=sd(WEALTH, na.rm=TRUE))%>% 
     # summarise  country data to find the mean wealth score
  arrange(desc(meanwel)) # arrange in descending order based on the meanwel score
print(PISA2018WelRank)

# For ESCS
PISA2018ESCSRank<-PISA_2018%>%
  select(CNT, ESCS)%>% # Select variables of interest
  group_by(CNT)%>% # group by country
  summarise(meanescs = mean(ESCS, na.rm=TRUE),
            sdescs=sd(ESCS, na.rm=TRUE))%>% 
     # summarise  country data to find  mean ESCS score
  arrange(desc(meanescs)) # arrange in descending order based on the meanescs score
print(PISA2018ESCSRank)



```
### Task 3 Comparing measures of wealth
::: question
To compare the correlation of different measures of wealth, produce graphs of the number of cars (`ST012Q02TA`), number of computers (`ST012Q06NA`) and `HOMEPOS` (home possessions) variable for the UK, Qatar and Japan. What patterns do you notice about the different measures? How can affluence best be reported?
:::

::: callout-note
To show multiple graphs in the same image, you can install the package `gridextra`:

```{r}
#| echo: true
#| eval: true

## install gridExtra:
# install.packages("gridExtra")
library(gridExtra)

# Store your plots as variables
p1 <- ggplot(PISA2018ESCSRank, aes(x=CNT, y=meanescs))+
  geom_bar(stat='identity') +
  theme(axis.text.x = element_text(angle = 90, vjust = 0.5, hjust=1))

p2 <- ggplot(PISA2018WelRank, aes(x=CNT, y=meanwel)) +
  geom_bar(stat='identity')+
  theme(axis.text.x = element_text(angle = 90, vjust = 0.5, hjust=1))

# You can then use the grid arrange function to plot both graphs, giving the number of graphs per row

grid.arrange(p1, p2, nrow = 2)
```

:::


```{r}

#| echo: true
#| eval: true
#| code-fold: true
#| code-summary: "Show the answer"

# Selecting, HOMEPOS, Item ST012Q02TA How many in your home: Cars, and Item ST012Q06NA How many in your home: Computers (desktop computer, portable laptop, or notebook)
library(gridExtra)
 
Pos <- PISA_2018 %>%
  select(CNT, HOMEPOS, ST012Q02TA, ST012Q06NA) %>%
  rename("Cars"="ST012Q02TA") %>%
  rename("Computers"="ST012Q06NA") %>%
  filter(CNT=="United Kingdom"|CNT=="Qatar"|CNT=="Japan")
 
p1 <- ggplot(Pos, aes(x=HOMEPOS)) +
  geom_histogram(fill="skyblue") +
  facet_grid(.~CNT)
 
p2 <- ggplot(Pos, aes(x=Cars)) +
  geom_bar(fill="green") +
  facet_grid(.~CNT) +
  theme(axis.text.x = element_text(angle = 90, vjust = 0.5, hjust=1))
 
p3 <- ggplot(Pos, aes(x=Computers)) +
  geom_bar(fill="red") +
  facet_grid(.~CNT) +
  theme(axis.text.x = element_text(angle = 90, vjust = 0.5, hjust=1))
 
grid.arrange(p1, p2, p3, nrow = 3)
```


### Task 4 Comparing distributions of wealth by country and gender
::: question
Use `geom_density` to produce density plots show the wealth of girls and boys in the sample for the UK, Japan and Qatar. 
:::

::: callout-note
To produce the same graph for multiple countries, remember you can use the `facet_grid` function.

```{r}
#| echo: true
#| eval: true
# Create a subset of ESCS scores by country for France, Germany and the US, including gender
PISAESCS <- PISA_2018 %>%
  select(CNT, ESCS, ST004D01T) %>%
  filter(CNT == "United States"|CNT == "France"| CNT == "Germany")

# Create a density plot, set to fill by gender
ggplot(PISAESCS, aes(x=ESCS, fill=ST004D01T))+
  geom_density(alpha=0.5)+ # set alpha (opacity) to 0.5 to show overlap
# to produce separate graphs for each country
  facet_grid(.~CNT) # i.e. facet by country





```

:::


```{r}

#| echo: true
#| eval: true
#| code-fold: true
#| code-summary: "Show the answer"
# Create a data frame for the UK Japan and Qatar including WEALTH and gender
PISAWel <- PISA_2018 %>%
  select(CNT, WEALTH, ST004D01T) %>%
  filter(CNT=="United Kingdom"|CNT=="Japan"|CNT=="Qatar")


# Create the density plot and then facet wrap by country, set fill by gender

ggplot(PISAWel, aes(x=WEALTH, fill=ST004D01T)) +
  geom_density(alpha=0.5) +
  facet_grid(.~CNT)
```

### Task 5 Plot scatter plots of Wealth against ESCS for the UK (and add colour by gender)
::: question
Use `geom_point` to produce a scatter plot of `WEALTH` scores against `ESCS` for students in the UK. Extension: change the colour of points by gender. What does the pattern show?
:::


```{r}

#| echo: true
#| eval: true
#| code-fold: true
#| code-summary: "Show the answer"
# Create a data frame for the UK including WEALTH and gender
PISAUKWel <- PISA_2018 %>%
  select(CNT, WEALTH, ESCS, ST004D01T) %>%
  filter(CNT == "United Kingdom")


# Create the scatter plot and set colour by gender

ggplot(PISAUKWel, aes(x = WEALTH, y = ESCS, colour = ST004D01T)) +
  geom_point(size = 0.1, alpha = 0.5) +
  xlab("Wealth") +
  ylab("ESCS")

```

### Task 6 Plot scatter plots of Wealth against ESCS for the UK, Japan, and Qatar (and add colour by gender)
::: question
Use `geom_point`, and `facet_grid(.~CNT)` to produce a scatter plot of `WEALTH` scores against `ESCS` for students in the UK, Japan and Qatar. Extension: change the colour of points by gender. What does the pattern show?
:::


```{r}

#| echo: true
#| eval: true
#| code-fold: true
#| code-summary: "Show the answer"
# Create a data frame for the UK including WEALTH and gender
PISAWel <- PISA_2018 %>%
  select(CNT, WEALTH, ESCS, ST004D01T) %>%
  filter(CNT=="United Kingdom"|CNT=="Japan"|CNT=="Qatar")


# Create the scatter plot and set colour by gender

ggplot(PISAWel, aes(x = WEALTH, y = ESCS, colour = ST004D01T)) +
  geom_point(size = 0.1, alpha = 0.5)+
  xlab("Wealth") +
  ylab("ESCS") +
  facet_grid(.~CNT)

```
### Task 8 Plot scatter plots of mean Wealth against mean ESCS for the whole data set
::: question
Create a summary data frame of mean `WEALTH` and mean `ESCS` scores for all the countries in the data. What patterns do you see? Extension: Use `geom_label_repel` to label countries with low scores.
:::

::: callout-note

Don't forget to use `group_by` and `summarise` to create a data frame to plot. Don't forget to add `na.rm=TRUE` to `mean` functions.


```{r}
#| echo: true
#| eval: true
# Create a summary of science and mathematics scores
PISASciMath <- PISA_2018 %>%
  select(CNT, PV1MATH, PV1SCIE) %>%
  group_by(CNT) %>%
  summarise(meansci=mean(PV1SCIE, na.rm=TRUE),
            meanmath=mean(PV1MATH, na.rm=TRUE))

# Create a scatter plot
ggplot(PISASciMath, aes(x=meansci, y=meanmath, colour="red")) +
  geom_point(alpha=0.5, size=0.5) 


```
For the extension: to label countries, first use `mutate` to create a new column. Then, use `if_else` to add either the country name (`CNT`) or nothing (`""`) to columns that meet some criteria (e.g. above or below some score). Notice that in the `if_else` function the syntax is `if_else(condition, <what to do if true>,<what to do in other cases>)`. 

For example, to change the label column to the country name if the mean science and mathematics score are over 520, and blank in other cases, we use :`mutate(label=if_else(meansci> 520 & meanmath > 520,CNT, ""))`

We can then use `geom_text_repel` which adds labels, but 'repels' them, making sure the text doesn't overlap.

```{r}
#| echo: true
#| eval: true
# Create a summary of science and mathematics scores
PISASciMath <- PISA_2018 %>%
  select(CNT, PV1MATH, PV1SCIE) %>%
  group_by(CNT) %>%
  summarise(meansci=mean(PV1SCIE, na.rm=TRUE),
            meanmath=mean(PV1MATH, na.rm=TRUE))

# mutate to add a column of labels for high scoring countries
PISASciMath <- PISASciMath %>%
  mutate(label=if_else(meansci> 520 & meanmath > 520,
                       CNT, ""))

# Create a scatter plot, set to fill by gender, and add geom_text_repel to add the labels
ggplot(PISASciMath, aes(x=meansci, y=meanmath, colour="red"))+
  geom_point(alpha=0.5, size=0.5)+
  geom_text_repel(label=PISASciMath$label)

  


```

:::


```{r}

#| echo: true
#| eval: true
#| code-fold: true
#| code-summary: "Show the answer"
# Create a summary of ESCS and wealth scores
PISAWelESCS <- PISA_2018 %>%
  select(CNT, ESCS, WEALTH) %>%
  group_by(CNT)%>%
  summarise(meanESCS=mean(ESCS, na.rm=TRUE),
            meanwel=mean(WEALTH, na.rm=TRUE))

# mutate to add a column of labels for high scoring countries
PISAWelESCS <- PISAWelESCS %>%
  mutate(label=if_else(meanESCS< (-1.0) & meanwel < (-1.5),
                       CNT, ""))

# Create a scatter plot, set to fill by gender
ggplot(PISAWelESCS, aes(x=meanwel, y=meanESCS, colour="red"))+
  geom_point(alpha=0.5, size=0.5) +
  geom_text_repel(label=PISAWelESCS$label)+
  xlab("Mean Wealth")+
  ylab("Mean ESCS")

```